# Capstone 2: Smoking and Cardiovascular Risk {#sec-capstone-smoking-cardiovascular-risk}

::: {.pbl-box .pbl-problem}
**The Problem**

Capstone 1's model found smoking's odds ratio for hypertension (2.25) wasn't statistically significant. Your professor now asks for a broader, standalone investigation: across *every* cardiovascular-relevant measurement in this dataset (systolic BP, diastolic BP, BMI, and hypertension status itself), does smoking status show a consistent pattern of association, or does it look different depending on which outcome you check?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Compare smokers and non-smokers across multiple cardiovascular-relevant measurements, choosing the correct test for each variable type, and synthesize the results into one coherent verdict.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 11 taught different tests for different variable-type combinations: two-sample t-tests for numeric outcomes, chi-square for two categorical variables. Which test fits "smoking status vs. systolic blood pressure" (numeric)? Which fits "smoking status vs. hypertension status" (both categorical)?
- If four separate comparisons are run (systolic BP, diastolic BP, BMI, hypertension), and only one shows significance, would that automatically mean smoking is unrelated to cardiovascular risk in general?
:::

## The R Concept

This capstone runs the *same* two-sample comparison logic (Level 11, Pages 5 and 7) across four different outcomes, methodically:

```{r}
#| eval: false
t.test(systolic_bp ~ smoker, data = participants)     # numeric vs. categorical
t.test(diastolic_bp ~ smoker, data = participants)    # numeric vs. categorical
t.test(bmi ~ smoker, data = participants)             # numeric vs. categorical
chisq.test(table(participants$smoker, participants$hypertension))  # categorical vs. categorical
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

t.test(systolic_bp ~ smoker, data = participants)
```

```

	Welch Two Sample t-test

data:  systolic_bp by smoker
t = -0.21305, df = 41.106, p-value = 0.8323
95 percent confidence interval:
 -9.750749  7.889630
sample estimates:
 mean in group No mean in group Yes
         129.5000          130.4306
```

## Explore

Try running the same `t.test()` pattern for `bmi` and `diastolic_bp` before looking at the Solve section below — form your own expectation first (Level 11's "Think Before Coding" habit) about whether either will show a significant difference, given that systolic blood pressure alone did not.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

participants$hypertension <- as.numeric(participants$systolic_bp >= 140)

bmi_test  <- t.test(bmi ~ smoker, data = participants)
dbp_test  <- t.test(diastolic_bp ~ smoker, data = participants)
hyper_tab <- table(participants$smoker, participants$hypertension)
hyper_test <- chisq.test(hyper_tab)

bmi_test$p.value
dbp_test$p.value
hyper_test$p.value
```

```
[1] 0.5458577
```

```
[1] 0.5266651
```

```
[1] 0.2480914
```

```{r}
#| eval: false
hyper_tab
prop.table(hyper_tab, margin = 1)
```

```

       0  1
  No  53 19
  Yes 14 10
```

```
           0         1
  No  0.7361111 0.2638889
  Yes 0.5833333 0.4166667
```

None of the four comparisons reaches statistical significance (all p-values well above 0.05): BMI (p = 0.546), diastolic BP (p = 0.527), systolic BP (p = 0.832, from the Demonstration), and hypertension status (p = 0.248). Yet the *directions* are consistent with clinical expectation throughout: smokers show a numerically higher hypertension rate (41.7% vs. 26.4%), higher mean diastolic BP (81.96 vs. 79.93), and higher mean systolic BP — none reaching significance, but none pointing the "wrong" way either (only BMI trends slightly lower among smokers, 26.6 vs. 27.3, a small and non-significant difference).

## Interpret the Result

This capstone's honest verdict is genuinely more nuanced than "smoking is/isn't related to cardiovascular risk": across four separate comparisons, **not one** reaches statistical significance in this 100-participant sample, but **three of four** point in the clinically expected direction (higher hypertension rate, higher diastolic and systolic BP among smokers). This is exactly the pattern expected from a real but modest effect studied in too small a sample — Levels 7, 11, and 12 all made the same point repeatedly: "not significant" describes *this sample's evidence*, not a settled claim that no true relationship exists. A well-designed follow-up study with more smokers (only 25 in this entire dataset) would be a reasonable next recommendation, rather than concluding smoking has no cardiovascular relevance here.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Concluding "smoking has no effect on cardiovascular risk" from four non-significant tests.** A consistent directional pattern across multiple related outcomes, even without any single test reaching significance, is meaningfully different evidence than results that scatter randomly in both directions — the first is suggestive; pure noise typically wouldn't line up this consistently.
- **Running four separate tests without acknowledging the multiple-comparisons issue.** Level 12 Page 5 explained why running many separate tests inflates the chance of a false positive — the same caution applies here in reverse: running four tests and finding *none* significant is a stronger form of evidence for "no detected effect" than if only one or two had been run.
- **Cherry-picking the smallest p-value (hypertension, p = 0.248) as "the closest to significant" and overselling it.** All four p-values are well above 0.05; none should be described as "trending toward significance" in a way that overstates the actual evidence.
:::

## Practice Problem

Run `t.test(age ~ smoker, data = participants)`. Does age differ significantly between smokers and non-smokers in this sample? How might an age difference between the two groups (if one exists) complicate interpreting the blood-pressure comparisons above?

## Challenge Problem

Fit `glm(hypertension ~ smoker, data = participants, family = binomial)` (a simple, one-predictor version of Capstone 1's larger model) and compare its `smoker` coefficient and p-value to the three-predictor model from Capstone 1. Does adjusting for age and BMI change smoking's estimated association with hypertension much?

## Key Takeaways

- Comparing groups across multiple related outcomes requires choosing the correct test for each variable-type pairing (t-tests for numeric outcomes, chi-square for two categorical variables) — the same decision framework from Level 11, applied systematically rather than to a single outcome at a time.
- A consistent directional pattern across several non-significant comparisons is more informative than a single isolated non-significant result — it suggests a real, if modest and statistically unconfirmed, underlying relationship rather than pure noise.
- Reporting "not significant" honestly (rather than as "no effect" or as "nearly significant") is a recurring discipline throughout this entire course, from Level 7's descriptive claims through this final capstone.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Across the four comparisons in this capstone, how many reached statistical significance at p < 0.05?
   a) All four
   b) Three of four
   c) One of four
   d) None of the four

2. **True or false.** A consistent directional pattern across several non-significant tests provides no more information than results that point in random, inconsistent directions.

3. **Predict the output.** Given a hypertension rate of 41.7% among smokers and 26.4% among non-smokers, but a chi-square p-value of 0.248, is this difference statistically significant at the conventional threshold?

4. **Short answer.** Why might a well-designed follow-up study with a larger number of smokers be a more appropriate conclusion than declaring smoking unrelated to cardiovascular risk in this dataset?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **d)** None of the four comparisons reached statistical significance in this sample.
2. **False.** A consistent direction across multiple related, non-significant comparisons is meaningfully more suggestive of a real underlying effect than results scattering randomly in both directions, even though neither pattern alone confirms significance.
3. No — a p-value of 0.248 is well above 0.05, so this difference is not statistically significant despite the descriptive gap.
4. This dataset contains only 25 smokers total, which limits statistical power to detect even a real, modest effect — recommending a larger sample is a scientifically appropriate response to "not enough evidence yet," rather than incorrectly treating a small sample's non-significant result as proof of no true relationship.

</details>
:::
