# Capstone 3: Geographic Patterns Across Counties {#sec-capstone-geographic-patterns}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 11's ANOVA found no significant county-level difference in blood pressure (F = 0.539, p = 0.708). But your professor points out that's only *one* health measurement — she wants a complete geographic analysis: check every relevant numeric measurement across the five counties, visualize the comparison properly, and cross-validate the result using the county-split clinic files from Level 10's automation, rather than trusting a single pre-combined CSV alone.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Test multiple numeric measurements for county-level differences, visualize the comparison, and independently verify the underlying data using Level 10's file-automation tools.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 11 tested blood pressure across 5 counties with `aov()`. Could the exact same approach apply to BMI, without writing a new kind of test?
- Level 10's automation capstone combined `data/clinics/*.csv` back into one dataset. If that automated reconstruction doesn't match `participants.csv` exactly, would you trust the geographic analysis at all?
:::

## The R Concept

This capstone chains three separate tools together: automated file reading (Level 10), a statistical test (Level 12), and a visualization (Level 8):

```{r}
#| eval: false
clinic_files <- list.files("data/clinics", pattern = "\\.csv$", full.names = TRUE)
clinic_data <- do.call(rbind, lapply(clinic_files, read.csv))

bmi_county_model <- aov(bmi ~ county, data = clinic_data)
summary(bmi_county_model)

ggplot(clinic_data, aes(x = county, y = bmi)) + geom_boxplot()
```

## Demonstration

```{r}
#| eval: false
clinic_files <- list.files("data/clinics", pattern = "\\.csv$", full.names = TRUE)
clinic_data <- do.call(rbind, lapply(clinic_files, read.csv))

nrow(clinic_data)
identical(sort(clinic_data$id), sort(read.csv("data/participants.csv")$id))
```

```
[1] 100
```

```
[1] TRUE
```

## Explore

Before trusting any geographic comparison, always confirm the automated reconstruction matches the original data exactly — as just demonstrated. Try `table(clinic_data$county)` to reproduce Level 10's known county counts (20/21/17/16/26) directly from the automated file-reading pipeline, independent of `participants.csv`.

## Solve the Problem

```{r}
#| eval: false
clinic_files <- list.files("data/clinics", pattern = "\\.csv$", full.names = TRUE)
clinic_data <- do.call(rbind, lapply(clinic_files, read.csv))

bmi_county_model <- aov(bmi ~ county, data = clinic_data)
summary(bmi_county_model)
```

```
            Df Sum Sq Mean Sq F value Pr(>F)
county       4    135    33.8   1.614  0.178
Residuals   90   1888    21.0
```

```{r}
#| eval: false
summary(bmi_county_model)[[1]][["Pr(>F)"]][1]
```

```
[1] 0.1776636
```

```{r}
#| eval: false
library(ggplot2)
ggplot(clinic_data, aes(x = county, y = bmi)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(x = "County", y = "BMI", title = "BMI Distribution by County")
```

```
[Renders a 5-box boxplot, one per county, with Springvale's box
 visibly positioned lowest (mean 25.24) and Cedar Falls' highest
 (mean 28.42), though the ANOVA shows this spread isn't statistically
 significant.]
```

Like blood pressure (Level 11, F = 0.539, p = 0.708), BMI shows no statistically significant county-level difference either (F = 1.614, p = 0.178) — despite a visually noticeable spread in the boxplot (Springvale's mean of 25.24 versus Cedar Falls' 28.42, roughly a 3-point difference). The automated reconstruction from `data/clinics/` exactly matches `participants.csv` (confirmed via `identical()` above), so this conclusion doesn't depend on trusting a single pre-combined file.

## Interpret the Result

Two separate health measurements (blood pressure and BMI), tested the same way, reach the same conclusion: no statistically significant geographic pattern across these five counties in this sample. This is a meaningfully different, more complete answer than checking blood pressure alone — a geographic health disparity claim resting on just one measurement would be considerably weaker evidence than one checked (and not found) across multiple relevant outcomes. The independent cross-validation against Level 10's automated file-reading pipeline also confirms this conclusion isn't an artifact of a single, possibly-corrupted source file.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Trusting a visually striking boxplot spread without checking the ANOVA's p-value.** Springvale's visibly lower BMI box might look like a real geographic pattern, but the F-test (p = 0.178) shows this spread is consistent with random sampling variation across 5 groups of modest size (14-26 participants each).
- **Skipping the cross-validation step against Level 10's clinic files.** Recomputing a result from an independently-reconstructed dataset (rather than only ever reading `participants.csv`) is exactly the kind of due diligence that catches a corrupted or outdated file before it silently produces a wrong conclusion.
- **Testing only one measurement and generalizing to "geography doesn't matter for health outcomes here."** This capstone's stronger conclusion — that *two* separate measurements both show no significant county effect — required deliberately checking more than one outcome, not stopping at the first test.
:::

## Practice Problem

Run `aov(age ~ county, data = clinic_data)` and `aov(diastolic_bp ~ county, data = clinic_data)`. Do either of these show a significant county-level difference, completing a four-measurement geographic comparison alongside systolic BP and BMI?

## Challenge Problem

Level 10's automation capstone reconstructed `participants.csv` from 5 separate clinic files. Modify the file-reading loop in this page's Demonstration to also record, for each file, its county name and row count into a small summary data frame — reproducing Level 10's known county counts (20, 21, 17, 16, 26) without ever reading `participants.csv` directly, only the 5 clinic files.

## Key Takeaways

- Testing a geographic hypothesis across multiple relevant measurements (not just one) produces meaningfully stronger evidence than checking a single outcome alone.
- Cross-validating a result against an independently-reconstructed version of the data (Level 10's automation) is a valuable due-diligence step before trusting a conclusion drawn from a single source file.
- A visually noticeable spread in a boxplot is not the same as statistical significance — always check the underlying test's p-value before interpreting a visualization as evidence of a real difference.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What was the purpose of reconstructing `clinic_data` from `data/clinics/*.csv` instead of just reading `participants.csv` directly?
   a) The clinic files contain different participants
   b) It cross-validates the analysis against an independently-assembled version of the same data
   c) `participants.csv` no longer exists
   d) It is required for `aov()` to work at all

2. **True or false.** BMI's ANOVA (F = 1.614, p = 0.178) shows a statistically significant difference across counties.

3. **Predict the output.** If `identical(sort(clinic_data$id), sort(participants$id))` returns `TRUE`, what does that confirm about the two datasets?

4. **Short answer.** Why is testing both blood pressure and BMI for a county effect stronger evidence than testing blood pressure alone?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** It cross-validates the analysis against an independently-assembled version of the same data, rather than trusting a single source file alone.
2. **False.** A p-value of 0.178 is well above 0.05, indicating no statistically significant county-level difference in BMI.
3. It confirms the two datasets contain the exact same set of participant IDs — strong evidence the automated reconstruction from the clinic files matches the original data with no participants lost, duplicated, or altered.
4. A single non-significant test could plausibly reflect that specific measurement's own quirks or insufficient power; finding the same non-significant conclusion across two independently-tested, clinically distinct measurements makes it considerably less likely that a real geographic effect is being missed simply due to chance.

</details>
:::
