# Identifying Missing Values Across a Dataset {#sec-identifying-missing}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor asks a broader question than any single page in Level 1 or Level 3 answered: "Before you analyze anything, give me a complete picture — which variables in `participants.csv` have missing data, and how much?" Checking `sum(is.na(participants$age))` one column at a time, as you did in Level 1, would work — but with 8 columns, you want a way to see all of them at once.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Systematically identify missing data across an entire dataset in one step, rather than checking one column at a time.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 1, what does `is.na(x)` return for a single vector `x`?
- If a data frame is really a collection of columns (Level 2), could you imagine applying `is.na()` and `sum()` to *every* column at once, rather than typing the same check eight separate times?
:::

## The R Concept

**`colSums(is.na(df))`** applies `is.na()` to every column of a data frame at once, then sums each column's `TRUE`/`FALSE` results — giving you a missing-value count per column in a single line:

```{r}
#| eval: false
colSums(is.na(df))
```

For an even quicker, broader first look, **`summary()`** on a whole data frame reports basic statistics for every column simultaneously, including an `NA's` count for any column that has missing values:

```{r}
#| eval: false
summary(df)
```

Between the two: `colSums(is.na(df))` gives you exactly the missing-count-per-column answer your supervisor asked for; `summary()` gives you that plus a broader statistical snapshot (min, max, mean for numeric columns; counts for character columns) in one glance.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

colSums(is.na(participants))
```

```
          id          age          sex       county  systolic_bp diastolic_bp
           0            6            0            0            4            0
         bmi       smoker
           5            0
```

```{r}
#| eval: false
summary(participants)
```

```
       id             age            sex               county
 Min.   :  1.00   Min.   :18.00   Length:100         Length:100
 1st Qu.: 25.75   1st Qu.:36.25   Class :character   Class :character
 Median : 50.50   Median :49.00   Mode  :character   Mode  :character
 Mean   : 50.50   Mean   :47.43   NA's  :6
 3rd Qu.: 75.25   3rd Qu.:58.75
 Max.   :100.00   Max.   :81.00
```

## Explore

Run `colSums(is.na(participants))` and confirm the total missing cells matches what you'd get by adding the six individual `sum(is.na(participants$column))` calls from earlier levels: `6 + 4 + 5 = 15`. Then try `sum(colSums(is.na(participants)))` — a single number summarizing the *entire* dataset's missingness at once.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

colSums(is.na(participants))
```

```
          id          age          sex       county  systolic_bp diastolic_bp
           0            6            0            0            4            0
         bmi       smoker
           5            0
```

You can now answer your supervisor completely and precisely: "Three variables have missing data — `age` (6 participants), `systolic_bp` (4 participants), and `bmi` (5 participants). The other five variables — `id`, `sex`, `county`, `diastolic_bp`, and `smoker` — are fully complete."

## Interpret the Result

`colSums(is.na(participants))` reveals the missingness pattern across the *entire* dataset in one line: 15 missing cells total, concentrated in exactly three numeric variables, with no missingness at all in the categorical variables or `diastolic_bp`. This kind of complete picture — rather than checking columns one at a time and hoping you didn't miss one — is exactly what a supervisor or a methods section of a paper expects before any analysis proceeds.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Checking missingness column by column and losing track.** With more than a handful of variables, it's easy to forget to check one — `colSums(is.na(df))` checks all of them, guaranteed, every time.
- **Confusing `colSums()` with `rowSums()`.** `colSums(is.na(df))` counts missing values *per column* (down each column); `rowSums(is.na(df))` counts missing values *per row* (across each row) — genuinely different questions, both useful, easy to mix up by name alone.
- **Treating `summary()`'s output as the full picture.** It's an excellent quick overview, but for character columns it only reports `Length`, `Class`, and `Mode` — not the counts or categories `table()` would show. Use both tools, not just one.
:::

## Practice Problem

Run `rowSums(is.na(participants))` and then `table(rowSums(is.na(participants)))` — this tells you how many participants have 0 missing values, how many have exactly 1, and so on. How many participants have zero missing data across all 8 variables?

## Challenge Problem

Using `colSums(is.na(participants))`, identify which single variable has the *most* missing data, and calculate what percentage of the 100 participants that represents (using arithmetic from Level 1).

## Key Takeaways

- `colSums(is.na(df))` reports the number of missing values in every column of a data frame at once, replacing repetitive one-column-at-a-time checks.
- `summary(df)` gives a broader statistical snapshot of every column simultaneously, including an `NA's` count where relevant.
- `rowSums(is.na(df))` answers a different, complementary question: how much missing data does each individual *row* (participant) have.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `colSums(is.na(df))` return?
   a) The total number of rows in `df`
   b) The number of missing values in each column
   c) The number of missing values in each row
   d) A logical vector the same size as `df`

2. **True or false.** `colSums()` and `rowSums()` answer the same question.

3. **Predict the output.** If a data frame has 3 columns and column `x` has 2 missing values while `y` and `z` have none, what would `colSums(is.na(df))` show for `x`?

4. **Write the code.** Write the single line of code to check missing-value counts across every column of a data frame called `survey_data`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The number of missing values in each column.
2. **False.** `colSums()` summarizes down each column (per variable); `rowSums()` summarizes across each row (per observation) — different, complementary questions.
3. `2`
4. `colSums(is.na(survey_data))`

</details>
:::
