# Removing Missing Values {#sec-removing-missing}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor wants a "complete case" analysis: every calculation should use only participants who have *no* missing data across `age`, `systolic_bp`, and `bmi` at all — not just excluding missing values one column at a time with `na.rm = TRUE`, but removing entire *rows* wherever any of the three variables is missing. `na.rm = TRUE` (Level 1) only ever worked within a single `mean()` call — it never actually removed a row from the dataset itself.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Remove rows with missing data from a data frame, either across specific columns or completely.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 2, how would you build a logical vector flagging which *rows* have a missing value in a specific column?
- If a row is missing `age` but has every other variable, should it be excluded from an analysis of `bmi` that has nothing to do with `age`? Does your answer depend on the specific question being asked?
:::

## The R Concept

**`na.omit()`** removes any row with *at least one* missing value, across *every* column of a data frame:

```{r}
#| eval: false
na.omit(df)
```

**`complete.cases()`** gives you more control — it returns a logical vector flagging which rows have no missing data, which you can then use with the row-indexing you learned in Level 2, optionally restricted to just the columns you care about:

```{r}
#| eval: false
complete.cases(df)                    # TRUE for rows complete across ALL columns
complete.cases(df[, c("age", "bmi")]) # TRUE for rows complete in just these two columns
df[complete.cases(df), ]              # keep only the fully complete rows
```

The distinction matters: `na.omit()` is quick but *all-or-nothing* across every column; `complete.cases()` lets you decide exactly which columns' completeness actually matters for the analysis at hand — precisely your supervisor's request.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

nrow(participants)
```

```
[1] 100
```

```{r}
#| eval: false
# na.omit() - removes a row if ANY column has a missing value
complete_all <- na.omit(participants)
nrow(complete_all)
```

```
[1] 85
```

```{r}
#| eval: false
# complete.cases() restricted to just the 3 columns that matter for this analysis
keep <- complete.cases(participants[, c("age", "systolic_bp", "bmi")])
complete_relevant <- participants[keep, ]
nrow(complete_relevant)
```

```
[1] 85
```

## Explore

These two approaches happen to agree here (85 rows) because `age`, `systolic_bp`, and `bmi` are the *only* columns with any missingness at all in this dataset (Level 4's first page confirmed `id`, `sex`, `county`, `diastolic_bp`, and `smoker` are fully complete). Try `nrow(na.omit(participants[, c("sex", "county")]))` — since neither column has any missing data, what would you expect, and does the result confirm it?

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

# Complete-case analysis restricted to exactly the variables your supervisor cares about
keep <- complete.cases(participants[, c("age", "systolic_bp", "bmi")])
participants_complete <- participants[keep, ]

nrow(participants_complete)
```

```
[1] 85
```

```{r}
#| eval: false
mean(participants_complete$age)
```

```
[1] 47.22353
```

Notice `mean()` no longer needs `na.rm = TRUE` at all here — every remaining row is already guaranteed complete for these three variables, so there's nothing left for `na.rm` to exclude.

## Interpret the Result

85 of the 100 participants have complete data across `age`, `systolic_bp`, and `bmi` — the other 15 are missing at least one of the three. Restricting to exactly these three columns (rather than `na.omit()`'s all-or-nothing approach across every column) matters in general, though here the two approaches happen to produce the same 85 rows since no other columns in this particular dataset have any missing values.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Using `na.omit()` on a whole data frame when only a few columns matter.** If a dataset had missingness in a column irrelevant to your analysis, `na.omit()` would still drop those rows unnecessarily — `complete.cases()` restricted to relevant columns avoids discarding usable data.
- **Losing track of how many rows were dropped.** Always compare `nrow()` before and after — silently going from 100 to 85 rows without noticing is a common way to accidentally misreport a sample size in a report.
- **Forgetting that removing rows changes more than one column at once.** Once you filter to complete cases for `age`, `systolic_bp`, and `bmi`, every *other* column (like `smoker`) also loses those same 15 rows — even though `smoker` itself had no missing data.
:::

## Practice Problem

Using `complete.cases()`, create a subset of `participants` that is complete for `bmi` alone, and report how many rows remain (it should match the earlier missingness count from the previous page).

## Challenge Problem

Compare `nrow(na.omit(participants))` to `nrow(participants[complete.cases(participants[, c("age", "systolic_bp", "bmi")]), ])`. They should match in this dataset — explain, in one or two sentences, *why* they match here, and describe a hypothetical change to the dataset that would make them produce different results.

## Key Takeaways

- `na.omit()` removes any row with missing data in *any* column; `complete.cases()` lets you restrict that check to just the columns relevant to your analysis.
- Always compare row counts before and after removing missing data, and report how many rows (and why) were excluded.
- Removing rows for missingness in one variable also removes those same rows' data for every other variable — a side effect worth being deliberate about.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `na.omit(df)` do?
   a) Replaces missing values with 0
   b) Removes any row with at least one missing value, across all columns
   c) Removes only missing values in numeric columns
   d) Removes columns with missing data

2. **True or false.** `complete.cases()` can be restricted to check completeness in just a subset of columns.

3. **Predict the output.** If `df` has 50 rows and 5 have missing `age`, and no other column has missing data, what would `nrow(na.omit(df))` return?

4. **Write the code.** Write the code to keep only the rows of `survey_data` that are complete for the column `income`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Removes any row with at least one missing value, across all columns.
2. **True.** `complete.cases(df[, c("col1", "col2")])` restricts the completeness check to just those columns.
3. `45`
4. `survey_data[complete.cases(survey_data[, "income"]), ]` (or equivalently, `survey_data[!is.na(survey_data$income), ]`)

</details>
:::
