# Removing Duplicates {#sec-removing-duplicates}

::: {.pbl-box .pbl-problem}
**The Problem**

You've confirmed participant #1 appears twice in the dataset, thanks to the last page's `duplicated()` check. Simply *knowing* about the duplicate isn't enough — your supervisor needs the dataset fixed, with exactly one copy of every participant, before any analysis proceeds.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Remove duplicate rows from a data frame, keeping exactly one copy of each.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From the last page, `duplicated()` returns `FALSE` for the first occurrence of a row and `TRUE` for every repeat. If you kept only the rows where this is `FALSE`, what would remain?
- Is there a shortcut function you might expect, given that Level 2 already introduced a function called `unique()` for vectors?
:::

## The R Concept

Two equivalent approaches remove duplicate rows, keeping the first occurrence of each:

```{r}
#| eval: false
df[!duplicated(df), ]   # keep only rows where duplicated() is FALSE
unique(df)               # shortcut: does exactly the same thing for a whole data frame
```

`unique()`, which you first met on vectors in Level 2, works the same way on entire data frames — it's really just a more convenient wrapper around `!duplicated(df)`.

```{r}
#| eval: false
nrow(df)                  # before
nrow(unique(df))          # after - should be nrow(df) - sum(duplicated(df))
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

participants_dup <- rbind(participants, participants[1, ])
nrow(participants_dup)
```

```
[1] 101
```

```{r}
#| eval: false
# Approach 1: explicit logical indexing
participants_fixed <- participants_dup[!duplicated(participants_dup), ]
nrow(participants_fixed)
```

```
[1] 100
```

```{r}
#| eval: false
# Approach 2: the unique() shortcut - identical result
participants_fixed2 <- unique(participants_dup)
nrow(participants_fixed2)
```

```
[1] 100
```

```{r}
#| eval: false
identical(participants_fixed, participants_fixed2)
```

```
[1] TRUE
```

## Explore

Confirm `identical(participants_fixed, participants)` — after removing the duplicate you added, does the dataset exactly match the original, untouched `participants`? This confirms the fix is complete and introduced no side effects beyond removing the one duplicate row.

## Solve the Problem

```{r}
#| eval: false
participants_dup <- rbind(participants, participants[1, ])

nrow(participants_dup)
```

```
[1] 101
```

```{r}
#| eval: false
participants_clean <- unique(participants_dup)

nrow(participants_clean)
```

```
[1] 100
```

```{r}
#| eval: false
sum(duplicated(participants_clean))
```

```
[1] 0
```

Exactly 100 rows remain — one copy of every participant, with the accidental double-entry from the two research assistants fully resolved.

## Interpret the Result

`nrow(participants_clean)` returning to `100` (down from the temporarily inflated `101`), combined with `sum(duplicated(participants_clean))` confirming `0` remaining duplicates, gives you concrete proof the dataset is now exactly as it should be — ready to hand back to your supervisor with confidence, not just a hopeful assumption.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Removing duplicates before confirming which rows they actually are.** Always run the detection step (previous page) first, so you know *what* you're removing and can sanity-check the before/after row counts make sense.
- **Assuming `unique()` and `duplicated()` disagree in some way.** They don't — `unique(df)` is defined to produce exactly the same result as `df[!duplicated(df), ]`; either is correct, and which one you reach for is purely a matter of style.
- **Not rechecking after removal.** Running `sum(duplicated(participants_clean))` afterward and confirming it's `0` is a cheap, valuable habit — don't just assume the removal worked because no error appeared.
:::

## Practice Problem

Create `test_dup <- rbind(participants, participants[50, ], participants[50, ])` (duplicating row 50 *twice*, so it appears three times total). Confirm `nrow(test_dup)` is `102`, then use `unique()` to reduce it back to `100`, and verify with `sum(duplicated(...))`.

## Challenge Problem

Recall the previous page's challenge: a duplicate participant with one *differing* value (like a corrected age) would not be caught by `duplicated()` or `unique()`, since they only catch exact row matches. Describe, in words, what additional check based on `id` alone (ignoring the other columns) would be needed to catch this trickier kind of duplicate — you don't need to write the code, just describe the logic.

## Key Takeaways

- `df[!duplicated(df), ]` and `unique(df)` are equivalent ways to remove duplicate rows, keeping the first occurrence of each.
- Always detect and confirm duplicates *before* removing them, and reconfirm with `sum(duplicated(...))` afterward that none remain.
- Both approaches only catch exact row matches — a "duplicate" participant with even one differing value requires a different, ID-based check (not covered by `duplicated()` alone).

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `unique(df)` do?
   a) Removes all rows with missing data
   b) Removes duplicate rows, keeping the first occurrence of each
   c) Removes the first occurrence of any repeated row
   d) Sorts the data frame

2. **True or false.** `df[!duplicated(df), ]` and `unique(df)` produce the same result.

3. **Predict the output.** If `df` has 20 rows and 3 are exact duplicates of earlier rows, what would `nrow(unique(df))` return?

4. **Write the code.** Write the code to remove duplicate rows from a data frame called `survey_data`, using the `unique()` shortcut.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Removes duplicate rows, keeping the first occurrence of each.
2. **True.** They are equivalent — `unique()` is a convenience wrapper around the same logic.
3. `17`
4. `survey_data <- unique(survey_data)`

</details>
:::
