# Combining Categories {#sec-combining-categories}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor wants a simpler regional comparison than five separate counties allow: "Group Cedar Falls and Millbrook as 'North Region,' and Oakford, Riverdale, and Springvale as 'South Region.' I don't need county-level detail for this particular report — I need exactly two regions." This is a genuinely different operation from recoding: several *different* original categories all need to collapse into the *same* new category.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Collapse several existing categories into a smaller number of broader groups.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From the recoding page, you mapped each old category to exactly one new label, one-to-one. What's different about mapping *several* old categories to the *same* new label?
- Could you use `%in%` (checking whether a value is one of several options) instead of a single `==` comparison, to flag "any of these counties belongs to the North Region"?
:::

## The R Concept

Combining categories extends the recoding pattern from the last page, using **`%in%`** to check membership in a *set* of values rather than equality with just one:

```{r}
#| eval: false
x %in% c("value1", "value2")   # TRUE if x matches ANY of these, not just one
```

```{r}
#| eval: false
df$new_col[df$old_col %in% c("CategoryA", "CategoryB")] <- "Combined Group"
```

This is the natural extension of `==` (from the recoding page) to "matches one of several things" rather than "matches exactly this one thing" — precisely what's needed when multiple original categories all collapse into the same broader group.

## Demonstration

```{r}
#| eval: false
county <- c("Cedar Falls", "Millbrook", "Oakford", "Riverdale", "Springvale", "Cedar Falls")

# Check membership in a set, rather than equality with a single value
county %in% c("Cedar Falls", "Millbrook")
```

```
[1]  TRUE  TRUE FALSE FALSE FALSE  TRUE
```

```{r}
#| eval: false
region <- county   # start from a copy, same pattern as recoding
region[county %in% c("Cedar Falls", "Millbrook")] <- "North Region"
region[county %in% c("Oakford", "Riverdale", "Springvale")] <- "South Region"

region
```

```
[1] "North Region" "North Region" "South Region" "South Region" "South Region"
[6] "North Region"
```

## Explore

Run `table(county, region)` — a cross-tabulation confirming every one of the five counties mapped into exactly one of the two intended regions, with no county left unmapped and no county mapped to both.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

participants$region <- participants$county
participants$region[participants$county %in% c("Cedar Falls", "Millbrook")] <- "North Region"
participants$region[participants$county %in% c("Oakford", "Riverdale", "Springvale")] <- "South Region"

table(participants$region)
```

```
North Region South Region
          41           59
```

```{r}
#| eval: false
# Verify every county mapped correctly and completely
table(participants$county, participants$region)
```

```
              North Region South Region
  Cedar Falls           20            0
  Millbrook             21            0
  Oakford                0           17
  Riverdale              0           16
  Springvale             0           26
```

Every county correctly collapsed into exactly one of the two intended regions — 41 participants in the North Region, 59 in the South Region — with the cross-tabulation confirming no county was missed or double-counted.

## Interpret the Result

The two-way table's diagonal-block pattern — Cedar Falls and Millbrook entirely in the North Region column, Oakford, Riverdale, and Springvale entirely in the South Region column, with zero counts everywhere else — confirms the regional grouping is complete and correct, exactly matching your supervisor's five-counties-into-two-regions request.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Using `==` instead of `%in%` for a multi-category match.** `county == c("Cedar Falls", "Millbrook")` does not do what you might expect — it compares element-by-element in a way that rarely matches your intent; `%in%` is the correct tool for "matches any of these."
- **Leaving a category unmapped by accident.** If you forget to include one of the five counties in either group, those rows keep their original county name in the new `region` column, silently mixed in among the properly grouped values — the cross-tabulation check catches this immediately.
- **Assuming combining categories is the same operation as recoding.** Recoding (previous page) is one-to-one; combining categories is many-to-one — the logic (`%in%` vs. `==`) reflects that real difference.
:::

## Practice Problem

Combine `participants$county` into three regions instead of two: `"Cedar Falls"` and `"Millbrook"` as `"North"`, `"Oakford"` and `"Riverdale"` as `"Central"`, and `"Springvale"` alone as `"East"`. Verify with a cross-tabulation.

## Challenge Problem

Your supervisor later asks for a *different* grouping of the same five counties for a separate report. Explain, in a sentence or two, why storing the *first* regional grouping in a new column (`region`) rather than overwriting `county` directly makes it easy to add this second, different grouping (perhaps `region_v2`) without losing either version or re-deriving anything from scratch.

## Key Takeaways

- `%in%` checks membership in a set of values, extending the single-value `==` comparison from the recoding page to "matches any of these."
- Combining categories is a many-to-one operation — several original categories collapse into one broader group — distinct from recoding's one-to-one relabeling.
- A two-way `table(old_column, new_column)` cross-tabulation confirms every original category was mapped exactly once, with nothing missed or double-mapped.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `x %in% c("A", "B")` check?
   a) Whether `x` equals `"A"` and `"B"` simultaneously
   b) Whether `x` matches either `"A"` or `"B"`
   c) Whether `x` is alphabetically between `"A"` and `"B"`
   d) Whether `x` is missing

2. **True or false.** Combining categories and recoding a variable are the same operation.

3. **Predict the output.**
   ```r
   x <- c("Red", "Blue", "Green")
   x %in% c("Red", "Green")
   ```

4. **Write the code.** Write the code to combine `"Yes"` and `"Sometimes"` into a single category `"Any use"` in a vector called `alcohol_use`, leaving `"No"` unchanged.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Whether `x` matches either `"A"` or `"B"`.
2. **False.** Recoding is one-to-one; combining categories is many-to-one — a genuinely different mapping.
3. `[1]  TRUE FALSE  TRUE`
4.
   ```r
   alcohol_use[alcohol_use %in% c("Yes", "Sometimes")] <- "Any use"
   ```

</details>
:::
