# `if_else()` {#sec-if-else-dplyr}

::: {.pbl-box .pbl-problem}
**The Problem**

A colleague's script uses `ifelse(is_followup, visit_date, NA)` to blank out a date column for baseline visits — and it silently converts every proper date into a plain number (the internal representation of a date), breaking every later step that expected an actual date. Nothing in `ifelse()` warned them this would happen.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand `if_else()`, the tidyverse's stricter version of `ifelse()`, and why it prevents exactly this kind of silent type-conversion bug.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 2, what does R do when you mix two different types together, like a number and a piece of text, inside `c()`?
- If `ifelse()`'s second and third arguments (the `TRUE` and `FALSE` results) were of two genuinely different types, what do you think should happen — a clear error, or a silent, best-effort guess?
:::

## The R Concept

**`if_else()`**, from the `dplyr` package (part of the tidyverse, arriving properly in Level 5), works exactly like `ifelse()` for straightforward cases, but is **stricter about types**: if the `TRUE` result and the `FALSE` result aren't compatible types, `if_else()` raises a clear error immediately, rather than silently coercing one of them into something unexpected.

```{r}
#| eval: false
library(dplyr)
if_else(condition, value_if_true, value_if_false)
```

```{r}
#| eval: false
# ifelse() silently converts dates to plain numbers - a common, hard-to-spot bug
ifelse(is_followup, visit_date, NA)

# if_else() raises a clear, immediate error instead of silently corrupting the type
if_else(is_followup, visit_date, NA)
```

The fix, once you know about the problem, is usually to make the `NA` explicitly typed to match: `if_else(is_followup, visit_date, as.Date(NA))` for a date column, for instance — `if_else()` forces you to be explicit rather than letting a mismatch pass silently.

## Demonstration

```{r}
#| eval: false
library(dplyr)

# A straightforward case - if_else() behaves identically to ifelse()
ages <- c(72, 45, 68)
if_else(ages >= 65, "Senior", "Non-senior")
```

```
[1] "Senior"     "Non-senior" "Senior"
```

```{r}
#| eval: false
# A type-mismatch case - ifelse() silently causes a problem, if_else() catches it
is_followup <- c(TRUE, FALSE, TRUE)
visit_date <- as.Date(c("2026-01-05", "2026-01-06", "2026-01-07"))

ifelse(is_followup, visit_date, NA)
```

```
[1] 20458    NA 20460
```

Notice the dates silently became plain numbers (each date's internal numeric representation) — exactly the bug from the opening problem.

```{r}
#| eval: false
if_else(is_followup, visit_date, NA)
```

```
Error in `if_else()`:
! `false` must be a <Date>, not `NA`.
```

`if_else()` refuses to guess, and instead tells you *exactly* what went wrong — the `FALSE` result must be a proper `Date`, not a bare, untyped `NA`.

## Explore

Fix the error by explicitly typing the `NA`: `if_else(is_followup, visit_date, as.Date(NA))`. Confirm this now works correctly, preserving the `Date` type throughout, with no silent conversion to plain numbers.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

is_followup <- c(TRUE, FALSE, TRUE)
visit_date <- as.Date(c("2026-01-05", "2026-01-06", "2026-01-07"))

visit_date_clean <- if_else(is_followup, visit_date, as.Date(NA))
visit_date_clean
```

```
[1] "2026-01-05" NA           "2026-01-07"
```

```{r}
#| eval: false
class(visit_date_clean)
```

```
[1] "Date"
```

The colleague's bug is now fixed and, more importantly, made impossible to silently reintroduce — `if_else()` would immediately raise an error again if the types ever mismatched in the future, rather than quietly corrupting the column the way `ifelse()` did.

## Interpret the Result

`class(visit_date_clean)` correctly reports `"Date"`, confirming the column's type was preserved throughout — the baseline visit correctly shows `NA` (a real, typed missing date) rather than the number `20459` (or similar) that `ifelse()` would have silently produced.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Assuming `if_else()` and `ifelse()` are simply two names for the same thing.** They agree on simple cases, but `if_else()`'s stricter type-checking is a deliberate, meaningful difference — not a stylistic quirk.
- **Being annoyed by an `if_else()` error instead of reading what it's telling you.** The error message in this page's demonstration says exactly what's wrong (`false must be a <Date>, not NA`) — it's surfacing a real problem `ifelse()` would have hidden, not being unnecessarily strict for no reason.
- **Forgetting to load `dplyr`.** Like any package function, `if_else()` requires `library(dplyr)` (or `library(tidyverse)`) first.
:::

## Practice Problem

Using `if_else()`, recreate the `age_group` variable from the previous page (`"Senior"` for 65+, `"Non-senior"` otherwise) on `participants$age`, and confirm it produces the same result as the `ifelse()` version.

## Challenge Problem

Explain, in two or three sentences, why a strict, error-raising function like `if_else()` is generally *safer* for production analysis code than a permissive, silently-converting one like `ifelse()` — even though the strict version is sometimes more annoying to work with in the moment.

## Key Takeaways

- `if_else()` (from `dplyr`) behaves like `ifelse()` on straightforward cases, but raises a clear error rather than silently converting types when the `TRUE` and `FALSE` results don't match.
- This strictness catches real bugs — like a `Date` column silently becoming plain numbers — immediately, rather than letting them surface much later and harder to trace.
- When `if_else()` errors on a type mismatch, the fix is usually to make both results explicitly the same type (like `as.Date(NA)` instead of a bare `NA`).

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `if_else()` do differently from `ifelse()`?
   a) It works only on numeric data
   b) It raises an error rather than silently converting mismatched types
   c) It cannot handle missing values
   d) It requires no package to be loaded

2. **True or false.** `ifelse()` can silently convert a `Date` column into plain numbers.

3. **Predict the output.**
   ```r
   if_else(c(TRUE, FALSE), "Yes", "No")
   ```

4. **Find the error.**
   ```r
   if_else(is_followup, visit_date, NA)
   ```
   where `visit_date` is a `Date` column.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** It raises an error rather than silently converting mismatched types.
2. **True.** This is exactly the bug `if_else()` is designed to catch and prevent.
3. `[1] "Yes" "No"`
4. The bare `NA` doesn't match the `Date` type of `visit_date`; it needs to be `as.Date(NA)` instead, so both possible results share the same type.

</details>
:::
