# `case_when()` {#sec-case-when}

::: {.pbl-box .pbl-problem}
**The Problem**

Back in Level 2, you built a four-category `bmi_category` variable using three *nested* `ifelse()` calls: `ifelse(bmi < 18.5, "Underweight", ifelse(bmi < 25, "Normal", ifelse(bmi < 30, "Overweight", "Obese")))`. It worked, but even at four categories it was already hard to read at a glance — and your supervisor now wants a five-category blood pressure classification. Nesting `ifelse()` four levels deep would be close to unreadable.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Use `case_when()` to handle several conditions at once, cleanly, without nesting `ifelse()` calls inside each other.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If you had five conditions to check in order, would you rather read them as five separate, parallel lines, or as `ifelse()` nested four levels deep?
- What should happen if a row matches none of your listed conditions — should it get a value at all, or should something else happen?
:::

## The R Concept

**`case_when()`**, from `dplyr` (arriving properly in Level 5), evaluates a series of conditions **in order**, top to bottom, and assigns the value from the **first** one that matches:

```{r}
#| eval: false
library(dplyr)
case_when(
  condition1 ~ value1,
  condition2 ~ value2,
  condition3 ~ value3,
  TRUE ~ default_value
)
```

Each line reads naturally: "condition `~` (results in) value." The final `TRUE ~ default_value` line acts as a catch-all — since `TRUE` always matches, it captures any row that didn't match an earlier condition (equivalent to the final `else` in a nested `ifelse()` chain, but far more readable at a glance).

```{r}
#| eval: false
case_when(
  bmi < 18.5 ~ "Underweight",
  bmi < 25   ~ "Normal",
  bmi < 30   ~ "Overweight",
  TRUE       ~ "Obese"
)
```

Notice this reads almost like a sentence: "if BMI is under 18.5, Underweight; if under 25 (having already excluded under-18.5), Normal;" and so on — each condition is only evaluated for rows that didn't match anything above it.

## Demonstration

```{r}
#| eval: false
library(dplyr)

bmi_values <- c(17.2, 22.5, 28.9, 33.1, NA)

case_when(
  bmi_values < 18.5 ~ "Underweight",
  bmi_values < 25   ~ "Normal",
  bmi_values < 30   ~ "Overweight",
  TRUE              ~ "Obese"
)
```

```
[1] "Underweight" "Normal"      "Overweight"  "Obese"       NA
```

Notice the `NA` BMI correctly produces `NA` in the result — even the catch-all `TRUE ~ "Obese"` doesn't fire for a missing input, because `case_when()` (like `ifelse()` and `if_else()`) correctly propagates missingness rather than guessing.

Compare this to the equivalent, harder-to-read nested `ifelse()` from Level 2:

```{r}
#| eval: false
ifelse(bmi_values < 18.5, "Underweight",
  ifelse(bmi_values < 25, "Normal",
    ifelse(bmi_values < 30, "Overweight", "Obese")))
```

Both produce the same result — `case_when()` simply reads far more clearly once there are more than two categories involved.

## Explore

Try adding a fifth category to the blood pressure classification: `"Normal"` (< 120), `"Elevated"` (120–129), `"Hypertension Stage 1"` (130–139), `"Hypertension Stage 2"` (140–179), and `"Hypertensive Crisis"` (180+), using `case_when()` on `participants$systolic_bp`. Notice how each additional category is just one more readable line, rather than another level of nesting.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

participants$bp_category <- case_when(
  participants$systolic_bp < 120 ~ "Normal",
  participants$systolic_bp < 130 ~ "Elevated",
  participants$systolic_bp < 140 ~ "Hypertension Stage 1",
  participants$systolic_bp < 180 ~ "Hypertension Stage 2",
  TRUE                            ~ "Hypertensive Crisis"
)

table(participants$bp_category, useNA = "always")
```

```
            Elevated                Normal Hypertension Stage 1
                  18                    32                    17
Hypertension Stage 2   Hypertensive Crisis                  <NA>
                  29                     0                     4
```

Five clinically meaningful categories, built in five clean, parallel lines — no nested nightmare of nine mismatched parentheses to debug, and the 4 participants with missing `systolic_bp` correctly show `NA` rather than being forced into any category.

## Interpret the Result

The blood pressure classification is now complete across all five clinical categories, with the missing-data behavior correctly preserved (4 participants with missing `systolic_bp` remain `NA`, not defaulted into `"Normal"` or any other category) — exactly the behavior you'd want, and considerably easier to read, extend, or hand off to a colleague than an equivalent nested `ifelse()` chain would have been.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting conditions are checked in order, top to bottom.** If you accidentally list a broader condition before a narrower one (like `systolic_bp < 180` before `systolic_bp < 120`), the narrower condition below it will *never* be reached, since the broader one already matched first.
- **Forgetting the final `TRUE ~ default` catch-all.** Without it, any row that doesn't match any listed condition becomes `NA` in the result — sometimes exactly what you want, but often an unintentional gap if you forgot a category.
- **Using `=` instead of `~`.** `case_when()`'s syntax specifically requires the tilde (`~`), not an equals sign — a common typo for anyone used to other assignment-style syntax in R.
:::

## Practice Problem

Using `case_when()`, classify `participants$age` into `"Child"` (under 18), `"Adult"` (18–64), and `"Senior"` (65+). Confirm the result with `table(..., useNA = "always")`.

## Challenge Problem

Rebuild Level 2's four-category `bmi_category` variable using `case_when()` instead of nested `ifelse()`, and confirm with `table()` that it produces exactly the same category counts as Level 2's original version (3 Underweight, 28 Normal, 39 Overweight, 25 Obese).

## Key Takeaways

- `case_when()` evaluates conditions in order and assigns the value from the first match — far more readable than nesting `ifelse()` calls for three or more categories.
- A final `TRUE ~ default_value` line acts as a catch-all for any row that matched nothing above it.
- Like `ifelse()` and `if_else()`, `case_when()` correctly propagates missing values — a missing input never gets forced into the catch-all category.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In `case_when()`, what does the final `TRUE ~ "default"` line do?
   a) Nothing; it's optional decoration
   b) Catches any row that didn't match an earlier condition
   c) Overrides every other condition
   d) Only applies to missing values

2. **True or false.** `case_when()` conditions are checked in order, and the first match wins.

3. **Predict the output.**
   ```r
   x <- c(5, 15, 25, NA)
   case_when(
     x < 10 ~ "low",
     x < 20 ~ "medium",
     TRUE   ~ "high"
   )
   ```

4. **Find the error.**
   ```r
   case_when(
     age = "Senior" ~ age >= 65,
     TRUE ~ "Non-senior"
   )
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Catches any row that didn't match an earlier condition.
2. **True.** Conditions are evaluated top to bottom, and the first `TRUE` match determines the result.
3. `[1] "low"    "medium" "high"   NA`
4. The condition and value are reversed, and `=` is used instead of `~`. It should be `age >= 65 ~ "Senior"`.

</details>
:::

## Where you've arrived

You started this level with a vague sense that real data needs fixing before it can be trusted, and a stack of tools scattered across Level 1 through Level 3 for spotting problems. You've now built a complete, systematic toolkit for actually *fixing* what you find: identifying, removing, and replacing missing values; detecting and removing duplicates; renaming and retyping variables in every direction; building new variables and recoding, combining, and binarizing categories; and — across these last four pages — the full progression from a single `if` decision to `ifelse()`'s vectorized version, `if_else()`'s type-safe version, and `case_when()`'s clean, multi-condition version.

That last tool, `case_when()`, is also your bridge into Level 5: it's a `dplyr` function, and Level 5 is where you meet the rest of the tidyverse — `filter()`, `mutate()`, `summarise()`, and the pipe `|>` — that will make everything you've done so far in base R read even more like the sentence you'd say the question in out loud.
