# Installing and Loading Packages {#sec-packages}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor emails you an Excel file (`.xlsx`) containing the community health survey data, not a CSV. You try to open it in R the way you've seen in a tutorial, and you get an error: `could not find function "read_excel"`. The tutorial clearly used that exact function.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand what an R package is, and install and load one so its functions become available to you.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If `read_excel` isn't a function R comes with automatically, where might it come from?
- Have you ever used an app that required you to download a separate "add-on" or "extension" before a feature worked? How is that similar to what might be happening here?
:::

## The R Concept

Base R — what you get immediately after installing R — includes a solid set of built-in functions, but it doesn't include everything. **Packages** are add-on collections of functions, data, and documentation that extend what R can do. Anyone can write and publish one; most official packages are hosted on **CRAN** (the Comprehensive R Archive Network).

Working with a package is a two-step process, and beginners often only do one of the two steps:

1. **Install it — once per computer.**
   ```r
   install.packages("readxl")
   ```
   This downloads the package and stores it on your machine permanently. You do not need to repeat this every time you use R — only when installing for the first time, or updating later.

2. **Load it — once per R session.**
   ```r
   library(readxl)
   ```
   This tells your *current* R session "make this package's functions available right now." Every time you close and reopen RStudio, you need to load the package again with `library()` — but you do **not** need to reinstall it.

Notice the quoting difference: `install.packages()` requires the package name in quotes (`"readxl"`), because it's just a piece of text identifying what to download. `library()` conventionally doesn't require quotes, though quoting it also works.

## Demonstration

```{r}
#| eval: false
# Step 1: install the package (only needs to be run once, ever, per computer)
install.packages("readxl")
```

```{r}
#| eval: false
# Step 2: load the package (needs to be run once per R session)
library(readxl)
```

Once loaded, functions from that package — like `read_excel()` — become available exactly as if they were built into R.

```{r}
#| eval: false
# Now this function exists and is ready to use
survey_data <- read_excel("data/survey.xlsx")
```

You can check which packages are already installed on your machine without loading anything:

```{r}
#| eval: false
# Lists every package installed locally (long output - often dozens or hundreds of rows)
installed.packages()[, "Package"] |> head()
```

## Explore

What do you think happens if you run `library(readxl)` *without* ever running `install.packages("readxl")` first, on a computer where it's never been installed? Try to predict the error message before you test it — it will say something to the effect of "there is no package called 'readxl'."

## Solve the Problem

The `could not find function "read_excel"` error happened because the tutorial's code assumed the `readxl` package was already loaded — but loading it requires two things to have happened first: the package must be *installed* on your computer, and `library()` must have been *run in your current session*. Since you'd done neither, R had no idea what `read_excel` meant.

The fix:

```{r}
#| eval: false
install.packages("readxl")  # only needed once
library(readxl)              # needed every session
survey_data <- read_excel("data/survey.xlsx")
```

## Interpret the Result

Once `library(readxl)` runs without error, it means the package's functions — including `read_excel()` — are now available in your current session. If `install.packages()` had failed (for example, due to no internet connection), `library()` would fail too, since there'd be nothing installed to load.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting to load the package each new session.** Installing is permanent; loading is not. A fresh RStudio session needs `library()` again even if you installed the package last week.
- **Reinstalling every time.** Running `install.packages()` at the top of every script is unnecessary and slow — install once, then just `library()` in future sessions.
- **Quoting inconsistency.** Forgetting the quotes in `install.packages(readxl)` (missing quotes) causes an error, since R tries to interpret `readxl` as an object rather than a package name.
- **Typos in package names.** `install.packages("read_xl")` will fail silently with a "package not available" message if the actual package is spelled `readxl` (no underscore).
:::

## Practice Problem

Write the two lines of code needed to install and then load a package called `janitor` (a real, commonly used package for cleaning messy data), assuming it has never been installed on this computer before.

## Challenge Problem

A classmate says their script works perfectly on their own laptop but produces `could not find function` errors on yours, even though you're both running the identical script. List two possible explanations related to what you learned on this page, and how you'd check each one.

## Key Takeaways

- Packages extend R with additional functions, data, and tools beyond what comes built in.
- `install.packages("name")` downloads a package once per computer; `library(name)` loads it once per session.
- Forgetting `library()` — not forgetting to reinstall — is the far more common cause of "could not find function" errors in day-to-day work.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Which line only needs to be run once, ever, on a given computer?
   a) `library(readxl)`
   b) `install.packages("readxl")`
   c) Both need to run every session
   d) Neither needs quotes

2. **True or false.** If you restart RStudio, previously installed packages must be reinstalled before you can use them again.

3. **Find the error.**
   ```r
   install.packages(readxl)
   ```

4. **Write the code.** Write the two lines needed to install and load a package called `lubridate`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `install.packages("readxl")` — installation is a one-time, per-computer action.
2. **False.** Installed packages remain installed; you only need to run `library()` again to load them into the new session.
3. The package name is missing quotes. It should be `install.packages("readxl")`.
4. `install.packages("lubridate")` then `library(lubridate)`.

</details>
:::
