# Objects {#sec-objects}

::: {.pbl-box .pbl-problem}
**The Problem**

Halfway through a script, you've now created `weight_kg`, `height_m`, `bmi`, `is_overweight`, and `bmi_rounded`. Your supervisor walks by and asks: "What exactly is stored in R right now?" You glance at your script but realize the script shows what you *ran*, not necessarily what currently *exists* in memory.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand what an R "object" is in general, and learn to inspect everything currently stored in your session.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Is there a difference between the *code that created* something and the *thing that now exists* as a result?
- If you assigned ten different values across a long script, how would you check, right now, what's actually available to use?
:::

## The R Concept

In R, virtually everything you create and store is called an **object**. A single number, a block of text, a list of a hundred ages, an entire imported spreadsheet — all of these become objects once assigned a name with `<-`. This is a deliberately broad, unifying idea: whether an object holds one value or a million, you interact with it the same way — by name.

Two ways to see what objects currently exist in your session:

1. **The Environment pane** (top-right in RStudio) — lists every object live, updating automatically as you create, modify, or remove them.
2. **`ls()`** — a function that lists all object names in your current session, directly in the Console.

```{r}
#| eval: false
ls()
```

To remove an object you no longer need:

```{r}
#| eval: false
rm(bmi_rounded)
```

And to clear everything at once (useful for starting fresh, though rarely needed mid-analysis):

```{r}
#| eval: false
rm(list = ls())
```

## Demonstration

```{r}
#| eval: false
weight_kg <- 78
height_m <- 1.70
bmi <- weight_kg / height_m^2
is_overweight <- bmi >= 25 & bmi < 30
bmi_rounded <- round(bmi, 1)

# See every object currently in this session
ls()
```

```
[1] "bmi"           "bmi_rounded"   "height_m"      "is_overweight" "weight_kg"
```

Note that `ls()` returns names in alphabetical order, not the order you created them — useful to know so you don't expect chronological order.

## Explore

Run `rm(bmi_rounded)` and then `ls()` again. Confirm `bmi_rounded` is gone from the list, while the other four objects remain. Then try printing `bmi_rounded` directly — what error message does R give you?

## Solve the Problem

```{r}
#| eval: false
ls()
```

```
[1] "bmi"           "bmi_rounded"   "height_m"      "is_overweight" "weight_kg"
```

You can now answer your supervisor precisely: "Right now, I have five objects stored — `bmi`, `bmi_rounded`, `height_m`, `is_overweight`, and `weight_kg`." You didn't need to scroll back through your whole script to answer confidently — `ls()` (or the Environment pane) gave you the live, current state directly.

## Interpret the Result

The output of `ls()` is not the *code* that built these objects — it's the *current inventory* of what those lines of code produced and left behind in memory. This is an important distinction: a script shows history; `ls()` and the Environment pane show present state.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Assuming the script and the Environment always match.** If you delete a line from your script but already ran it, the object it created still exists in your session until you explicitly remove it or restart R.
- **Using an object that "shouldn't" exist anymore.** Because R doesn't automatically clean up after you, leftover objects from earlier experimentation can accidentally get reused later in a script, producing confusing results.
- **Confusing `rm()` with deleting a file.** `rm(bmi)` removes the *object* from your current session's memory — it has no effect on any saved script or data file.
:::

## Practice Problem

Create three objects: `n_enrolled <- 250`, `n_completed <- 211`, and `site_name <- "Riverdale Clinic"`. Run `ls()` and confirm all three appear. Then remove `site_name` with `rm()` and run `ls()` again to confirm it's gone.

## Challenge Problem

Without running any code, predict what `ls()` would return immediately after a fresh RStudio session starts, before you've created any objects. Then verify your prediction is correct — reason through what an empty environment should report.

## Key Takeaways

- Nearly everything you create in R — numbers, text, entire datasets — is stored as an object once assigned a name.
- `ls()` and the Environment pane both show what currently exists in your session, which can differ from what your script currently contains.
- `rm()` removes an object from memory but never touches any saved file.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `ls()` show?
   a) The lines of code in your current script
   b) The names of all objects currently in your session
   c) A list of installed packages
   d) The contents of your working directory folder

2. **True or false.** Deleting a line of code from your script automatically removes the corresponding object from your session.

3. **Predict the output.**
   ```r
   a <- 1
   b <- 2
   rm(a)
   ls()
   ```

4. **Write the code.** Write the command that removes every object currently in your session at once.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The names of all objects currently in your session.
2. **False.** Removing a line from your script has no effect on objects already created in memory; you must explicitly run `rm()` or restart R to clear them.
3. `[1] "b"` — only `b` remains after `a` was removed.
4. `rm(list = ls())`

</details>
:::
