# Logical Data {#sec-logical-data}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor asks: "Which participants are current smokers, and which are 65 or older?" You need to flag each participant with a clear yes/no answer for both questions, so these flags can later be used to filter and count people — but you're not sure what kind of value a "yes/no" answer should even be stored as in R.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand R's logical data type (`TRUE`/`FALSE`) and the comparison operators used to create it.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- What are all the possible answers to a genuine yes/no question — is there really a third option?
- If you wanted R to check "is this participant's age 65 or more," what symbol would represent "is greater than or equal to"?
:::

## The R Concept

**Logical** data has exactly two possible values: `TRUE` or `FALSE` (always written in capital letters, never quoted — `"TRUE"` is character text, not logical). Logical values are most often produced by **comparison operators**:

| Operator | Meaning |
|---|---|
| `==` | equal to |
| `!=` | not equal to |
| `>` | greater than |
| `<` | less than |
| `>=` | greater than or equal to |
| `<=` | less than or equal to |

You can also combine logical values using:

| Operator | Meaning |
|---|---|
| `&` | and (both must be TRUE) |
| `\|` | or (at least one must be TRUE) |
| `!` | not (flips TRUE to FALSE, and vice versa) |

The single most common beginner mistake is confusing `=` (assignment) with `==` (equality comparison) — they look almost identical but do fundamentally different things.

## Demonstration

```{r}
#| eval: false
age <- 68
is_senior <- age >= 65
is_senior
```

```
[1] TRUE
```

```{r}
#| eval: false
smoker_status <- "Yes"
is_smoker <- smoker_status == "Yes"
is_smoker
```

```
[1] TRUE
```

```{r}
#| eval: false
# Combining two conditions with &
high_risk <- is_senior & is_smoker
high_risk
```

```
[1] TRUE
```

```{r}
#| eval: false
# Flipping a logical value with !
is_nonsmoker <- !is_smoker
is_nonsmoker
```

```
[1] FALSE
```

## Explore

Change `age` to `50` and rerun `is_senior <- age >= 65`. What do you get, and does `high_risk` change automatically? (Recall from the objects page: it won't, unless you rerun the calculation.) Try `age == "68"` (comparing a number to quoted text) — does it still work as expected?

## Solve the Problem

```{r}
#| eval: false
age <- 71
smoker_status <- "Yes"

is_senior <- age >= 65
is_smoker <- smoker_status == "Yes"

is_senior
```

```
[1] TRUE
```

```{r}
#| eval: false
is_smoker
```

```
[1] TRUE
```

Both flags now clearly and unambiguously answer your supervisor's two questions for this participant, ready to be reused later (for example, to count or filter many participants at once — a technique covered in Level 5).

## Interpret the Result

`is_senior` being `TRUE` means this participant meets the "65 or older" criterion; `is_smoker` being `TRUE` means they reported current smoking. Stored as logical values rather than text, these flags can be directly combined (as with `high_risk` above) or later used to filter a whole dataset down to exactly the participants who match both conditions.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing `=` and `==`.** `age = 65` *assigns* `65` to `age`; `age == 65` *checks whether* `age` equals `65`. Using the wrong one produces very different, sometimes silently wrong, results.
- **Quoting TRUE/FALSE unnecessarily.** `is_smoker <- "TRUE"` creates character text, not a logical value — drop the quotes: `is_smoker <- TRUE`.
- **Misusing `&` versus `|`.** `age >= 65 & age <= 75` requires *both* conditions; using `|` instead would match ages 65 and under *or* 75 and over — the opposite of what's usually intended for an age range.
- **Forgetting operator precedence.** Comparisons are evaluated before `&`/`|`, but when combining several conditions, parentheses make your intent explicit and prevent subtle bugs.
:::

## Practice Problem

A participant has `systolic_bp <- 145`. Create a logical object `has_elevated_bp` that is `TRUE` if `systolic_bp` is greater than or equal to `130`.

## Challenge Problem

A participant has `age <- 58` and `bmi <- 31.2`. Create a logical object `needs_followup` that is `TRUE` only if the participant is *both* under 65 *and* has a BMI of 30 or above (using `&`), then create a second object `either_condition` using `|` instead, and explain in one sentence why the two results differ.

## Key Takeaways

- Logical values are `TRUE` or `FALSE` (unquoted); comparison operators (`==`, `!=`, `>`, `<`, `>=`, `<=`) produce them.
- `=` assigns a value; `==` checks equality — mixing these up is one of the most common beginner errors.
- `&` (and), `|` (or), and `!` (not) combine or invert logical values, and parentheses help make combined conditions unambiguous.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `age == 65` do?
   a) Assigns 65 to age
   b) Checks whether age equals 65, returning TRUE or FALSE
   c) Rounds age to 65
   d) Creates a character string

2. **True or false.** `"TRUE"` and `TRUE` are the same value in R.

3. **Predict the output.**
   ```r
   x <- 10
   y <- 20
   (x < y) & (y < 30)
   ```

4. **Find the error.**
   ```r
   is_adult <- age = 18
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Checks whether `age` equals `65`, returning `TRUE` or `FALSE`.
2. **False.** `"TRUE"` (quoted) is character text; `TRUE` (unquoted) is logical — they have different classes.
3. `[1] TRUE` — both `10 < 20` and `20 < 30` are true, and `&` requires both.
4. The code uses `=` where `==` was intended for comparison (and the logic is also reversed). It likely should read `is_adult <- age >= 18`.

</details>
:::
