# Getting Help in R {#sec-getting-help}

::: {.pbl-box .pbl-problem}
**The Problem**

You just learned that `mean()` has an argument called `na.rm`. But you also notice `round()` takes a second argument you've been supplying without really understanding — `round(bmi, 1)` — and you're not sure what that `1` actually controls, or what other arguments `round()` might accept that you don't know about yet.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Learn to use R's built-in help system to answer your own questions about any function, without needing to search the internet or ask a colleague first.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If a function's documentation were built directly into R, what symbol might you type before a function's name to pull it up?
- What information would you expect a function's help page to contain — just examples, or something more structured?
:::

## The R Concept

Every function that ships with R (and every well-made package function) has a **help page** describing exactly what it does. Three ways to access it:

```{r}
#| eval: false
?round        # opens the help page for round() in the Help pane
help(round)   # identical to ?round
args(round)   # shows just the function's arguments, without the full page
```

A typical help page has a consistent structure worth learning to skim:

- **Description** — a short summary of what the function does.
- **Usage** — the function's exact syntax, including argument names and their default values.
- **Arguments** — an explanation of each argument.
- **Value** — what the function returns.
- **Examples** — runnable code demonstrating typical use, at the very bottom (often the fastest place to start reading).

For functions with special characters in their name (like `if` or `+`), you need to wrap the name in backticks or quotes: `` ?`+` `` or `help("if")`.

## Demonstration

```{r}
#| eval: false
?round
```

Opens the help page in the Help pane, showing (abbreviated):

```
round(x, digits = 0)

Arguments:
x        a numeric vector.
digits   integer indicating the number of decimal places to round to.
```

```{r}
#| eval: false
args(round)
```

```
function (x, digits = 0)
NULL
```

This confirms: `round()`'s second argument is called `digits`, and it defaults to `0` if you don't supply it — which is exactly why `round(bmi, 1)` rounds to 1 decimal place, since `1` is being matched to the `digits` argument by its position.

## Explore

Run `?mean` and scroll to the Arguments section. Confirm you can now see, in R's own documentation, the `na.rm` argument you used on the previous page, along with its default value. Then try `example(round)` — a function that runs the worked examples straight from the help page.

## Solve the Problem

```{r}
#| eval: false
?round
```

Reading the help page's Usage section — `round(x, digits = 0)` — answers your question directly: the second argument to `round()` is named `digits`, and it controls how many decimal places the result is rounded to. `round(bmi, 1)` rounds `bmi` to 1 decimal place; had you written `round(bmi, 0)` (or just `round(bmi)`, relying on the default), it would round to the nearest whole number instead.

## Interpret the Result

Knowing that `round()`'s second argument is specifically called `digits` — not just "some number you put there" — means you can now write clearer code by naming it explicitly: `round(bmi, digits = 1)`, which is more self-explanatory to a future reader than the shorter, positional `round(bmi, 1)`.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Jumping to a search engine before checking `?function`.** Built-in help is often faster and always version-matched to the R you're actually running — search results may reference an outdated or different version of a function.
- **Being intimidated by a help page's dense formatting.** Skip straight to the Examples section at the bottom first — it's usually the fastest way to understand a function's practical use before reading the more formal Arguments section.
- **Forgetting quotes or backticks for special names.** `?if` alone will error; you need `` ?`if` `` or `help("if")`.
- **Not realizing argument defaults are shown in Usage.** `round(x, digits = 0)` tells you `digits` defaults to `0` if you don't specify it — this is often the fastest way to discover a function's default behavior.
:::

## Practice Problem

Use `?sum` to answer: does `sum()` have an `na.rm` argument, similar to `mean()`? Confirm your answer by reading the Arguments section.

## Challenge Problem

You encounter an unfamiliar function, `quantile()`, in a colleague's script. Use `?quantile` to identify what its default argument (besides `x`) is called, and what value it defaults to when not specified.

## Key Takeaways

- `?function_name` (or `help(function_name)`) opens R's built-in documentation in the Help pane.
- A help page's Usage section shows argument names and their defaults; the Examples section (often the fastest starting point) shows runnable code.
- Checking built-in help first is usually faster and more reliable than searching the web, especially for base R functions.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Which command opens the help page for `sd()`?
   a) `sd?`
   b) `?sd`
   c) `help sd`
   d) `sd.help()`

2. **True or false.** A function's help page always lists its default argument values, if it has any.

3. **Predict the output.** What section of a help page would you check to find out what a function *returns*?

4. **Write the code.** Write the command to view the help page for the function `paste`, using the `help()` form rather than `?`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `?sd`
2. **True.** The Usage section of a help page shows every argument along with any default value it has.
3. The **Value** section describes what the function returns.
4. `help(paste)`

</details>
:::
