# Wide Format vs. Long Format {#sec-wide-vs-long}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor gives you `data/bp_visits_wide.csv`: 28 participants, one row each, with separate columns `sbp_visit1`, `sbp_visit2`, `sbp_visit3` for blood pressure at three clinic visits. You're asked to plot how blood pressure changes over time, using `visit` as one axis. But there's no single `visit` column to plot against — the visit information is baked into the *column names themselves*, not into any column's values.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Recognize the difference between wide format and long format data, and understand why some tasks require one shape or the other.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- In `sbp_visit1`, `sbp_visit2`, `sbp_visit3`, is "which visit" stored as a *value* somewhere, or as part of a *column name*?
- If you wanted to `group_by(visit)` (Level 5) and compute a mean blood pressure per visit, could you do that if "visit" isn't an actual column containing values like `1`, `2`, `3`?
:::

## The R Concept

**Wide format** stores one row per observational unit (here, one row per participant), with separate columns for each repeated measurement:

```
id  sbp_visit1  sbp_visit2  sbp_visit3
1   125         120         117
2   123         121         129
```

**Long format** stores one row per *combination* of observational unit and measurement occasion, with the measurement occasion itself living in a column:

```
id  visit   sbp
1   visit1  125
1   visit2  120
1   visit3  117
2   visit1  123
2   visit2  121
2   visit3  129
```

Both represent the *exact same information* — no data is gained or lost — but long format has an actual `visit` column you can `filter()`, `group_by()`, or plot against, while wide format encodes that same information into column names that dplyr functions can't directly reference as values.

## Demonstration

```{r}
#| eval: false
bp_wide <- read.csv("data/bp_visits_wide.csv")

names(bp_wide)
```

```
[1] "id"           "county"       "clinic_site"  "sbp_visit1"
[5] "sbp_visit2"   "sbp_visit3"   "weight_visit1" "weight_visit2"
[9] "weight_visit3"
```

```{r}
#| eval: false
nrow(bp_wide)
```

```
[1] 28
```

28 rows — one per participant — with visit information spread across six separate measurement columns (three for `sbp`, three for `weight`).

## Explore

Try `mean(bp_wide$sbp_visit1, na.rm = TRUE)`, then `mean(bp_wide$sbp_visit2, na.rm = TRUE)`, then `mean(bp_wide$sbp_visit3, na.rm = TRUE)` — three separate lines, one per visit, because there's no single `sbp` column to `group_by(visit)` on. This repetition is exactly the signal that a reshape is needed.

## Solve the Problem

```{r}
#| eval: false
bp_wide <- read.csv("data/bp_visits_wide.csv")

# Wide format: three repeated columns, one row per participant
head(bp_wide[, c("id", "sbp_visit1", "sbp_visit2", "sbp_visit3")], 3)
```

```
  id sbp_visit1 sbp_visit2 sbp_visit3
1  1        125        120        117
2  2        123        121        129
3  3        130        128        124
```

There is no way to `group_by()` a "visit" here, because "visit" isn't a column of values — it's spread across three column *names*. The next page's `pivot_longer()` is exactly the tool that fixes this.

## Interpret the Result

`bp_wide` is a perfectly valid, common way to store repeated-measures data — and it's often the *easier* format for a human to read at a glance, comparing one participant's three visits side by side. But it's the wrong shape for `group_by(visit)`-style analysis or for plotting change over time, which is why reshaping tools exist at all: not to "fix" wide format, but to convert between the shape that's easy to *read* and the shape that's easy to *analyze*.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Assuming one format is objectively "correct."** Neither wide nor long is inherently better — wide is often better for reading and data entry; long is often better for `group_by()`-style analysis and plotting. The right shape depends on the task.
- **Trying to `group_by()` a value that's actually a column name.** `group_by(bp_wide, visit)` fails outright — there is no `visit` column in wide format; the visit information lives in the column *names* `sbp_visit1`/`sbp_visit2`/`sbp_visit3`, not in any column's values.
- **Forgetting reshaping doesn't change the underlying data.** Converting wide to long (or back) doesn't add, remove, or recalculate anything — it only changes which pieces of information are stored as column names versus column values.
:::

## Practice Problem

Look at `bp_wide`'s `weight_visit1`, `weight_visit2`, `weight_visit3` columns. Without writing any reshaping code yet, describe in words what the long-format equivalent would look like: what would the column names be, and how many rows would it have?

## Challenge Problem

If `bp_wide` has 28 rows and 3 visits' worth of `sbp` columns, how many rows would a long-format version (one row per participant-visit combination, `sbp` only) have? What if you also included the `weight` measurements in the same long format?

## Key Takeaways

- Wide format has one row per unit, with repeated measurements spread across multiple columns.
- Long format has one row per unit-and-occasion combination, with the occasion itself stored as an actual column of values.
- Both formats hold identical information — the right choice depends on whether you're reading data by eye or feeding it into `group_by()`-style analysis or plotting.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In wide format, how is repeated measurement information (like "visit number") typically stored?
   a) In an actual column of values
   b) Encoded in the column names
   c) It cannot be stored in wide format at all
   d) As row names

2. **True or false.** Converting data from wide to long format changes the underlying values.

3. **Predict the output.** If `bp_wide` has 28 rows and 3 visit columns for `sbp`, how many rows would the fully reshaped long format (`sbp` only) have?

4. **Short answer.** Why can't you `group_by(bp_wide, visit)` directly on the wide-format data?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Encoded in the column names (e.g., `sbp_visit1`, `sbp_visit2`).
2. **False.** Reshaping only changes the *shape* of the data (which information is a column name vs. a column value) — no values are added, removed, or recalculated.
3. 84 rows (28 participants × 3 visits).
4. There is no `visit` column in the wide-format data — the visit information is embedded in the column names (`sbp_visit1`, etc.), not stored as values `group_by()` can reference.

</details>
:::
