# `pivot_wider()` {#sec-pivot-wider}

::: {.pbl-box .pbl-problem}
**The Problem**

You've reshaped to long format and computed mean blood pressure per county per visit with `group_by(county, visit)` + `summarise()` — but the result is 15 rows (5 counties × 3 visits) stacked vertically, which is awkward for a report table. Your professor wants a compact table with one row per county and one column per visit, so all three visits can be compared side by side at a glance.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Convert long-format data back into wide format using `tidyr::pivot_wider()` — the reverse of `pivot_longer()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If `pivot_longer()` takes several columns and stacks them into two, what would you expect the *reverse* operation to do with those same two columns?
- Looking at a long-format table with a `visit` column and a `mean_sbp` column, which one should become new column *names*, and which one should fill in the new column *values*?
:::

## The R Concept

**`pivot_wider()`** takes one column of names and one column of values, and spreads them back out into separate columns:

```{r}
#| eval: false
pivot_wider(df,
  names_from = name_column,
  values_from = value_column
)
```

```{r}
#| eval: false
pivot_wider(county_visit_means,
  names_from = visit,
  values_from = mean_sbp
)
```

This is exactly the reverse of `pivot_longer()`'s `names_to`/`values_to`: `names_from` says which column's *values* should become new column *names*, and `values_from` says which column's values should fill in those new columns.

## Demonstration

```{r}
#| eval: false
library(dplyr)
library(tidyr)

bp_wide <- read.csv("data/bp_visits_wide.csv")

bp_long_sbp <- pivot_longer(bp_wide,
  cols = starts_with("sbp_visit"),
  names_to = "visit",
  names_prefix = "sbp_",
  values_to = "sbp"
)

county_visit_means <- bp_long_sbp |>
  group_by(county, visit) |>
  summarise(mean_sbp = round(mean(sbp, na.rm = TRUE), 2), .groups = "drop")

county_visit_means
```

```
# A tibble: 15 x 3
   county      visit  mean_sbp
   <chr>       <chr>     <dbl>
 1 Cedar Falls visit1   136.00
 2 Cedar Falls visit2   130.25
 3 Cedar Falls visit3   128.00
 4 Millbrook   visit1   122.17
 5 Millbrook   visit2   120.33
 6 Millbrook   visit3   122.00
 7 Oakford     visit1   120.67
 8 Oakford     visit2   117.17
 9 Oakford     visit3   119.50
10 Riverdale   visit1   123.33
11 Riverdale   visit2   119.00
12 Riverdale   visit3   120.67
13 Springvale  visit1   126.33
14 Springvale  visit2   121.17
15 Springvale  visit3   116.17
```

15 rows, exactly as `group_by(county, visit)` would produce (5 counties × 3 visits) — readable, but each county's three visits are spread across three separate rows.

```{r}
#| eval: false
report_table <- pivot_wider(county_visit_means,
  names_from = visit,
  values_from = mean_sbp
)

report_table
```

```
# A tibble: 5 x 4
  county      visit1 visit2 visit3
  <chr>        <dbl>  <dbl>  <dbl>
1 Cedar Falls 136.00 130.25 128.00
2 Millbrook   122.17 120.33 122.00
3 Oakford     120.67 117.17 119.50
4 Riverdale   123.33 119.00 120.67
5 Springvale  126.33 121.17 116.17
```

Now 5 rows (one per county) and 3 visit columns — exactly the compact report table your professor asked for.

## Explore

Try `pivot_wider()` on `bp_long_sbp` directly (not the summarised version), using `names_from = visit, values_from = sbp`. Confirm this reconstructs something very close to the original `bp_wide` data frame — `pivot_wider()` is the true inverse of `pivot_longer()`, so reshaping long-to-wide and back again returns you to (almost) where you started.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)
library(tidyr)

bp_wide <- read.csv("data/bp_visits_wide.csv")

bp_long_sbp <- pivot_longer(bp_wide,
  cols = starts_with("sbp_visit"),
  names_to = "visit",
  names_prefix = "sbp_",
  values_to = "sbp"
)

county_visit_report <- bp_long_sbp |>
  group_by(county, visit) |>
  summarise(mean_sbp = round(mean(sbp, na.rm = TRUE), 2), .groups = "drop") |>
  pivot_wider(names_from = visit, values_from = mean_sbp)

county_visit_report
```

```
# A tibble: 5 x 4
  county      visit1 visit2 visit3
  <chr>        <dbl>  <dbl>  <dbl>
1 Cedar Falls 136.00 130.25 128.00
2 Millbrook   122.17 120.33 122.00
3 Oakford     120.67 117.17 119.50
4 Riverdale   123.33 119.00 120.67
5 Springvale  126.33 121.17 116.17
```

Cedar Falls starts with the highest mean blood pressure (136.00) and shows the clearest downward trend across visits (136.00 → 130.25 → 128.00), while Springvale shows the largest overall drop from visit 1 to visit 3 (126.33 → 116.17).

## Interpret the Result

`county_visit_report` compresses 15 long-format rows into 5 wide-format rows, one per county, with visits laid out side by side — exactly the shape a reader scanning a report wants, even though `group_by()` + `summarise()` (Level 5) only knows how to produce the longer, stacked version. Note that Cedar Falls has only 4 participants in this subset (versus 6 in each other county), so its more dramatic-looking trend rests on a notably smaller sample.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `pivot_wider()` needs unique name-value combinations.** If `county_visit_means` had two rows with the same `county` and `visit` (a duplicate), `pivot_wider()` would need to combine them into a list-column rather than a single value — a common source of confusing "list of lists" output for beginners.
- **Confusing `names_from` and `values_from`.** `names_from` picks which column's *values* become the new column headers; `values_from` picks which column's values fill in underneath them. Swapping them produces an error or nonsensical column names.
- **Forgetting small group sizes remain small after reshaping.** Cedar Falls' n=4 doesn't become more reliable just because its numbers are now displayed in a wide table instead of a long one — reshaping changes layout, not sample size.
:::

## Practice Problem

Use `group_by(county, visit)` and `summarise()` to compute the mean `weight` per county per visit (after first pivoting the weight columns to long format), then use `pivot_wider()` to build a county-by-visit weight report table.

## Challenge Problem

Starting from `bp_long_sbp`, use `pivot_wider(names_from = visit, values_from = sbp)` directly (no `group_by()`/`summarise()` first) to reconstruct a wide table. Compare its columns to the original `bp_wide`'s `sbp_visit1`/`sbp_visit2`/`sbp_visit3` columns — are the values identical?

## Key Takeaways

- `pivot_wider(df, names_from = ..., values_from = ...)` is the reverse of `pivot_longer()`, spreading long rows back into wide columns.
- `names_from` selects which column's values become new column names; `values_from` selects which column fills them in.
- `pivot_wider()` requires unique name-value combinations per row group — duplicates produce list-columns instead of single values.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `names_from` specify in `pivot_wider()`?
   a) Which column's values become new column names
   b) Which column holds the final row count
   c) Which column to group by
   d) Which column's values fill in the new columns

2. **True or false.** `pivot_wider()` is the reverse operation of `pivot_longer()`.

3. **Predict the output.** If a long-format table has 5 counties and 3 visits (15 rows total), how many rows would `pivot_wider(names_from = visit, values_from = mean_sbp)` produce?

4. **Find the error.**
   ```r
   pivot_wider(county_visit_means, names_from = mean_sbp, values_from = visit)
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **a)** Which column's values become new column names.
2. **True.** `pivot_wider()` reverses what `pivot_longer()` does, spreading long rows back into wide columns.
3. 5 rows — one per county, since each county's 3 visit-rows collapse into 1 wide row with 3 visit columns.
4. `names_from` and `values_from` are swapped — this would try to use the numeric `mean_sbp` values as column *names* and the text `visit` values as the data filling them in, which is backwards. It should be `names_from = visit, values_from = mean_sbp`.

</details>
:::
