# `unite()` {#sec-unite}

::: {.pbl-box .pbl-problem}
**The Problem**

After `pivot_longer()`, your data has separate `id` and `visit` columns — 84 rows, but no single column that uniquely labels each participant-visit combination as one identifier. Your professor wants a single `record_id` column (like `"1_visit1"`) to use as a unique row label when exporting the long-format data for a database import.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Combine two or more columns into one, using `tidyr::unite()` — the reverse of `separate()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From the previous page, `separate()` split one compound column into several. What would you expect the reverse operation — combining several columns into one — to need as arguments?
- If you combined `id` (`1`) and `visit` (`"visit1"`) with an underscore, what single value would you expect the result to be?
:::

## The R Concept

**`unite()`** combines multiple columns into one, joined by a separator:

```{r}
#| eval: false
unite(df, new_col, col1, col2, sep = "separator")
```

```{r}
#| eval: false
unite(bp_long_sbp, record_id, id, visit, sep = "_")
```

This is exactly the reverse of `separate()`: instead of naming a `sep` to split *by*, you name a `sep` to join *with*. The new column name (`record_id`) comes first, followed by the columns to combine, in the order you want them joined.

## Demonstration

```{r}
#| eval: false
library(tidyr)

bp_long_sbp <- pivot_longer(bp_wide,
  cols = starts_with("sbp_visit"),
  names_to = "visit",
  names_prefix = "sbp_",
  values_to = "sbp"
)

bp_united <- unite(bp_long_sbp, record_id, id, visit, sep = "_")

head(bp_united[, c("record_id", "county", "sbp")], 4)
```

```
  record_id   county sbp
1    1_visit1  Oakford 125
2    1_visit2  Oakford 120
3    1_visit3  Oakford 117
4    2_visit1 Riverdale 123
```

`id` and `visit` are gone from the result; `record_id` has taken their place, combining both pieces of information into one text value.

## Explore

Try `length(unique(bp_united$record_id))` and compare it to `nrow(bp_united)`. Since every participant-visit combination is unique, these two numbers should match exactly — confirming `record_id` works as a genuine unique row identifier.

## Solve the Problem

```{r}
#| eval: false
library(tidyr)

bp_long_sbp <- pivot_longer(bp_wide,
  cols = starts_with("sbp_visit"),
  names_to = "visit",
  names_prefix = "sbp_",
  values_to = "sbp"
)

bp_united <- unite(bp_long_sbp, record_id, id, visit, sep = "_")

length(unique(bp_united$record_id))
```

```
[1] 84
```

```{r}
#| eval: false
nrow(bp_united)
```

```
[1] 84
```

84 unique `record_id` values for 84 rows — confirming every participant-visit combination produced a genuinely unique identifier, exactly what your professor needs for the database export.

## Interpret the Result

`record_id` now serves as a single, human-readable unique key (`"1_visit1"`, `"1_visit2"`, ...) for each row of the long-format data — a common requirement before exporting to external systems that expect one unique identifier column rather than a combination of several.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `unite()` converts everything to text.** Even though `id` started as a number, `record_id` is a character column after uniting — you can no longer do arithmetic on the `id` portion without separating it back out first.
- **Choosing a `sep` that could reappear inside the combined values.** If `visit` values themselves contained an underscore, later trying to `separate()` this `record_id` back apart with `sep = "_"` could split it into more pieces than intended.
- **Forgetting `unite()` removes the original columns by default.** Like `separate()`, `unite()` drops `id` and `visit` from the result unless you specify `remove = FALSE` to keep them alongside the new combined column.
:::

## Practice Problem

Use `unite()` to combine `county` and `visit` into a single column called `county_visit`, separated by a dash. How many unique `county_visit` values result?

## Challenge Problem

Use `unite()` to create `record_id` (as shown above) with `remove = FALSE`, keeping the original `id` and `visit` columns alongside the new combined one. Then use `separate()` to split `record_id` back apart, and confirm with `identical()` that you recover the original `id` and `visit` values (keeping in mind `unite()` converts `id` to text, so you may need `as.character()` on the original for a fair comparison).

## Key Takeaways

- `unite(df, new_col, col1, col2, sep = "...")` combines multiple columns into one text column, joined by a separator — the reverse of `separate()`.
- `unite()` always produces a character (text) column, even if the original columns were numeric.
- Like `separate()`, `unite()` removes the original columns by default unless `remove = FALSE` is specified.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `unite(df, id_visit, id, visit, sep = "_")` produce?
   a) Two new columns, `id` and `visit`, split apart
   b) One new column, `id_visit`, combining `id` and `visit` with an underscore
   c) A count of unique `id` values
   d) A filtered data frame with only matching rows

2. **True or false.** `unite()` preserves the original numeric type of a column if it was numeric before combining.

3. **Predict the output.** What would `unite(df, key, county, visit, sep = "-")` produce for a row where `county = "Oakford"` and `visit = "visit1"`?

4. **Find the error.**
   ```r
   unite(bp_long_sbp, id, visit, record_id, sep = "_")
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** One new column, `id_visit`, combining `id` and `visit` with an underscore.
2. **False.** `unite()` always converts the combined result to a character (text) column, even if the inputs were numeric.
3. `"Oakford-visit1"`.
4. The new column name must come first, followed by the columns to combine — here `id` and `visit` are existing columns, but `record_id` (the intended new name) is listed last as if it were a third column to combine, which isn't what was likely intended. It should be `unite(bp_long_sbp, record_id, id, visit, sep = "_")`.

</details>
:::
