# Pivoting Multiple Measurements at Once {#sec-multiple-value-columns}

::: {.pbl-box .pbl-problem}
**The Problem**

Back on the `pivot_longer()` page, a Common Mistake warned that pivoting only `sbp_visit1`/`sbp_visit2`/`sbp_visit3` leaves `weight_visit1`/`weight_visit2`/`weight_visit3` untouched — duplicating each participant's weight values across all three of their new long rows. Your professor actually wants **both** measurements reshaped together, correctly: one row per participant-visit, with separate `sbp` and `weight` columns side by side.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Reshape multiple measurement types to long format simultaneously, using `pivot_longer()`'s special `.value` token.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Look at the column name `sbp_visit1`. It actually encodes *two* pieces of information at once — which measurement, and which visit. Can you identify each piece?
- If `pivot_longer()` could split a column name into pieces the same way `separate()` splits a column's *values*, what would you want the "measurement" piece to become, versus the "visit" piece?
:::

## The R Concept

Column names like `sbp_visit1` and `weight_visit1` encode **two** pieces of information joined by an underscore: which measurement (`sbp` or `weight`), and which visit (`visit1`). The special `.value` token in `names_to`, combined with `names_sep`, tells `pivot_longer()` to split each column name into pieces and use *one specific piece* (wherever `.value` is placed) as the name of a whole new **column**, rather than folding it into a single generic value column:

```{r}
#| eval: false
pivot_longer(df,
  cols = matches("visit"),
  names_to = c(".value", "visit"),
  names_sep = "_"
)
```

```{r}
#| eval: false
pivot_longer(bp_wide,
  cols = c(starts_with("sbp_visit"), starts_with("weight_visit")),
  names_to = c(".value", "visit"),
  names_sep = "_"
)
```

Here, `sbp_visit1` splits at the underscore into `"sbp"` and `"visit1"`. Because `.value` occupies the *first* position in `names_to`, that first piece (`"sbp"`) becomes an actual **column name** in the result, while the second piece (`"visit1"`) becomes a value in a `visit` column — giving you `sbp` and `weight` as two separate, properly reshaped columns, instead of duplicating one across the other.

## Demonstration

```{r}
#| eval: false
library(tidyr)

bp_wide <- read.csv("data/bp_visits_wide.csv")

bp_long_both <- pivot_longer(bp_wide,
  cols = c(starts_with("sbp_visit"), starts_with("weight_visit")),
  names_to = c(".value", "visit"),
  names_sep = "_"
)

names(bp_long_both)
```

```
[1] "id"          "county"      "clinic_site" "visit"       "sbp"
[6] "weight"
```

Both `sbp` and `weight` now exist as genuine, separate columns — not one duplicated across the other's long rows.

```{r}
#| eval: false
head(bp_long_both[, c("id", "visit", "sbp", "weight")], 6)
```

```
  id  visit sbp weight
1  1 visit1 125  101.4
2  1 visit2 120  100.8
3  1 visit3 117  101.0
4  2 visit1 123   80.1
5  2 visit2 121   79.6
6  2 visit3 129   79.0
```

## Explore

Try `nrow(bp_long_both)`. It should still be 84 (28 participants × 3 visits) — reshaping two measurements together doesn't create extra rows, since `.value` correctly keeps `sbp` and `weight` as separate columns within the same row, rather than stacking them into additional rows.

## Solve the Problem

```{r}
#| eval: false
library(tidyr)

bp_wide <- read.csv("data/bp_visits_wide.csv")

bp_long_both <- pivot_longer(bp_wide,
  cols = c(starts_with("sbp_visit"), starts_with("weight_visit")),
  names_to = c(".value", "visit"),
  names_sep = "_"
)

nrow(bp_long_both)
```

```
[1] 84
```

```{r}
#| eval: false
sum(!is.na(bp_long_both$sbp))
```

```
[1] 82
```

```{r}
#| eval: false
sum(!is.na(bp_long_both$weight))
```

```
[1] 82
```

84 rows (not 168), with both `sbp` and `weight` correctly present as their own columns in every row — exactly the properly reshaped structure your professor asked for, with no duplication of one measurement across the other.

## Interpret the Result

`bp_long_both` is now genuinely "tidy": one row per participant-visit combination, with each measurement type in its own column. This is the format that would let you `group_by(visit)` and compute means of *both* `sbp` and `weight` in a single `summarise(across(...))` call (Level 5), something the earlier single-measurement reshape couldn't offer without either duplicating data or reshaping twice.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting to include `.value` in `names_to`.** Without it, `pivot_longer()` has no way to know that `"sbp"` and `"weight"` should become separate columns rather than both being lumped into one undifferentiated value column.
- **Choosing the wrong `names_sep`.** `sbp_visit1` splits correctly on the *first* underscore into `"sbp"` and `"visit1"` — but if a measurement name itself contained an underscore, the split would produce more pieces than the two `names_to` expects.
- **Reshaping each measurement separately when they didn't need to be.** Doing two separate `pivot_longer()` calls (one for `sbp`, one for `weight`) and then trying to `merge` them back together is far more error-prone than using `.value` to reshape both in a single step.
:::

## Practice Problem

Confirm that `bp_long_both`'s `sbp` column, restricted to `visit == "visit1"`, matches `bp_wide$sbp_visit1` exactly (hint: use `filter()` from Level 5, then compare with `identical()` after matching row order by `id`).

## Challenge Problem

Use `bp_long_both`, `group_by(visit)`, and `summarise(across(c(sbp, weight), ~ mean(.x, na.rm = TRUE)))` (Level 5's `across()`) to compute mean blood pressure and mean weight for each visit in a single call.

## Key Takeaways

- The `.value` token in `pivot_longer()`'s `names_to` lets you reshape multiple measurement types into separate columns at once, rather than lumping them all into a single generic value column.
- `names_sep` (or `names_pattern`) tells `pivot_longer()` how to split each compound column name into its `.value` and other pieces.
- Reshaping multiple measurements together in one call avoids both row-count duplication and the awkwardness of merging separately reshaped pieces back together.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does the `.value` token do inside `pivot_longer()`'s `names_to`?
   a) Deletes that piece of the column name
   b) Uses that piece of the split column name as an actual new column name
   c) Converts that piece to numeric
   d) Has no special meaning; it's just a placeholder string

2. **True or false.** Reshaping `sbp` and `weight` together with `.value` produces more rows than reshaping `sbp` alone.

3. **Predict the output.** After the `.value`-based reshape, how many columns would `bp_long_both` have among `id`, `county`, `clinic_site`, `visit`, `sbp`, `weight`?

4. **Find the error.**
   ```r
   pivot_longer(bp_wide,
     cols = c(starts_with("sbp_visit"), starts_with("weight_visit")),
     names_to = "visit",
     names_sep = "_"
   )
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Uses that piece of the split column name as an actual new column name.
2. **False.** Both reshapes produce 84 rows — `.value` keeps `sbp` and `weight` as separate columns within the same rows, rather than adding extra rows.
3. 6 columns: `id`, `county`, `clinic_site`, `visit`, `sbp`, `weight`.
4. `names_to` is missing the `.value` token — with just `"visit"`, `pivot_longer()` expects only one piece per column name and can't know that `"sbp"` and `"weight"` should become separate columns. It should be `names_to = c(".value", "visit")`.

</details>
:::
