# Keys and Relationships {#sec-keys-and-relationships}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor now hands you two separate files: `data/baseline_enrollment.csv` (20 participants enrolled at the start of the study) and `data/followup_visit.csv` (20 follow-up visit records, collected a year later by a different field team). Some participants appear in both files; some appear in only one. Before writing any code, you need to understand exactly how these two files relate to each other — otherwise even a correct-looking merge could silently drop or duplicate participants.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand what a "key" is, and recognize the different possible relationships between two datasets before attempting to combine them.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Both files have an `id` column. What do you think that column's job is, compared to every other column in each file?
- If a participant enrolled at baseline but was never seen again at follow-up, what would you expect to happen to that participant's row in `followup_visit.csv`?
:::

## The R Concept

A **key** is the column (or combination of columns) that identifies which row in one dataset corresponds to which row in another. Here, `id` is the key linking `baseline_enrollment.csv` and `followup_visit.csv` — every row in each file represents one participant, identified by that same `id` value.

Two datasets sharing a key can relate to each other in several ways:

- **Every id in both files matches exactly** (a clean one-to-one relationship) — rare in real longitudinal data.
- **Some ids exist in one file but not the other** — participants who dropped out before follow-up, or new recruits added at follow-up.
- **One id in one file matches multiple rows in the other** (a one-to-many relationship) — for example, if `followup_visit.csv` recorded two visits for some participants.

Before merging anything, always inspect both key columns first.

## Demonstration

```{r}
#| eval: false
baseline <- read.csv("data/baseline_enrollment.csv")
followup <- read.csv("data/followup_visit.csv")

nrow(baseline)
```

```
[1] 20
```

```{r}
#| eval: false
nrow(followup)
```

```
[1] 20
```

Both files have 20 rows — but that doesn't mean they contain the same 20 participants.

```{r}
#| eval: false
sum(baseline$id %in% followup$id)
```

```
[1] 15
```

```{r}
#| eval: false
sum(!(baseline$id %in% followup$id))
```

```
[1] 5
```

## Explore

Try `sum(followup$id %in% baseline$id)` and `sum(!(followup$id %in% baseline$id))`. Compare these counts to the ones above for `baseline`. Notice the numbers aren't symmetric in an obvious way until you think it through: some `followup` ids are genuinely new participants who weren't in `baseline` at all.

## Solve the Problem

```{r}
#| eval: false
baseline <- read.csv("data/baseline_enrollment.csv")
followup <- read.csv("data/followup_visit.csv")

# ids in baseline but NOT in followup - likely dropouts
setdiff(baseline$id, followup$id)
```

```
[1] 16 17 18 19 20
```

```{r}
#| eval: false
# ids in followup but NOT in baseline - likely new recruits
setdiff(followup$id, baseline$id)
```

```
[1] 21 22 23 24 25
```

```{r}
#| eval: false
# ids present in both files
length(intersect(baseline$id, followup$id))
```

```
[1] 15
```

Exactly as your professor's description hinted: 5 baseline participants (ids 16-20) have no follow-up record at all — likely dropouts — while 5 follow-up ids (21-25) don't appear in baseline at all — likely participants recruited after the study began. 15 ids (1-15) appear in both files.

## Interpret the Result

Before writing a single `join()` call, you now know exactly what to expect: a merge that keeps only matching ids will produce 15 rows; a merge that keeps every baseline id will produce 20 (with 5 missing follow-up values); a merge that keeps every id from either file will produce 25. Knowing this *ahead of time* is exactly what lets you catch a bug if your eventual join produces an unexpected row count.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Assuming two files with the same number of rows describe the same units.** `baseline` and `followup` both happen to have 20 rows here, but they represent partially different sets of 20 participants — row count alone tells you nothing about whether the *same* individuals are present.
- **Merging without checking the key column's uniqueness first.** If `id` had duplicate values within a single file (a one-to-many relationship), a join could produce more rows than either original file — always check `length(unique(df$id)) == nrow(df)` before assuming a one-to-one relationship.
- **Not deciding in advance what should happen to non-matching ids.** Different join types (next several pages) handle non-matches completely differently — deciding whether you want to keep, drop, or investigate them should happen *before* you pick which join function to use, not after.
:::

## Practice Problem

Use `length(unique(baseline$id)) == nrow(baseline)` to confirm `id` is a genuinely unique key in `baseline_enrollment.csv` (no participant appears twice). Do the same check for `followup_visit.csv`.

## Challenge Problem

Without using any `join()` function yet, use `setdiff()` and `intersect()` to predict, in writing, how many rows each of the following would produce: (a) a merge keeping only ids present in both files, (b) a merge keeping every id from `baseline`, (c) a merge keeping every id from `followup`, (d) a merge keeping every id from either file. Confirm your predictions once you reach the following pages.

## Key Takeaways

- A key is the column that identifies which row in one dataset matches which row in another — here, `id`.
- Two datasets can relate as one-to-one, or with unmatched ids on either side (dropouts, new recruits) — always check both directions with `setdiff()` before merging.
- Predicting the expected row count of a merge *before* running it is the best way to catch a join gone wrong.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `setdiff(baseline$id, followup$id)` return?
   a) ids present in both files
   b) ids in `baseline` that are NOT in `followup`
   c) ids in `followup` that are NOT in `baseline`
   d) The total number of participants

2. **True or false.** Two data frames with the same number of rows must describe the exact same set of individuals.

3. **Predict the output.** Given 20 `baseline` ids and 20 `followup` ids, with 15 ids common to both, how many ids are in `baseline` only?

4. **Short answer.** Why is it useful to compute `setdiff()` and `intersect()` before running any join function?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** ids in `baseline` that are NOT in `followup`.
2. **False.** Row count alone says nothing about whether the same individuals are represented — as seen here, both files have 20 rows but only 15 individuals overlap.
3. 5 (20 total baseline ids minus 15 shared with followup).
4. Knowing the expected overlap ahead of time lets you predict how many rows each join type should produce, so you can immediately catch a mistake if the actual result doesn't match your prediction.

</details>
:::
