# `semi_join()` and `anti_join()` {#sec-semi-anti-join}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor asks two more specific questions, neither of which needs any *columns* combined from the second file — just a **filtered view** of the first file, based on whether a match exists elsewhere. First: "Show me only baseline participants who *do* have a follow-up record — but keep just the baseline columns, nothing from follow-up." Second: "Show me exactly which baseline participants were lost to follow-up — again, baseline columns only."
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Filter one data frame based on whether its key matches another, without adding any of the second data frame's columns, using `dplyr::semi_join()` and `dplyr::anti_join()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Every join so far (`left_join()`, `inner_join()`, `right_join()`, `full_join()`) *added columns* from the second data frame. If a function only *filtered rows* of the first data frame instead, would its result need any new columns at all?
- If `semi_join()` keeps rows *with* a match, what would you expect its natural counterpart to keep instead?
:::

## The R Concept

**`semi_join()`** keeps only the rows of the first data frame that **have** a matching key in the second — but adds none of the second data frame's columns:

```{r}
#| eval: false
semi_join(df1, df2, by = "key_column")
```

```{r}
#| eval: false
semi_join(baseline, followup, by = "id")
```

**`anti_join()`** does the opposite: it keeps only the rows of the first data frame that **do not** have a matching key in the second — again, adding no new columns:

```{r}
#| eval: false
anti_join(df1, df2, by = "key_column")
```

```{r}
#| eval: false
anti_join(baseline, followup, by = "id")
```

Both are **filtering joins** — unlike `left_join()`/`inner_join()`/`right_join()`/`full_join()` (all "mutating joins" that add columns), `semi_join()` and `anti_join()` only ever return a subset of the first data frame's own rows and columns.

## Demonstration

```{r}
#| eval: false
library(dplyr)

baseline <- read.csv("data/baseline_enrollment.csv")
followup <- read.csv("data/followup_visit.csv")

has_followup <- semi_join(baseline, followup, by = "id")

nrow(has_followup)
```

```
[1] 15
```

```{r}
#| eval: false
names(has_followup)
```

```
[1] "id"           "age"          "sex"          "county"       "baseline_bmi"
```

15 rows — the participants with a follow-up match — but notice: only `baseline`'s own 5 columns, with nothing from `followup` attached (unlike `inner_join()`, which would have added `followup_sbp` and `followup_smoker` too).

```{r}
#| eval: false
lost_to_followup <- anti_join(baseline, followup, by = "id")

lost_to_followup$id
```

```
[1] 16 17 18 19 20
```

## Explore

Try `anti_join(followup, baseline, by = "id")` — reversing which file is first. Confirm this returns the 5 *new recruits* (ids 21-25) instead of the 5 dropouts — `anti_join()`'s result depends entirely on which data frame is listed first, exactly like every join so far.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

baseline <- read.csv("data/baseline_enrollment.csv")
followup <- read.csv("data/followup_visit.csv")

lost_to_followup <- anti_join(baseline, followup, by = "id")
new_recruits <- anti_join(followup, baseline, by = "id")

lost_to_followup[, c("id", "county", "age")]
```

```
   id      county age
16 16   Millbrook  45
17 17  Springvale  79
18 18     Oakford  26
19 19 Cedar Falls  80
20 20     Oakford  55
```

```{r}
#| eval: false
new_recruits$id
```

```
[1] 21 22 23 24 25
```

Two clean, separate reports for your professor: exactly the 5 baseline participants lost to follow-up (ids 16-20, ages ranging from 26 to 80), and exactly the 5 new recruits who joined only at follow-up (ids 21-25) — each extracted with a single filtering join, no unwanted columns attached.

## Interpret the Result

`semi_join()` and `anti_join()` answer "which rows match (or don't)?" without the complexity of merged, `NA`-filled columns — often exactly what's needed when the goal is identifying *who* falls into a group, not analyzing their combined data. Compare this to Page 10's `inner_join()`, which answers a related but different question ("give me the matched data, combined") by adding columns rather than just filtering rows.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Reaching for `inner_join()` when you only needed a filter.** If your goal is just "which participants match," `semi_join()` gives a cleaner result than `inner_join()` followed by manually dropping the unwanted follow-up columns.
- **Forgetting `anti_join()`'s direction matters enormously.** `anti_join(baseline, followup, ...)` (dropouts) and `anti_join(followup, baseline, ...)` (new recruits) answer completely different questions — mixing them up reverses which group you're actually reporting on.
- **Expecting `semi_join()` to add any columns at all.** Unlike every other join in this level, `semi_join()`'s result never has more columns than its *first* argument — if you need `followup`'s columns too, `semi_join()` is the wrong tool; reach for `inner_join()` instead.
:::

## Practice Problem

Use `semi_join(followup, baseline, by = "id")` to find which follow-up participants also have a baseline record. How many are there, and does this match `inner_join()`'s row count from Page 10?

## Challenge Problem

Confirm that `nrow(semi_join(baseline, followup, by = "id")) + nrow(anti_join(baseline, followup, by = "id"))` equals `nrow(baseline)` exactly. Explain in words why this must always be true for any two data frames and any key column.

## Key Takeaways

- `semi_join(df1, df2, by = "key")` keeps only `df1`'s rows that have a match in `df2`, with none of `df2`'s columns added.
- `anti_join(df1, df2, by = "key")` keeps only `df1`'s rows that have **no** match in `df2` — again, no new columns.
- Both are filtering joins, distinct from the "mutating joins" (`left_join()`, `inner_join()`, `right_join()`, `full_join()`) that combine columns from both data frames.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** How many columns does `semi_join(baseline, followup, by = "id")` return, if `baseline` has 5 columns and `followup` has 3?
   a) 8 (5 + 3)
   b) 5 (only `baseline`'s columns)
   c) 3 (only `followup`'s columns)
   d) 2 (only the shared key and one other)

2. **True or false.** `anti_join(baseline, followup, by = "id")` and `anti_join(followup, baseline, by = "id")` return the same rows.

3. **Predict the output.** Given 20 baseline ids and 15 with a follow-up match, how many rows would `anti_join(baseline, followup, by = "id")` return?

4. **Short answer.** When would you prefer `semi_join()` over `inner_join()`, even though both only keep matching rows?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** 5 — `semi_join()` never adds columns from the second data frame, regardless of how many it has.
2. **False.** The two calls answer different questions entirely — dropouts (baseline-only) versus new recruits (followup-only) — and generally return completely different rows.
3. 5 rows — the baseline ids (16-20) with no matching follow-up record.
4. When you only need to know *which* rows match, not to actually combine data from both files — `semi_join()` avoids adding unwanted columns that you'd otherwise have to `select()` away after an `inner_join()`.

</details>
:::
