# What Is a Function? {#sec-what-is-a-function}

::: {.pbl-box .pbl-problem}
**The Problem**

Since Level 1, you've typed `mean(participants$age, na.rm = TRUE)` without ever asking what `mean()` actually *is* — you just trust it takes a vector in and gives a number back. Now your professor asks you to compute something `mean()` doesn't do: the percentage of missing values in a column. There's no built-in `pct_missing()` function in R. You'll have to build the tool yourself.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand what a function fundamentally is — a reusable, named piece of code that takes input and produces output — as preparation for writing your own.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- When you call `mean(participants$age, na.rm = TRUE)`, what goes *in* (the input), and what comes *out* (the output)?
- If you had to explain `sum(is.na(x)) / length(x) * 100` to a colleague as a single reusable idea, what would you want to name it?
:::

## The R Concept

A **function** is a named, reusable block of code that takes zero or more **inputs** (called *arguments* or *parameters*), does something with them, and produces an **output** (called the *return value*). Every function you've used since Level 1 — `mean()`, `sum()`, `read.csv()`, `filter()`, `ggplot()` — follows this exact same pattern:

```{r}
#| eval: false
function_name(input1, input2, ...)   # produces some output
```

`mean(participants$age, na.rm = TRUE)` takes a vector (`participants$age`) and a setting (`na.rm = TRUE`) as input, and returns a single number as output. There is nothing magical about built-in functions — they're written in exactly the same style you're about to learn to write your own functions in.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

# Using a built-in function - input goes in, output comes out
mean(participants$age, na.rm = TRUE)
```

```
[1] 47.42553
```

```{r}
#| eval: false
# The pattern you'll replicate: some calculation, wrapped up and named
sum(is.na(participants$age)) / length(participants$age) * 100
```

```
[1] 6
```

The second line computes the percentage of missing `age` values by hand — exactly the calculation you'd want to name and reuse, the same way someone once wrote `mean()` so you'd never have to type `sum(x) / length(x)` yourself.

## Explore

Try running `sum(is.na(participants$systolic_bp)) / length(participants$systolic_bp) * 100` and `sum(is.na(participants$bmi)) / length(participants$bmi) * 100`. Notice you're retyping the same pattern with only the column name changing — exactly the repetition a function is meant to eliminate.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

age_pct_missing <- sum(is.na(participants$age)) / length(participants$age) * 100
bp_pct_missing <- sum(is.na(participants$systolic_bp)) / length(participants$systolic_bp) * 100
bmi_pct_missing <- sum(is.na(participants$bmi)) / length(participants$bmi) * 100

age_pct_missing
```

```
[1] 6
```

```{r}
#| eval: false
bp_pct_missing
```

```
[1] 4
```

```{r}
#| eval: false
bmi_pct_missing
```

```
[1] 5
```

Three correct answers (6%, 4%, 5% missing) — but three nearly identical lines of code, differing only in which column is referenced. This is precisely the kind of repetition the next page's custom function will eliminate.

## Interpret the Result

`age`, `systolic_bp`, and `bmi` are each missing a small percentage of their values (6%, 4%, and 5% respectively) — useful individually, but the *process* of computing each one was identical. Recognizing "I keep typing the same pattern with only one thing changing" is the signal that a function belongs here — the same signal that should have applied a hundred times over across every prior level, now finally something you can act on.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Assuming built-in and custom functions work fundamentally differently.** They don't — `mean()` and a function you write yourself both take input and produce output the exact same way; the only difference is who wrote them and when.
- **Not noticing repeated code as a signal to write a function.** Copy-pasting the same calculation three times with one small change (Page 1's `age`/`systolic_bp`/`bmi` example) is the clearest sign a function is overdue — the next page turns exactly this repetition into one reusable tool.
- **Confusing a function's *definition* with *calling* it.** Writing a function (next page) and calling it are two separate steps — defining `pct_missing <- function(x) {...}` doesn't run anything by itself; you still have to call `pct_missing(participants$age)` afterward to get a result.
:::

## Practice Problem

Compute the percentage missing for `diastolic_bp` and `smoker` by hand, using the same `sum(is.na(x)) / length(x) * 100` pattern. Confirm both are 0%, since Level 7 found no missing values in either column.

## Challenge Problem

Before reading the next page, try to write out in plain English (not R code) the steps a function called `pct_missing` would need to perform, given any column as input. What would its single input be? What single output would it produce?

## Key Takeaways

- A function takes input (arguments) and produces output (a return value) — every function you've used since Level 1 follows this same basic pattern.
- Built-in functions like `mean()` and functions you write yourself work identically in kind — the only difference is authorship.
- Retyping the same calculation with only one small change is the clearest signal that code belongs in a function instead.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What are a function's inputs called?
   a) Return values
   b) Arguments (or parameters)
   c) Comments
   d) Vectors

2. **True or false.** Built-in functions like `mean()` work fundamentally differently from functions a user writes.

3. **Predict the output.** What would `sum(is.na(participants$diastolic_bp)) / length(participants$diastolic_bp) * 100` return, given `diastolic_bp` has no missing values?

4. **Short answer.** What's the clearest signal that a repeated piece of code should be turned into a function?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Arguments (or parameters).
2. **False.** Built-in and user-written functions work identically in kind — both take input and produce output the same way.
3. `0` — with no missing values, the calculation gives 0%.
4. Typing the same (or nearly the same) calculation more than once, with only a small piece — like a column name — changing between repetitions.

</details>
:::
