# Writing Your First Function {#sec-writing-your-first-function}

::: {.pbl-box .pbl-problem}
**The Problem**

You've identified the repeated pattern (previous page): `sum(is.na(x)) / length(x) * 100`. Now you need to actually turn it into a real, callable tool — something you can invoke as `pct_missing(participants$age)` the same way you'd call `mean(participants$age)`, instead of retyping the full calculation every time.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Write your own function using R's `function()` keyword, with a defined input and a returned output.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Looking at `sum(is.na(x)) / length(x) * 100`, what single input (`x`) does this calculation need to run?
- If a function's last line is a value (not assigned to anything), do you think R automatically treats that as the function's output?
:::

## The R Concept

A function is created with the `function()` keyword, listing its argument names in parentheses, followed by a block of code in curly braces `{}`:

```{r}
#| eval: false
function_name <- function(argument1, argument2) {
  # code using argument1 and argument2
  result   # the last evaluated expression is returned automatically
}
```

R automatically returns whatever the **last line** of the function evaluates to — no explicit `return()` needed, though it's allowed and sometimes used for clarity or to exit early:

```{r}
#| eval: false
pct_missing <- function(x) {
  sum(is.na(x)) / length(x) * 100
}
```

## Demonstration

```{r}
#| eval: false
pct_missing <- function(x) {
  sum(is.na(x)) / length(x) * 100
}

participants <- read.csv("data/participants.csv")

pct_missing(participants$age)
```

```
[1] 6
```

```{r}
#| eval: false
pct_missing(participants$systolic_bp)
```

```
[1] 4
```

```{r}
#| eval: false
pct_missing(participants$bmi)
```

```
[1] 5
```

The exact same three answers as the previous page (6%, 4%, 5%) — but now produced by calling one short, named function three times, instead of retyping the full calculation each time.

## Explore

Try `pct_missing(participants$diastolic_bp)`. Confirm it correctly returns `0`, exactly matching your hand calculation from the previous page's practice problem — your function works on *any* vector passed to it, not just the ones you tested first.

## Solve the Problem

```{r}
#| eval: false
pct_missing <- function(x) {
  sum(is.na(x)) / length(x) * 100
}

participants <- read.csv("data/participants.csv")

pct_missing(participants$id)
```

```
[1] 0
```

```{r}
#| eval: false
pct_missing(participants$smoker)
```

```
[1] 0
```

```{r}
#| eval: false
pct_missing(participants$county)
```

```
[1] 0
```

`id`, `smoker`, and `county` all correctly return `0` — none of these columns have any missing values, and `pct_missing()` handles them exactly as correctly as it handled the columns that *did* have missing data, without needing any special-casing.

## Interpret the Result

`pct_missing()` now works as a genuine, general-purpose tool — callable on any column in `participants`, or in any other dataset entirely, without ever needing to see or modify its internal code again. This is the core payoff of writing a function: the calculation was written once, correctly, and now every future use of it is just a short function call.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting the curly braces `{}` for multi-line functions.** A single-line function body can skip the braces (`pct_missing <- function(x) sum(is.na(x)) / length(x) * 100`), but any function with more than one line needs `{}` to group them together.
- **Assuming a function needs an explicit `return()` to produce output.** R automatically returns the last evaluated expression — `return()` is optional and mainly useful for exiting a function early (covered later in this level) or for clarity in longer functions.
- **Confusing the argument name (`x`) with any specific column.** `x` inside the function body is just a placeholder name — it takes on whatever value is passed in when the function is called (`participants$age`, `participants$bmi`, or anything else), and has no meaning outside the function itself.
:::

## Practice Problem

Write a function called `range_width` that takes a numeric vector `x` and returns the difference between its maximum and minimum values (using `na.rm = TRUE` inside), then test it on `participants$bmi`.

## Challenge Problem

Write a function called `mean_and_sd` that takes a numeric vector `x` and returns *both* the mean and standard deviation as a named vector (hint: `c(mean = ..., sd = ...)`), handling missing values appropriately. Test it on `participants$age` and confirm both values match Level 7's earlier findings (mean 47.43, SD 15.03).

## Key Takeaways

- A function is created with `function(arguments) { code }`, and is called later by name with actual values in place of its arguments.
- R automatically returns the last evaluated expression from a function — an explicit `return()` is optional, not required.
- Once written and tested, a function becomes a reusable tool that works on any appropriate input, not just the specific examples used while building it.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does R return from a function by default, with no explicit `return()`?
   a) Nothing — an error occurs
   b) The first evaluated expression
   c) The last evaluated expression
   d) `NULL`, always

2. **True or false.** Curly braces `{}` are required for every function, even one-line ones.

3. **Predict the output.** Given `pct_missing <- function(x) sum(is.na(x)) / length(x) * 100`, what would `pct_missing(c(1, NA, 3, NA))` return?

4. **Find the error.**
   ```r
   pct_missing <- function(x)
     sum(is.na(x)) / length(x) * 100
     cat("Done!")
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **c)** The last evaluated expression.
2. **False.** A single-line function body can omit the curly braces; they're only required when the body spans multiple lines that need to be grouped together.
3. `50` — 2 of the 4 values are `NA`, so `2 / 4 * 100 = 50`.
4. Without curly braces, only the very next single line (`sum(is.na(x)) / length(x) * 100`) is treated as the function's body — the `cat("Done!")` line runs separately, outside the function, every time this code is sourced, rather than being part of `pct_missing()`'s behavior. Wrapping both lines in `{}` would fix this ambiguity.

</details>
:::
