# Default Argument Values {#sec-default-arguments}

::: {.pbl-box .pbl-problem}
**The Problem**

You want a function that reports a numeric variable's mean and standard deviation, rounded for a clean report — but different reports need different precision: a quick internal check might want whole numbers, while a final publication table wants 2 decimal places. Writing two separate functions (`describe_rounded` and `describe_precise`) for what's really the same calculation feels wasteful.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Give a function argument a default value, so it works sensibly when called with fewer arguments, while still allowing that default to be overridden.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If most of the time you want 2 decimal places, but occasionally want more or fewer, would you rather type `digits = 2` every single time, or have `2` be assumed unless you say otherwise?
- From Level 2, `mean(x, na.rm = TRUE)` — is `na.rm` required, or does it have some default value if you omit it entirely?
:::

## The R Concept

A **default argument value** is specified with `=` directly in the function definition, and is used automatically whenever that argument is omitted from the call:

```{r}
#| eval: false
function_name <- function(x, digits = 2) {
  # digits is 2 unless the caller specifies otherwise
}
```

```{r}
#| eval: false
describe_numeric <- function(x, digits = 2) {
  c(
    mean = round(mean(x, na.rm = TRUE), digits),
    sd = round(sd(x, na.rm = TRUE), digits)
  )
}
```

Calling `describe_numeric(participants$age)` uses `digits = 2` automatically; calling `describe_numeric(participants$age, digits = 0)` overrides it explicitly. This is exactly the same pattern behind `mean(x, na.rm = TRUE)`'s `na.rm` argument, which defaults to `FALSE` unless you explicitly set it to `TRUE`.

## Demonstration

```{r}
#| eval: false
describe_numeric <- function(x, digits = 2) {
  c(
    mean = round(mean(x, na.rm = TRUE), digits),
    sd = round(sd(x, na.rm = TRUE), digits)
  )
}

participants <- read.csv("data/participants.csv")

describe_numeric(participants$age)
```

```
 mean    sd
47.43 15.03
```

```{r}
#| eval: false
describe_numeric(participants$age, digits = 0)
```

```
mean   sd
  47   15
```

Calling the function with no `digits` argument at all uses the default (2 decimal places); explicitly passing `digits = 0` overrides that default, rounding to whole numbers instead.

## Explore

Try `describe_numeric(participants$bmi, digits = 4)`. Confirm the function happily accepts any value for `digits`, not just 0 or 2 — the default is only ever a *fallback*, never a restriction on what values are allowed.

## Solve the Problem

```{r}
#| eval: false
describe_numeric <- function(x, digits = 2) {
  c(
    mean = round(mean(x, na.rm = TRUE), digits),
    sd = round(sd(x, na.rm = TRUE), digits)
  )
}

participants <- read.csv("data/participants.csv")

describe_numeric(participants$systolic_bp)
```

```
  mean     sd
129.84  18.87
```

```{r}
#| eval: false
describe_numeric(participants$systolic_bp, digits = 1)
```

```
 mean    sd
129.8  18.9
```

Both calls compute the exact same underlying mean and SD — only the rounding precision changes, controlled entirely by whether `digits` is left at its default (2) or explicitly overridden (1).

## Interpret the Result

`describe_numeric()` now serves both the quick-internal-check case (default 2 decimals) and the more precise or more rounded reporting case (any `digits` value explicitly requested), without needing two separate functions or repeating the `round()`-wrapped calculation twice. A well-chosen default lets the common case stay simple, while still leaving full flexibility available when needed.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Choosing a default that silently produces wrong-looking results for typical use.** A poorly chosen default (like `digits = 10`) could produce overly precise, hard-to-read output for the common case — defaults should reflect what's usually wanted, not just any arbitrary value.
- **Forgetting default arguments can reference other arguments.** More advanced function definitions can have one default depend on another argument's value — while not needed here, it's worth knowing defaults aren't limited to fixed constants.
- **Assuming a default value means an argument can never be changed.** A default is only what's used *if omitted* — every default-valued argument remains fully overridable at the calling site, exactly as shown with `digits = 0` and `digits = 1` above.
:::

## Practice Problem

Modify `pct_missing()` from Page 2 to accept a `digits` argument with a default of `1`, rounding its result before returning it. Test it on `participants$age` both with and without specifying `digits`.

## Challenge Problem

Write a function called `flag_high_bp` that takes a numeric vector of systolic blood pressure values and a `threshold` argument defaulting to `130` (Level 4's hypertension cutoff), returning `TRUE`/`FALSE` for each value indicating whether it exceeds the threshold. Test it once with the default threshold, and once with `threshold = 140` (a stricter cutoff), and compare how many values are flagged under each.

## Key Takeaways

- A default argument value (`argument = default_value` in the function definition) is used automatically whenever that argument is omitted from a call.
- Defaults make the common case simple to call, while every default remains fully overridable when a different value is needed.
- This is the exact same pattern behind built-in functions like `mean(x, na.rm = TRUE)`, whose `na.rm` argument defaults to `FALSE`.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In `function(x, digits = 2)`, when is `digits` equal to `2`?
   a) Always, regardless of how the function is called
   b) Only when the caller explicitly writes `digits = 2`
   c) Whenever the caller omits the `digits` argument entirely
   d) Never — defaults are just documentation, not actual behavior

2. **True or false.** A default argument value can always be overridden by the caller.

3. **Predict the output.** Given `describe_numeric <- function(x, digits = 2) {...}`, what would `digits` equal inside a call to `describe_numeric(participants$bmi)`?

4. **Find the error.**
   ```r
   describe_numeric <- function(x, digits) {
     c(mean = round(mean(x, na.rm = TRUE), digits))
   }
   describe_numeric(participants$age)
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **c)** Whenever the caller omits the `digits` argument entirely.
2. **True.** A default is only ever a fallback for when an argument is omitted — it never prevents the caller from specifying a different value.
3. `2` — the default value, since the call didn't specify `digits`.
4. `digits` has no default value here, so calling `describe_numeric(participants$age)` without specifying `digits` produces an error (`argument "digits" is missing, with no default`). Adding `digits = 2` (or any default) to the function definition would fix this.

</details>
:::
