# Anonymous Functions {#sec-anonymous-functions}

::: {.pbl-box .pbl-problem}
**The Problem**

You want to quickly compute the range (max minus min) of `bmi`, just this once, without needing to reuse the calculation anywhere else afterward. Writing a full named function (`range_width <- function(x) {...}`, Page 2's practice problem) just to use it a single time feels like overkill — you don't need a name for something you'll never call again.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Write a throwaway, unnamed ("anonymous") function for a one-off calculation, without assigning it a name first.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Every function you've written so far has followed the pattern `name <- function(...) {...}`. What part of that pattern is actually *required* for a function to work, versus just a convenience for reusing it later?
- If you never plan to call a function more than once, does it still need a name at all?
:::

## The R Concept

A function doesn't need to be assigned a name to be used immediately — this is called an **anonymous function**:

```{r}
#| eval: false
function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
```

R also provides a shorter syntax for anonymous functions, using a backslash `\` instead of the full word `function`:

```{r}
#| eval: false
\(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
```

Both versions define the exact same function — the `\(x)` shorthand is simply less typing, and is especially common when passing a quick function directly into another function (like `sapply()`, previewed next page), rather than defining it separately first.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

# Anonymous function, called immediately - no name assigned
(function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))(participants$bmi)
```

```
[1] 23
```

```{r}
#| eval: false
# The shorthand \(x) syntax - identical result
(\(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))(participants$bmi)
```

```
[1] 23
```

Both versions compute the exact same range width for `bmi` (38.6 − 15.6 = 23) without ever creating a named, reusable function object — appropriate here, since this calculation is only needed this one time.

## Explore

Try wrapping the same anonymous function around `participants$age` instead of `participants$bmi`. Notice you can reuse the *pattern* freely by copy-pasting it — what you're avoiding isn't reuse of the code itself, but the overhead of formally naming and storing a function object you'll only call once.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

age_range_width <- (\(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))(participants$age)
bmi_range_width <- (\(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE))(participants$bmi)

age_range_width
```

```
[1] 63
```

```{r}
#| eval: false
bmi_range_width
```

```
[1] 23
```

Both range widths (63 for age, 23 for BMI) computed correctly using the shorthand anonymous function syntax, with no named function ever created or stored for later reuse.

## Interpret the Result

Anonymous functions are most useful exactly in situations like this: a one-off calculation, or (as the next page shows) a quick calculation passed directly as an argument into another function like `sapply()`. Using a full named function definition for something this disposable would add unnecessary ceremony to genuinely one-time code.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Using an anonymous function for something you actually need to call repeatedly.** If you find yourself copy-pasting the same anonymous function multiple times, that's a signal (echoing Page 1) it should become a properly named function instead — anonymous functions are for genuine one-off use, not disguised repetition.
- **Forgetting the parentheses needed to call an anonymous function immediately.** Writing `\(x) x + 1(5)` doesn't call the function with `5` as intended — the whole anonymous function definition needs to be wrapped in its own parentheses first, then called: `(\(x) x + 1)(5)`.
- **Confusing the modern `\(x)` shorthand with older R syntax.** The backslash shorthand was introduced in R 4.1 — code written for older R versions will only use the full `function(x)` form; both are equivalent in current R, so either is fine to use going forward.
:::

## Practice Problem

Write an anonymous function (using the `\(x)` shorthand) that computes the coefficient of variation (`sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100`) and apply it immediately to `participants$systolic_bp`.

## Challenge Problem

Explain, in your own words, the difference between when you'd choose a named function (Pages 2-6) versus an anonymous one (this page). Give one example scenario from this course so far where each would be the more appropriate choice.

## Key Takeaways

- An anonymous function has no name and can be defined and called immediately, appropriate for genuinely one-off calculations.
- R's `\(x)` shorthand is equivalent to the full `function(x)` syntax, just less typing — especially handy when passing a quick function directly as an argument to another function.
- Repeated use of the same anonymous function is a signal it should become a properly named, reusable function instead.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `\(x)` mean in modern R syntax?
   a) A comment
   b) Shorthand for defining an anonymous function
   c) A file path separator
   d) An error-checking operator

2. **True or false.** An anonymous function must always be assigned to a variable name before it can be called.

3. **Predict the output.** What would `(\(x) x * 2)(10)` return?

4. **Short answer.** What's the clearest signal that an anonymous function should instead be rewritten as a named one?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Shorthand for defining an anonymous function (equivalent to `function(x)`).
2. **False.** An anonymous function can be defined and called immediately, with no name ever assigned.
3. `20`.
4. If the same anonymous function is being copy-pasted and reused in multiple places, that repetition is the same signal from Page 1 indicating the code should become one properly named, reusable function instead.

</details>
:::
