# Applying a Function Across Columns: `sapply()` and `lapply()` {#sec-apply-family}

::: {.pbl-box .pbl-problem}
**The Problem**

You have `pct_missing()` (Page 2), a working function for one column at a time. But `participants` has 8 columns, and typing `pct_missing(participants$id)`, `pct_missing(participants$age)`, `pct_missing(participants$sex)` ... eight separate times is exactly the repetition functions were supposed to eliminate in the first place. You need to apply `pct_missing()` to *every column at once*, automatically.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Apply a function across every column of a data frame at once using `sapply()` and `lapply()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 2 introduced vectorized operations like `x + 1`, which apply automatically to every element of a vector. Could a *function* similarly be applied automatically to every *column* of a data frame?
- A data frame is really a list of columns underneath. If a function could be applied to every element of a list, would that also apply it to every column of a data frame?
:::

## The R Concept

**`sapply()`** ("simplify apply") applies a function to every element of a vector or list — including every column of a data frame — and simplifies the result into a vector when possible:

```{r}
#| eval: false
sapply(df, function_name)
```

**`lapply()`** ("list apply") does the exact same thing, but always returns a **list** instead of simplifying — useful when the results don't simplify neatly into a single vector:

```{r}
#| eval: false
lapply(df, function_name)
```

## Demonstration

```{r}
#| eval: false
pct_missing <- function(x) {
  sum(is.na(x)) / length(x) * 100
}

participants <- read.csv("data/participants.csv")

sapply(participants, pct_missing)
```

```
         id         age         sex      county systolic_bp diastolic_bp
          0           6           0           0           4            0
        bmi      smoker
          5           0
```

One call, `sapply(participants, pct_missing)`, applied `pct_missing()` to all 8 columns automatically — no more retyping `pct_missing(participants$...)` eight separate times.

```{r}
#| eval: false
lapply(participants, pct_missing)
```

```
$id
[1] 0

$age
[1] 6

$sex
[1] 0
... (output continues as a list, one entry per column)
```

`lapply()` produces the same underlying values, but as a list (one named entry per column) rather than `sapply()`'s simplified named vector.

## Explore

Try `sapply(participants, class)` — applying the built-in `class()` function (Level 2) across every column instead of `pct_missing()`. This confirms `sapply()` works with *any* function, built-in or custom, as long as it's designed to take a single column as input.

## Solve the Problem

```{r}
#| eval: false
pct_missing <- function(x) {
  sum(is.na(x)) / length(x) * 100
}

participants <- read.csv("data/participants.csv")

missing_report <- sapply(participants, pct_missing)

missing_report
```

```
         id         age         sex      county systolic_bp diastolic_bp
          0           6           0           0           4            0
        bmi      smoker
          5           0
```

```{r}
#| eval: false
missing_report[missing_report > 0]
```

```
        age systolic_bp         bmi
          6           4           5
```

`sapply()` computed the missing-value percentage for all 8 columns in one call, and a follow-up logical filter (`missing_report > 0`, Level 2) immediately narrows the result down to just the 3 columns that actually have any missingness — `age` (6%), `systolic_bp` (4%), and `bmi` (5%).

## Interpret the Result

This is exactly the level's opening mission delivered: `pct_missing()` was written once (Page 2), and `sapply()` applied it across the entire dataset automatically — no repetition, no retyping the calculation eight times, and the result immediately usable for further filtering (`missing_report > 0`) to focus attention on only the columns that need it.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `sapply()` and `lapply()` pass one column at a time to the function, not the whole data frame at once.** `pct_missing()` receives `participants$age` as its `x` on one call, then `participants$systolic_bp` on the next — never the entire `participants` data frame in a single call.
- **Expecting `sapply()` to always return a clean vector.** `sapply()` simplifies when it can, but if a function's results don't have a consistent length or type across every column, `sapply()` falls back to returning a list — the same result `lapply()` would have given directly.
- **Applying a function written for a single value (like `bp_category()`, Page 4) directly to an entire column without `sapply()`.** `bp_category()`'s `if` statement expects one number, not a vector — calling `bp_category(participants$systolic_bp)` directly errors; wrapping it as `sapply(participants$systolic_bp, bp_category)` (or vectorizing the function itself) is what correctly applies it across every value.
:::

## Practice Problem

Use `sapply(participants, class)` to check the storage type of every column. Which columns are `"character"`, and which are `"integer"` or `"numeric"`?

## Challenge Problem

Write a function called `n_unique` that returns the number of distinct non-missing values in a vector (hint: use `length(unique(...))` combined with removing `NA`s first). Apply it with `sapply()` across every column of `participants`, and confirm `county` shows exactly 5 distinct values, matching Level 4's finding.

## Key Takeaways

- `sapply(df, function_name)` applies a function to every column of a data frame, simplifying the result into a vector when possible.
- `lapply(df, function_name)` does the same thing but always returns a list, useful when results don't simplify neatly.
- Both `sapply()` and `lapply()` pass one column at a time to the function — the function itself is still written exactly as if handling a single vector, exactly as on Pages 2-4.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `sapply(participants, pct_missing)` do?
   a) Applies `pct_missing()` to the entire data frame as one single input
   b) Applies `pct_missing()` separately to each column, simplifying the results into a vector
   c) Filters `participants` to only rows with missing values
   d) Modifies `participants` in place

2. **True or false.** `lapply()` always returns a simplified vector, exactly like `sapply()`.

3. **Predict the output.** Given `sapply(participants, pct_missing)`'s result, what would `missing_report["bmi"]` return?

4. **Find the error.**
   ```r
   bp_category(participants$systolic_bp)
   # intended: classify every participant's blood pressure at once
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Applies `pct_missing()` separately to each column, simplifying the results into a vector.
2. **False.** `lapply()` always returns a list; `sapply()` attempts to simplify the result into a vector (or matrix) when possible.
3. `5` — the percentage of missing `bmi` values.
4. `bp_category()` (Page 4) was written expecting a single value, using `if`, which errors on a vector input. It should be wrapped with `sapply()`: `sapply(participants$systolic_bp, bp_category)`, applying the single-value function across every element correctly.

</details>
:::
