# Repeating Values with `rep()` {#sec-rep}

::: {.pbl-box .pbl-problem}
**The Problem**

Your study enrolled participants at two clinics: the first 50 rows in your dataset came from the Riverdale clinic, and the last 50 came from the Oakford clinic. You need a `site` label for every one of the 100 rows — but typing `"Riverdale"` fifty times and `"Oakford"` fifty times is exactly the kind of repetitive task you'd hope R could handle for you.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Generate a vector of repeated values automatically, without typing each repetition individually.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Is "Riverdale, Riverdale, Riverdale, ..., Oakford, Oakford, Oakford" the same pattern as "Riverdale, Oakford, Riverdale, Oakford, ..."? Which one matches your actual dataset?
- What two pieces of information would you need to describe a repeating pattern: what to repeat, and what else?
:::

## The R Concept

**`rep()`** repeats a value or vector according to one of two different logics, and the difference between them matters a great deal:

```{r}
#| eval: false
rep(x, times = n)   # repeats the WHOLE vector n times, in sequence
rep(x, each = n)    # repeats EACH element n times before moving to the next
```

```{r}
#| eval: false
rep(c("A", "B"), times = 3)   # "A" "B" "A" "B" "A" "B"
rep(c("A", "B"), each = 3)    # "A" "A" "A" "B" "B" "B"
```

Notice both calls use the same inputs but produce entirely different orderings — `times` repeats the *pattern*, `each` repeats *every element* before advancing.

## Demonstration

```{r}
#| eval: false
rep(c("Riverdale", "Oakford"), times = 2)
```

```
[1] "Riverdale" "Oakford"   "Riverdale" "Oakford"
```

```{r}
#| eval: false
rep(c("Riverdale", "Oakford"), each = 2)
```

```
[1] "Riverdale" "Riverdale" "Oakford"   "Oakford"
```

```{r}
#| eval: false
# Repeating a single value
rep("Enrolled", times = 5)
```

```
[1] "Enrolled" "Enrolled" "Enrolled" "Enrolled" "Enrolled"
```

## Explore

Try `rep(c("Riverdale", "Oakford"), each = 2, times = 2)` — `each` and `times` can be combined. Predict the output before running it: does `each` or `times` get applied first?

## Solve the Problem

```{r}
#| eval: false
# 50 participants from Riverdale, followed by 50 from Oakford
site <- rep(c("Riverdale", "Oakford"), each = 50)

length(site)
```

```
[1] 100
```

```{r}
#| eval: false
table(site)
```

```
site
  Oakford Riverdale
       50        50
```

Using `each = 50` (not `times = 50`) correctly produces 50 consecutive `"Riverdale"` values followed by 50 consecutive `"Oakford"` values — matching exactly how the real dataset was collected, clinic by clinic.

## Interpret the Result

`table(site)` confirms 50 participants are labeled `"Oakford"` and 50 are labeled `"Riverdale"`, generated correctly in one line rather than 100 individual entries — and critically, in the right *order* to match the two-clinic-block structure of the actual data collection.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Using `times` when you meant `each` (or vice versa).** `rep(c("A", "B"), times = 50)` alternates A/B one hundred times; `rep(c("A", "B"), each = 50)` produces fifty A's followed by fifty B's — these look similar in code but produce completely different data.
- **Assuming `rep()` output length always matches your target automatically.** Always check `length()` after building a repeated vector, especially before attaching it as a new column to a data frame with a fixed number of rows.
- **Forgetting `rep()` only repeats an existing pattern.** If the real-world groups aren't evenly sized (say 62 Riverdale participants and 38 Oakford), a simple `rep(..., each = 50)` will be wrong — you'd need `rep(c("Riverdale", "Oakford"), times = c(62, 38))`, where `times` can also take a vector specifying how many times each element repeats individually.
:::

## Practice Problem

Create a vector `dose_group` representing 3 participants assigned to `"Placebo"` and 3 assigned to `"Treatment"`, in that block order (not alternating), using `rep()` with the appropriate argument.

## Challenge Problem

A study has 3 sites with uneven enrollment: 20 participants at Site A, 35 at Site B, and 45 at Site C (100 total). Use `rep()` with a vector passed to `times` (as hinted in the Common Mistakes box) to generate the full 100-length `site` vector in the correct block order.

## Key Takeaways

- `rep(x, times = n)` repeats the entire pattern `n` times; `rep(x, each = n)` repeats every individual element `n` times before moving to the next.
- These two produce very different orderings from the same inputs — always double-check which one matches your actual data structure.
- `times` can also accept a vector, letting each element repeat a different number of times — useful for unevenly sized groups.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `rep(c("A", "B"), each = 2)` produce?
   a) `"A" "B" "A" "B"`
   b) `"A" "A" "B" "B"`
   c) `"A" "A" "A" "B"`
   d) An error

2. **True or false.** `rep(x, times = n)` and `rep(x, each = n)` always produce the same output.

3. **Predict the output.**
   ```r
   rep(1:3, times = 2)
   ```

4. **Find the error.** A student wants 10 consecutive `"Male"` labels followed by 10 consecutive `"Female"` labels, but writes:
   ```r
   sex <- rep(c("Male", "Female"), times = 10)
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `"A" "A" "B" "B"`
2. **False.** `times` repeats the whole pattern in sequence; `each` repeats every element before moving on — they produce different orderings.
3. `[1] 1 2 3 1 2 3`
4. The student used `times` when they meant `each` — `rep(c("Male", "Female"), times = 10)` alternates Male/Female 10 times (20 values total, alternating), not 10 Males followed by 10 Females. It should be `rep(c("Male", "Female"), each = 10)`.

</details>
:::
