# Positive Indexing {#sec-positive-indexing}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor asks: "What was the age of the very first participant enrolled? And can you show me the ages of the first ten participants, so I can spot-check the data entry?" You have a 100-value `age` vector, but you've only ever printed the *whole* thing at once — you don't yet know how to pull out just one value, or just a few, from a specific position.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Extract specific elements from a vector by their position, using square brackets.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If someone asked you for "the 3rd item on this list," how would you count to find it — starting from 0 or from 1?
- What if you wanted several positions at once, like the 1st, 2nd, and 3rd — could you reuse something you learned two pages ago (sequences) to describe "1 through 10"?
:::

## The R Concept

**Indexing** means extracting one or more elements from a vector using square brackets, `[ ]`, containing the position(s) you want:

```{r}
#| eval: false
x[position]
```

Critically, **R is 1-indexed** — the first element is at position `1`, not `0` (unlike some other programming languages you may encounter later, such as Python). This is one of the most important structural facts to internalize early.

```{r}
#| eval: false
ages <- c(46, 53, 59, 49, 61)
ages[1]      # the FIRST element - not the second
ages[3]      # the third element
ages[1:3]    # the first three elements, using the sequence range from the last page
```

You can also pass a vector of specific, non-consecutive positions: `ages[c(1, 3, 5)]` returns the 1st, 3rd, and 5th elements.

## Demonstration

```{r}
#| eval: false
ages <- c(46, 53, 59, 49, 61, 18, 62, 68, 33, 36)

ages[1]
```

```
[1] 46
```

```{r}
#| eval: false
ages[1:10]
```

```
 [1] 46 53 59 49 61 18 62 68 33 36
```

```{r}
#| eval: false
ages[c(2, 5, 9)]
```

```
[1] 53 61 33
```

## Explore

Try `ages[11]` — one position past the end of a 10-element vector. What does R return, and why do you think it doesn't produce an error? Then try `ages[0]` — an unusual edge case worth seeing once.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

# Age of the very first participant enrolled
participants$age[1]
```

```
[1] 46
```

```{r}
#| eval: false
# Ages of the first ten participants
participants$age[1:10]
```

```
 [1] 46 53 59 49 61 18 62 68 33 36
```

## Interpret the Result

The first participant enrolled was 46 years old, and the first ten participants ranged across ages including 18, 33, and 68 — exactly the kind of quick spot-check your supervisor asked for, pulled directly from specific positions in the `age` column without scrolling through the entire 100-row dataset.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Assuming R is 0-indexed.** `ages[0]` does not return the first element — it returns an empty vector of length zero, since position `0` doesn't correspond to any real element in R's 1-indexed system.
- **Using `()` instead of `[]`.** `ages(1)` is not valid indexing syntax — parentheses are for function calls, not for extracting elements from a vector.
- **Indexing past the end silently.** `ages[11]` on a 10-element vector doesn't error — it returns `NA`, since R interprets it as "the value at a position that doesn't exist." This can silently introduce missing values into further calculations if you're not paying attention.
:::

## Practice Problem

Using `participants$bmi`, extract the BMI of the 15th participant, and separately, the BMIs of the first 5 participants.

## Challenge Problem

Extract the ages of the 1st, 50th, and 100th participants in a single line of code (hint: you'll need `c()` inside the square brackets). Then explain, in one sentence, why `participants$age[1, 50, 100]` (with commas but no `c()`) would not work the way you might expect.

## Key Takeaways

- Square brackets `[ ]` extract elements from a vector by position; R is 1-indexed, so the first element is at position `1`.
- A range (`1:10`) or an arbitrary set of positions (`c(2, 5, 9)`) can both be used inside `[ ]`.
- Indexing past the end of a vector doesn't error — it silently returns `NA`, which is worth watching for.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `x[1]` return for `x <- c(10, 20, 30)`?
   a) `20`
   b) `10`
   c) An error, since indexing starts at 0
   d) The entire vector

2. **True or false.** R is 0-indexed, like Python.

3. **Predict the output.**
   ```r
   x <- c(10, 20, 30, 40)
   x[c(1, 4)]
   ```

4. **Find the error.**
   ```r
   ages(3)
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `10` — the first element, since R is 1-indexed.
2. **False.** R is 1-indexed; the first element is at position 1, not 0.
3. `[1] 10 40`
4. Parentheses are used for function calls, not indexing. It should be `ages[3]`.

</details>
:::
