# Named Vectors {#sec-named-vectors}

::: {.pbl-box .pbl-problem}
**The Problem**

You've calculated the mean systolic blood pressure for each of the five counties in your survey — Cedar Falls, Millbrook, Oakford, Riverdale, and Springvale — and stored the five results in a plain numeric vector. A week later, you look at that vector again: `128.16, 133.30, 129.71, 124.36, 131.46`. Which number belonged to which county? You'd have to recompute everything to remember.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Attach labels directly to the values in a vector, so each number is self-documenting and can be looked up by name instead of position.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If a plain vector only lets you retrieve a value by position (`x[2]`), what extra piece of information would let you retrieve it by a meaningful label instead?
- Have you seen anything like this before — a set of labeled values you can look up by name rather than counting positions? (Hint: a data frame's *columns* already work this way.)
:::

## The R Concept

A vector can have a **name attached to every element**, either by supplying names directly inside `c()`, or by assigning them afterward with `names()`:

```{r}
#| eval: false
c(name1 = value1, name2 = value2, ...)
```

```{r}
#| eval: false
x <- c(10, 20, 30)
names(x) <- c("a", "b", "c")
```

Once named, you can index by name using `[ ]` with a quoted string, in addition to the position-based indexing you already know:

```{r}
#| eval: false
x["b"]     # retrieves the element named "b", regardless of its position
```

`names()` works both ways: calling it alone shows the current names; assigning to it (`names(x) <- ...`) sets new ones.

## Demonstration

```{r}
#| eval: false
mean_bp_by_county <- c(
  "Cedar Falls" = 128.16,
  "Millbrook"   = 133.30,
  "Oakford"     = 129.71,
  "Riverdale"   = 124.36,
  "Springvale"  = 131.46
)

mean_bp_by_county
```

```
Cedar Falls    Millbrook      Oakford    Riverdale   Springvale
     128.16       133.30       129.71       124.36       131.46
```

```{r}
#| eval: false
mean_bp_by_county["Millbrook"]
```

```
Millbrook
    133.3
```

```{r}
#| eval: false
names(mean_bp_by_county)
```

```
[1] "Cedar Falls" "Millbrook"   "Oakford"     "Riverdale"   "Springvale"
```

## Explore

Try `mean_bp_by_county[["Millbrook"]]` (double brackets) instead of single brackets — notice the output no longer includes the name label, just the bare number. This distinction between `[ ]` and `[[ ]]` will matter even more once you reach lists later in this level. Then try retrieving a county that doesn't exist, like `mean_bp_by_county["Oak Grove"]` — what does R return?

## Solve the Problem

```{r}
#| eval: false
# Building the named vector directly from the data, using tools from earlier in this level
participants <- read.csv("data/participants.csv")

counties <- unique(participants$county)
mean_bp_by_county <- sapply(counties, function(cty) {
  mean(participants$systolic_bp[participants$county == cty], na.rm = TRUE)
})
names(mean_bp_by_county) <- counties

# A week later, you can still look this up instantly and unambiguously
mean_bp_by_county["Millbrook"]
```

```
Millbrook
   133.30
```

::: {.callout-note}
## A quick note on `sapply()`
This solution uses `sapply()` to avoid repeating the same filter-and-mean calculation five times by hand — you haven't formally learned it yet (that's covered in Level 10), so don't worry about the syntax here. The point of this page is what happens *after*: attaching names so the result is self-documenting.
:::

## Interpret the Result

`mean_bp_by_county["Millbrook"]` instantly and unambiguously returns 133.30 mmHg — the mean systolic blood pressure among Millbrook participants — without you needing to remember whether Millbrook was the first, second, or third value in the vector. The name *is* the lookup key, which is far more durable than remembering positional order.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Misspelling or mis-casing a name when looking it up.** `mean_bp_by_county["millbrook"]` (lowercase) returns `NA`, not an error — names are matched exactly and are case-sensitive, just like object names in Level 1.
- **Confusing `[ ]` and `[[ ]]` on named vectors.** `x["b"]` returns a named single-element vector; `x[["b"]]` returns just the bare value with no name attached — a subtle difference that becomes more important with lists, later in this level.
- **Forgetting names can become out of sync.** If you reorder or filter a named vector's values without also reordering `names()`, the labels will silently no longer match the values they're attached to — always build both together, as in the solution above.
:::

## Practice Problem

Run `table(participants$county)` to get the real participant counts per county, then build a named vector `n_participants_by_county` from that output. Retrieve just the Springvale count by name.

## Challenge Problem

You have `mean_bp_by_county` from this page. Use `which.max()` (look it up with `?which.max` if needed, applying what you learned about getting help in Level 1) to identify which county has the *highest* mean blood pressure, and extract both its name and its value in one expression.

## Key Takeaways

- Vectors can carry names alongside their values, either set inside `c(name = value)` or assigned afterward with `names()`.
- Named elements can be retrieved by name using `x["name"]`, which is more durable and self-documenting than remembering positional order.
- `x["name"]` keeps the name attached to the result; `x[["name"]]` returns just the bare value.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** How do you retrieve the element named `"b"` from vector `x`, keeping its name?
   a) `x[2]`
   b) `x["b"]`
   c) `x(b)`
   d) `x.b`

2. **True or false.** Name-based indexing in R is case-sensitive.

3. **Predict the output.**
   ```r
   x <- c(low = 10, high = 90)
   x["low"]
   ```

4. **Find the error.** A student wants the mean BP for Riverdale and writes:
   ```r
   mean_bp_by_county["riverdale"]
   ```
   but gets `NA`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `x["b"]`
2. **True.** Names must match exactly, including capitalization.
3. `  low \n   10` (a named single-element vector showing `low: 10`).
4. The name was written in lowercase (`"riverdale"`) but the actual name is `"Riverdale"` (capitalized) — names are case-sensitive, so the lookup silently returns `NA` instead of erroring.

</details>
:::
