# Accessing Rows {#sec-accessing-rows}

::: {.pbl-box .pbl-problem}
**The Problem**

Your supervisor flags an unusual result and asks: "Can you pull up participant #42's entire record? I want to see everything we have on them — age, county, blood pressure, all of it." You know how to grab a single *column* across all rows. Now you need the opposite: a single *row*, across all columns.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Extract one or more complete rows from a data frame, keeping every column.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- A data frame is two-dimensional — rows and columns. If vector indexing used one number in `[ ]`, how many numbers (or positions) might a data frame need?
- Do you think leaving one of those two positions blank might mean something specific, like "give me all of them"?
:::

## The R Concept

Data frames are indexed with **two** positions inside the brackets, separated by a comma: `df[rows, columns]`. Leaving either side blank means "all of them":

```{r}
#| eval: false
df[row_number, ]     # one full row, every column (note the comma!)
df[, column_number]  # one full column, every row
df[row_number, column_number]  # a single specific cell
```

This comma is what distinguishes row-and-column indexing from the single-bracket column shortcut you saw on the last page (`df["age"]`, no comma, meant something entirely different — a column by name).

```{r}
#| eval: false
participants[42, ]      # participant #42's complete record, every column
participants[1:5, ]     # the first 5 participants, every column
participants[42, "age"] # just participant #42's age - a specific cell
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

participants[42, ]
```

```
   id age  sex    county systolic_bp diastolic_bp  bmi smoker
42 42  53 Male Springvale         135           86 30.4    Yes
```

```{r}
#| eval: false
participants[1:3, ]
```

```
  id age    sex   county systolic_bp diastolic_bp  bmi smoker
1  1  46 Female  Oakford         125           79 35.1     No
2  2  53 Female Riverdale         123           80 27.7     No
3  3  59 Female  Oakford         130           77 28.3     No
```

```{r}
#| eval: false
participants[42, "age"]
```

```
[1] 53
```

## Explore

Try `participants[42]` — the same thing, but *without* the comma. Notice this doesn't give you the row at all; it behaves like column-selection instead (specifically, it tries to select the 42nd *column*, which doesn't exist in an 8-column data frame, producing an error or unexpected result). The comma is not optional decoration — it fundamentally changes what R thinks you're asking for.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

# Participant #42's entire record, every column
participants[42, ]
```

```
   id age  sex    county systolic_bp diastolic_bp  bmi smoker
42 42  53 Male Springvale         135           86 30.4    Yes
```

Your supervisor now has the complete picture for participant #42 in one line: 53 years old, male, from Springvale, with a systolic BP of 135, diastolic of 86, BMI of 30.4, and a current smoker.

## Interpret the Result

`participants[42, ]` returns a one-row data frame containing every variable recorded for that single participant — exactly the "everything we have on them" your supervisor asked for, rather than a bare vector, since a row genuinely mixes multiple data types (numbers and text together) that only a data frame (not a vector) can hold.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting the comma.** `participants[42, ]` (row 42, all columns) and `participants[42]` (attempting to select the 42nd column) are completely different operations — the comma is essential, not optional.
- **Confusing row *position* with the `id` column's value.** `participants[42, ]` returns the 42nd *row* in the data frame — this only matches `id == 42` because the data happens to be sorted and unfiltered; after any filtering or reordering (Level 5 onward), row position and `id` value can diverge completely.
- **Leaving a trailing comma off by habit from vector indexing.** Vectors use one index (`x[5]`); data frames need two, separated by a comma, even when you want "all columns" (in which case the space after the comma is simply left blank, not omitted).
:::

## Practice Problem

Extract the complete records for participants 10 through 15 (using a range) from `participants`.

## Challenge Problem

Extract just participant #7's `county` and `smoker` values in a single expression (hint: the column position can also be a vector of column names, like `c("county", "smoker")`, inside the same two-position bracket syntax).

## Key Takeaways

- Data frames use two-position indexing, `df[rows, columns]`, separated by a comma; leaving one side blank means "all."
- Omitting the comma entirely changes the meaning of the expression — it no longer refers to rows at all.
- A single row returns a one-row data frame (since rows can mix types); a single column returns a vector (via `$` or `[[ ]]`) or a one-column data frame (via `[ ]`, from the previous page).

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `df[5, ]` return?
   a) The 5th column
   b) The 5th row, all columns
   c) The first 5 rows
   d) An error

2. **True or false.** `df[5, ]` and `df[5]` do the same thing.

3. **Predict the output.**
   ```r
   df <- data.frame(x = c(10, 20, 30), y = c("a", "b", "c"))
   df[2, ]
   ```

4. **Find the error.** A student wants participant #10's record and writes:
   ```r
   participants[10]
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The 5th row, all columns.
2. **False.** `df[5, ]` returns row 5 (all columns); `df[5]` attempts to select column 5 — the comma changes the meaning entirely.
3.
   ```
     x y
   2 20 b
   ```
4. Missing the comma — `participants[10]` attempts to select the 10th column, not row 10. It should be `participants[10, ]`.

</details>
:::
