# Matrices {#sec-matrices}

::: {.pbl-box .pbl-problem}
**The Problem**

A lab tracks each participant's systolic blood pressure across four consecutive monthly visits. You could store this as a data frame — but every value here is the exact same type (numeric blood pressure readings), and you want to do genuinely mathematical operations across the whole grid, like computing each participant's visit-to-visit change. A data frame's flexibility (mixed column types) isn't actually needed here, and a simpler, more strictly numeric grid would do.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Create and index a matrix — a rectangular grid where every single value shares the same type.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If every value in your grid is a number, what would you lose by using a data frame instead of something simpler?
- A data frame's columns can each be a different type. What might it mean for a structure to *require* every value, across the whole grid, to be the same type?
:::

## The R Concept

A **matrix** is a two-dimensional grid where, unlike a data frame, *every* value must be the same type (all numeric, or all character — never mixed). Matrices are common in more mathematical or statistical contexts (like a correlation matrix in Level 11) where genuine matrix arithmetic matters.

```{r}
#| eval: false
matrix(data, nrow = ..., ncol = ..., byrow = FALSE)
```

```{r}
#| eval: false
bp_matrix <- matrix(
  c(120, 125, 118, 130, 132, 128, 145, 140, 138),
  nrow = 3, ncol = 3, byrow = TRUE
)
```

The `byrow` argument controls how the flat vector of values gets arranged into the grid: `byrow = TRUE` fills row by row; the default, `byrow = FALSE`, fills column by column instead — a frequent source of confusion, since both produce a valid-looking matrix with the *same numbers in different positions*.

Indexing a matrix uses the same two-position bracket syntax as a data frame: `m[row, col]`.

## Demonstration

```{r}
#| eval: false
bp_matrix <- matrix(
  c(120, 125, 118,
    130, 132, 128,
    145, 140, 138),
  nrow = 3, ncol = 3, byrow = TRUE
)
bp_matrix
```

```
     [,1] [,2] [,3]
[1,]  120  125  118
[2,]  130  132  128
[3,]  145  140  138
```

```{r}
#| eval: false
# Row 1, column 2 - a specific reading
bp_matrix[1, 2]
```

```
[1] 125
```

```{r}
#| eval: false
# All of row 2 (one participant's readings across all visits)
bp_matrix[2, ]
```

```
[1] 130 132 128
```

```{r}
#| eval: false
class(bp_matrix)
```

```
[1] "matrix" "array"
```

## Explore

Rebuild the same matrix with `byrow = FALSE` instead (or simply omit `byrow` entirely, since `FALSE` is the default) and compare the printed result to the one above — same nine numbers, arranged completely differently. This is exactly the kind of mistake that produces a matrix that *looks* fine but silently has every value in the wrong position.

## Solve the Problem

```{r}
#| eval: false
# Three participants (rows), four monthly visits (columns)
bp_matrix <- matrix(
  c(120, 125, 118, 122,
    130, 132, 128, 126,
    145, 140, 138, 135),
  nrow = 3, ncol = 4, byrow = TRUE
)
bp_matrix
```

```
     [,1] [,2] [,3] [,4]
[1,]  120  125  118  122
[2,]  130  132  128  126
[3,]  145  140  138  135
```

```{r}
#| eval: false
# Change from visit 1 to visit 4, for every participant at once
bp_matrix[, 4] - bp_matrix[, 1]
```

```
[1]   2  -4 -10
```

## Interpret the Result

The change vector `2, -4, -10` shows that participant 1's blood pressure rose slightly by visit 4, while participants 2 and 3 saw decreases of 4 and 10 mmHg, respectively — a calculation performed across the entire matrix in a single line, exactly the kind of operation a matrix's uniform numeric structure makes easy.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing `byrow = TRUE` and `byrow = FALSE`.** The same flat vector of numbers produces a completely different arrangement depending on this argument — always verify the printed matrix matches your intended layout before trusting any calculation on it.
- **Assuming a matrix can mix types like a data frame.** `matrix(c(1, "two", 3), nrow = 1)` silently coerces everything to character, exactly like `c()` did in the vectors page — matrices inherit this same all-one-type constraint.
- **Mixing up matrix indexing with data frame indexing habits.** Both use `m[row, col]`, but a matrix has no column *names* by default (just `[,1]`, `[,2]`, etc.) unless you explicitly assign them with `colnames()` — attempting `bp_matrix["visit_1"]` without first setting names will fail.
:::

## Practice Problem

Create a 2×3 matrix of BMI values `c(22.1, 24.5, 27.3, 19.8, 31.2, 26.0)` filled by row, and extract the entire second row.

## Challenge Problem

Using the `bp_matrix` from the solution above, calculate the mean blood pressure for each participant across their four visits (hint: look up `rowMeans()` using the help-page skills from Level 1 — `?rowMeans`).

## Key Takeaways

- A matrix is a two-dimensional grid where every value must share the same type — unlike a data frame, which allows different types per column.
- `byrow` controls whether the input vector fills the matrix row by row or column by column, and mixing this up silently misplaces every value.
- Matrix indexing uses the same `m[row, col]` syntax as data frames, making genuinely mathematical, whole-grid operations (like `bp_matrix[, 4] - bp_matrix[, 1]`) straightforward.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What must be true of every value in a matrix?
   a) All values must be unique
   b) All values must be the same type
   c) All values must be positive
   d) All values must be named

2. **True or false.** `byrow = TRUE` and `byrow = FALSE` always produce the same matrix from the same input vector.

3. **Predict the output.**
   ```r
   m <- matrix(1:6, nrow = 2, byrow = TRUE)
   m[1, ]
   ```

4. **Find the error.**
   ```r
   matrix(c(1, "two", 3), nrow = 1)
   ```
   A student expects this to raise an error because it mixes a number and text. What actually happens instead?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** All values must be the same type.
2. **False.** They arrange the same values differently — row by row versus column by column — producing different matrices unless the data happens to be symmetric.
3. `[1] 1 2 3` (the first row, since `byrow = TRUE` fills `1, 2, 3` into row 1 and `4, 5, 6` into row 2).
4. It doesn't error — R silently coerces everything to character, exactly like `c()` does, producing `matrix(c("1", "two", "3"), nrow = 1)`.

</details>
:::
