# Lists {#sec-lists}

::: {.pbl-box .pbl-problem}
**The Problem**

You've written a small summary of your analysis: the study name (`"Community Health Survey"`, a single piece of text), the full `participants` data frame, a numeric vector of the three counties with the highest mean blood pressure, and a logical flag for whether the analysis is complete. These four things have completely different shapes and types — there's no vector, matrix, or single data frame that could hold a data frame, a vector, and a single value all together.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Bundle several different, unrelated objects — of any type or shape — into a single container.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Every structure you've learned so far in this level (vectors, matrices, data frames) enforces some kind of consistency — same type, or same length. What if you needed a container with *no* such restriction at all?
- If you already know `[[ ]]` extracts a bare value from a named vector, could something similar work for extracting one whole, distinct item out of a bundle of very different items?
:::

## The R Concept

A **list** is R's most flexible structure: it can hold any combination of objects — vectors, data frames, matrices, single values, even other lists — with no requirement that they share a type, length, or shape.

```{r}
#| eval: false
list(name1 = object1, name2 = object2, ...)
```

```{r}
#| eval: false
analysis_summary <- list(
  study_name = "Community Health Survey",
  data = participants,
  top_counties = c("Millbrook", "Springvale", "Oakford"),
  complete = TRUE
)
```

Accessing list elements uses the same double-bracket idea from named vectors, but it matters even more here: **`[[ ]]`** extracts one element exactly as it is (a data frame stays a data frame, a vector stays a vector); **`[ ]`** returns a *smaller list* containing that element, still wrapped in list packaging — usually not what you want when you're after the actual content.

```{r}
#| eval: false
analysis_summary[["study_name"]]   # the bare character value
analysis_summary["study_name"]     # a one-element LIST containing that value
analysis_summary$study_name        # $ also works for named list elements, like data frames
```

## Demonstration

```{r}
#| eval: false
analysis_summary <- list(
  study_name = "Community Health Survey",
  n_participants = 100,
  top_counties = c("Millbrook", "Springvale", "Oakford"),
  complete = TRUE
)

analysis_summary
```

```
$study_name
[1] "Community Health Survey"

$n_participants
[1] 100

$top_counties
[1] "Millbrook"   "Springvale"  "Oakford"

$complete
[1] TRUE
```

```{r}
#| eval: false
analysis_summary[["top_counties"]]
```

```
[1] "Millbrook"   "Springvale"  "Oakford"
```

```{r}
#| eval: false
class(analysis_summary["top_counties"])
```

```
[1] "list"
```

```{r}
#| eval: false
class(analysis_summary[["top_counties"]])
```

```
[1] "character"
```

## Explore

Try `analysis_summary$data <- participants` to add the entire `participants` data frame as a fifth element of the list — demonstrating that a list can hold something as large and complex as an entire dataset alongside small individual values. Then run `length(analysis_summary)` — what do you think it counts: total values across everything, or top-level elements?

## Solve the Problem

```{r}
#| eval: false
analysis_summary <- list(
  study_name = "Community Health Survey",
  data = participants,
  top_counties = c("Millbrook", "Springvale", "Oakford"),
  complete = TRUE
)

length(analysis_summary)
```

```
[1] 4
```

```{r}
#| eval: false
analysis_summary$study_name
```

```
[1] "Community Health Survey"
```

```{r}
#| eval: false
nrow(analysis_summary$data)
```

```
[1] 100
```

The list `analysis_summary` now holds all four genuinely different pieces — a piece of text, an entire 100-row data frame, a 3-element character vector, and a single logical value — as one coherent object, with each piece still fully accessible and usable in its original form.

## Interpret the Result

`length(analysis_summary)` returns `4`, confirming the list has four *top-level* elements — it does not count all 100 rows inside the nested data frame, or every character in `study_name`; it counts the list's own direct components. This distinction between "top-level elements" and "everything nested inside" becomes important as you work with more complex, real-world list structures — including, later in the course, the output of statistical models, which are almost always returned as lists.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing `[ ]` and `[[ ]]` on a list.** `analysis_summary["study_name"]` returns a one-element *list* (still wrapped); `analysis_summary[["study_name"]]` returns the bare character value inside — nearly every bug involving "why does this look like text but won't behave like text" traces back to this exact distinction.
- **Assuming `length()` counts everything nested inside.** `length(analysis_summary)` counts the 4 top-level elements, not the 100 rows of the nested data frame or the 3 elements of the nested vector — each nested object has its own separate `length()` or `nrow()`.
- **Forgetting that list elements can be added or changed with `$` or `[[ ]]`, just like data frame columns.** `analysis_summary$new_field <- "value"` adds a new list element on the fly, exactly like adding a new column to a data frame.
:::

## Practice Problem

Create a list `patient_record` with elements `id = 42`, `age = 53`, `vitals = c(135, 86)` (systolic and diastolic BP), and `notes = "Follow-up needed"`. Extract just the `vitals` element as a bare vector (not wrapped in a list).

## Challenge Problem

Add a fifth element to `patient_record` called `flagged`, set to `TRUE`, using `$`. Then write the code to check `length(patient_record)` before and after adding it, confirming it increases by exactly one — regardless of how large or small the new element itself is.

## Key Takeaways

- A list can hold any combination of objects — different types, different lengths, different shapes — with no consistency requirement at all.
- `[[ ]]` (or `$`, for named elements) extracts an element exactly as it is; `[ ]` returns a smaller list still wrapping that element.
- `length()` on a list counts top-level elements only, not everything nested inside them — a distinction that matters once lists hold complex objects like entire data frames.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What can a list hold?
   a) Only vectors of the same type
   b) Only numeric values
   c) Any combination of objects, regardless of type or shape
   d) Only data frames

2. **True or false.** `mylist[["x"]]` and `mylist["x"]` always return the exact same thing.

3. **Predict the output.**
   ```r
   l <- list(a = 1, b = c(2, 3), c = "text")
   length(l)
   ```

4. **Find the error.** A student expects `mylist["score"]` to behave like a bare number so they can do math with it directly, but gets an error.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **c)** Any combination of objects, regardless of type or shape.
2. **False.** `[[ ]]` returns the bare element; `[ ]` returns a one-element list still wrapping it — they are structurally different.
3. `[1] 3` — three top-level elements (`a`, `b`, `c`), regardless of how many values are nested inside `b`.
4. `mylist["score"]` returns a one-element list, not a bare number — arithmetic functions can't operate on a list directly. The fix is `mylist[["score"]]` or `mylist$score`.

</details>
:::
