# The `summary()` Function {#sec-summary-function}

::: {.pbl-box .pbl-problem}
**The Problem**

You've now computed the mean, median, quartiles, and missing-value count of `age` across three separate pages, using `mean()`, `median()`, `quantile()`, and `sum(is.na(...))` one at a time (Level 4). Your professor wants all of these numbers at once, in one glance, without four separate function calls every time you look at a new variable.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Use `summary()` to get a complete numeric snapshot — minimum, quartiles, mean, maximum, and missing count — in a single call.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 4, what six or seven numbers did `summary(participants$age)` report?
- If `summary()` already reports the median as "Median" and the quartiles as "1st Qu." / "3rd Qu.", is there anything this level has covered so far that `summary()` *doesn't* include?
:::

## The R Concept

**`summary()`** produces the "five-number summary" (minimum, 1st quartile, median, 3rd quartile, maximum) plus the mean, all in one call — and automatically reports a count of missing values if any exist, with no `na.rm` needed:

```{r}
#| eval: false
summary(x)
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

summary(participants$age)
```

```
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's
  18.00   36.25   49.00   47.43   58.75   81.00       6
```

One call reports the minimum (18), 1st quartile (36.25), median (49), mean (47.43), 3rd quartile (58.75), maximum (81), and missing count (6) — everything from the previous three pages, condensed into a single readable line.

## Explore

Try `summary(participants)` — calling `summary()` on the **entire data frame** at once, not just one column. Notice it produces a compact summary of every column simultaneously: numeric columns get the five-number-summary treatment, while character columns (like `county`) get a simpler length/class/mode description instead.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

summary(participants$bmi)
```

```
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's
  15.60   23.35   27.70   27.12   30.20   38.60       5
```

```{r}
#| eval: false
summary(participants$systolic_bp)
```

```
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's
  90.00  116.75  128.50  129.84  145.25  179.00       4
```

Both variables' complete descriptive profiles — spread, center, and missingness all at once — ready to drop directly into your professor's report without needing four separate function calls per variable.

## Interpret the Result

`summary()` confirms what the earlier pages found piece by piece: BMI's mean (27.12) and median (27.70) sit close together, suggesting a roughly symmetric distribution, while systolic blood pressure's mean (129.84) and median (128.50) are similarly close. Both variables report a handful of missing values (5 and 4 respectively) automatically — no separate `sum(is.na(...))` call needed.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `summary()` needs no `na.rm` argument.** Unlike `mean()`, `median()`, `var()`, `sd()`, and `quantile()`, `summary()` handles missing values automatically, reporting them in a dedicated `NA's` row rather than requiring `na.rm = TRUE`.
- **Applying `summary()` to a character column and expecting numeric output.** `summary(participants$county)` reports length, class, and mode (the *storage type* trap from Page 1) rather than quartiles — `summary()` behaves differently depending on the column's type.
- **Treating `summary()`'s output as a data frame you can immediately manipulate.** `summary()` returns a special `table`-like object formatted for reading, not something you'd typically pipe into `filter()` or `mutate()` — for further computation, go back to the individual functions (`mean()`, `quantile()`, etc.).
:::

## Practice Problem

Run `summary(participants)` on the whole data frame and identify which columns show an `NA's` count, and which don't.

## Challenge Problem

Compare `summary(participants$age)`'s reported mean to the value from `mean(participants$age, na.rm = TRUE)` computed on Page 1. Confirm they match exactly, then explain in words why `summary()` doesn't need an explicit `na.rm` argument to produce this correct result.

## Key Takeaways

- `summary(x)` reports the minimum, quartiles, mean, maximum, and missing-value count for a numeric column in one call.
- `summary()` automatically excludes missing values from its calculations and reports their count separately — no `na.rm` argument required.
- `summary()` behaves differently for character or factor columns than for numeric ones, and can also summarize an entire data frame at once.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Which of these does `summary()` NOT report for a numeric column?
   a) Mean
   b) Standard deviation
   c) Median
   d) Maximum

2. **True or false.** `summary()` requires an explicit `na.rm = TRUE` argument to handle missing values correctly.

3. **Predict the output.** If a column has no missing values at all, would `summary()` still show an `NA's` row?

4. **Short answer.** Why might you still want to compute `sd()` separately, even after running `summary()`?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Standard deviation is not part of `summary()`'s output — only the five-number summary plus the mean.
2. **False.** `summary()` handles missing values automatically, reporting their count in a dedicated row without any `na.rm` argument needed.
3. No — the `NA's` row only appears when there is at least one missing value to report.
4. `summary()` doesn't report standard deviation at all, even though it's one of the most commonly cited spread measures alongside the mean — `sd()` must still be called separately to get it.

</details>
:::
