# Cross-Tabulations {#sec-cross-tabulations}

::: {.pbl-box .pbl-problem}
**The Problem**

You know 25% of all participants smoke (previous page) — but your professor's actual question is more specific: "Does the smoking rate differ between men and women?" A single overall percentage can't answer this; you need to look at smoking status *separately within each sex* — exactly the kind of two-variable breakdown `table()` alone (with one argument) can't produce.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Build a two-way frequency table — a cross-tabulation — showing how two categorical variables relate to each other.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 5, `count(participants, sex, smoker)` produced counts for every sex-and-smoking-status combination as separate *rows*. Could `table()` produce something similar, but shaped like a grid instead?
- If you wanted to know "what percentage of *women* smoke" specifically, would you divide by the total number of participants, or just the total number of women?
:::

## The R Concept

**`table()` with two arguments** builds a cross-tabulation — a grid with one variable's categories as rows and the other's as columns:

```{r}
#| eval: false
table(row_variable, column_variable)
```

```{r}
#| eval: false
table(participants$sex, participants$smoker)
```

To convert this into row-wise percentages (e.g., "of all women, what percent smoke?"), use `prop.table()` with `margin = 1` (percentages *within each row*):

```{r}
#| eval: false
prop.table(table(participants$sex, participants$smoker), margin = 1)
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sex_smoker_table <- table(participants$sex, participants$smoker)
sex_smoker_table
```

```
        No Yes
Female  42  10
Male    33  15
```

Now you can see, at a glance, that among the 52 women, 42 don't smoke and 10 do — and among the 48 men, 33 don't smoke and 15 do.

```{r}
#| eval: false
round(prop.table(sex_smoker_table, margin = 1) * 100, 1)
```

```
          No  Yes
Female  80.8 19.2
Male    68.8 31.2
```

`margin = 1` computes each row's percentages *within that row* — 80.8% and 19.2% for women sum to 100%, and 68.8% and 31.2% for men sum to 100% separately, rather than all four percentages summing to 100% together.

## Explore

Try `prop.table(sex_smoker_table)` with **no** `margin` argument at all. Notice all four percentages now sum to 100% *overall*, answering a different question ("what percent of the entire study is female non-smokers?") than the row-wise version above ("what percent of women don't smoke?").

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sex_smoker_table <- table(participants$sex, participants$smoker)
sex_smoker_pct <- round(prop.table(sex_smoker_table, margin = 1) * 100, 1)

sex_smoker_table
```

```
        No Yes
Female  42  10
Male    33  15
```

```{r}
#| eval: false
sex_smoker_pct
```

```
          No  Yes
Female  80.8 19.2
Male    68.8 31.2
```

Yes — the smoking rate does differ meaningfully by sex in this sample: 31.2% of men smoke, compared to 19.2% of women, a 12-percentage-point difference directly answering your professor's question.

## Interpret the Result

Reporting the overall 25% smoking rate (previous page) alone would have hidden this sex difference entirely — the cross-tabulation reveals that men in this sample smoke at nearly double the rate of women. As always, this is a descriptive pattern in n=48 men and n=52 women, not yet a tested statistical claim (Level 11 covers how to test whether a difference like this is likely to reflect a real, generalizable pattern versus chance).

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting to specify `margin` and misreading the resulting percentages.** Without `margin = 1` or `margin = 2`, `prop.table()` computes percentages relative to the *entire table*, not within each row or column — a very different (and often not what you intended) question.
- **Confusing `margin = 1` (row percentages) with `margin = 2` (column percentages).** `margin = 1` makes each *row* sum to 100%; `margin = 2` makes each *column* sum to 100% instead — mixing these up flips which variable you're computing percentages "within."
- **Drawing a causal conclusion from a descriptive cross-tabulation.** "Men smoke more than women in this sample" is a valid descriptive finding — it is not, by itself, evidence that being male *causes* higher smoking rates; cross-tabulations describe patterns, they don't establish causation.
:::

## Practice Problem

Build a cross-tabulation of `county` and `smoker`, then compute column percentages (`margin = 2`) to answer: "of all current smokers, what percentage live in each county?"

## Challenge Problem

Use `table(participants$sex, participants$smoker)` together with `margin = 2` (instead of `margin = 1`) to compute: "of all smokers, what percentage are male versus female?" Explain in words why this answers a different question than the row-percentage version already computed above.

## Key Takeaways

- `table(var1, var2)` builds a two-way cross-tabulation, showing how two categorical variables relate.
- `prop.table(cross_tab, margin = 1)` computes row-wise percentages; `margin = 2` computes column-wise percentages; omitting `margin` computes percentages relative to the whole table.
- A cross-tabulation can reveal patterns (like a sex difference in smoking rate) that an overall single-variable summary completely hides — but it remains a descriptive finding, not a causal claim.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `margin = 1` mean in `prop.table(tab, margin = 1)`?
   a) Percentages computed within each column
   b) Percentages computed within each row
   c) Percentages computed relative to the whole table
   d) It has no effect

2. **True or false.** A cross-tabulation can reveal patterns that an overall single-variable frequency table would hide.

3. **Predict the output.** In `sex_smoker_table`, do the row percentages for "Female" (`80.8`, `19.2`) sum to 100?

4. **Short answer.** Why is "men smoke more than women in this sample" a safer statement than "being male causes higher smoking rates," based on this cross-tabulation alone?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Percentages computed within each row.
2. **True.** The overall 25% smoking rate hid a substantial sex difference (19.2% vs. 31.2%) that only the cross-tabulation revealed.
3. Yes — `80.8 + 19.2 = 100`, exactly as expected for row percentages.
4. A cross-tabulation only describes an association observed in this particular sample — it doesn't rule out other explanations (like differences in age, county, or other unmeasured factors between the male and female participants) that a purely descriptive table can't account for.

</details>
:::
