# Level 7 Capstone: A Descriptive Statistics Report {#sec-descriptive-stats-capstone}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor's final request for this level: write the opening paragraph of a study report — the "Table 1" every published health study begins with, describing the sample before any deeper analysis. It needs to state, clearly and correctly: how many participants, their age and BMI (center and spread), the sex and smoking breakdown, and how blood pressure varies by county — pulling together every tool from this entire level.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Assemble a complete descriptive statistics report from `participants.csv`, combining central tendency, spread, frequency tables, and grouped summaries into one coherent set of findings.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Looking back over this level's seven pages, which functions would you need to describe a single *numeric* variable like `age` completely (center, spread, and missingness)?
- Which functions would you need to describe a single *categorical* variable like `sex` or `smoker` completely (counts and percentages)?
:::

## The R Concept

A complete descriptive report typically combines, for each numeric variable: `mean()`, `median()`, `sd()`, `range()`, and a missing-value count; and for each categorical variable: `table()` and `prop.table()`. `dplyr`'s `group_by()` + `summarise()` (Level 5) extends this to grouped numeric summaries, like blood pressure by county.

```{r}
#| eval: false
numeric_summary <- function(x) {
  c(
    n_valid = sum(!is.na(x)),
    n_missing = sum(is.na(x)),
    mean = round(mean(x, na.rm = TRUE), 2),
    sd = round(sd(x, na.rm = TRUE), 2),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE)
  )
}
```

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

nrow(participants)
```

```
[1] 100
```

```{r}
#| eval: false
round(mean(participants$age, na.rm = TRUE), 2)
```

```
[1] 47.43
```

```{r}
#| eval: false
round(sd(participants$age, na.rm = TRUE), 2)
```

```
[1] 15.03
```

```{r}
#| eval: false
round(prop.table(table(participants$sex)) * 100, 1)
```

```
Female   Male
  52.0   48.0
```

## Explore

Try wrapping the numeric-summary calculations above into a small reusable function (a Level 9 preview), taking a column as input and returning a named vector of statistics — so you don't have to retype `mean()`, `sd()`, `median()`, `min()`, and `max()` separately for every variable in the report.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

# Sample size
n_total <- nrow(participants)

# Age: center, spread, missingness
age_mean <- round(mean(participants$age, na.rm = TRUE), 2)
age_sd <- round(sd(participants$age, na.rm = TRUE), 2)
age_median <- median(participants$age, na.rm = TRUE)
age_missing <- sum(is.na(participants$age))

# BMI: center and spread
bmi_mean <- round(mean(participants$bmi, na.rm = TRUE), 2)
bmi_sd <- round(sd(participants$bmi, na.rm = TRUE), 2)

# Sex breakdown
sex_pct <- round(prop.table(table(participants$sex)) * 100, 1)

# Smoking status
smoker_pct <- round(prop.table(table(participants$smoker)) * 100, 1)

# Blood pressure by county
bp_by_county <- participants |>
  group_by(county) |>
  summarise(n = n(), mean_sbp = round(mean(systolic_bp, na.rm = TRUE), 2)) |>
  arrange(desc(mean_sbp))

n_total
```

```
[1] 100
```

```{r}
#| eval: false
age_mean
```

```
[1] 47.43
```

```{r}
#| eval: false
sex_pct
```

```
Female   Male
  52.0   48.0
```

```{r}
#| eval: false
bp_by_county
```

```
# A tibble: 5 x 3
  county          n mean_sbp
  <chr>       <int>    <dbl>
1 Millbrook      21   133.30
2 Springvale     26   131.46
3 Oakford        17   129.71
4 Cedar Falls    20   128.16
5 Riverdale      16   124.36
```

Every number your professor's Table 1 paragraph needs is now computed: sample size, age center and spread, sex breakdown, smoking rate, and blood pressure ranked by county.

## Interpret the Result

Putting it into a written paragraph:

*"The study sample included 100 participants, with a mean age of 47.4 years (SD = 15.0, 6 missing) and mean BMI of 27.1 (SD = 4.6). The sample was 52% female and 48% male, and 25% of participants reported current smoking. Mean systolic blood pressure ranged from 124.4 mmHg in Riverdale to 133.3 mmHg in Millbrook across the study's five counties."*

Every number in that paragraph traces directly back to a specific function call from this level — nothing was estimated or eyeballed. This is exactly the kind of opening paragraph a published health study begins with, built entirely from tools covered across eight pages.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Reporting a mean without its missing-value count.** "Mean age was 47.4" is incomplete without also noting that 6 of 100 ages were missing — a reader needs to know how much of the sample that mean actually reflects.
- **Rounding at different points inconsistently.** Rounding `age_mean` to 2 decimals but leaving `bp_by_county`'s means unrounded (or rounded differently) makes a report look inconsistent — decide on a rounding convention and apply it uniformly.
- **Writing the narrative paragraph before double-checking every number against its source calculation.** It's easy to transpose two percentages, or misremember which county had the highest blood pressure, when translating numbers into prose — always re-verify each stated number against the code that produced it before finalizing a report.
:::

## Practice Problem

Extend the report to include the smoking percentage by sex (from Level 7's cross-tabulation page) as an additional sentence in the paragraph.

## Challenge Problem

Write a complete "Table 1" for just the subset of participants aged 65 and older (Level 5's `filter()`), reporting sample size, mean age, mean BMI, sex breakdown, and smoking rate for this older subgroup specifically. How does this subgroup's profile compare to the full sample's?

## Key Takeaways

- A complete descriptive report combines central tendency, spread, and missingness for numeric variables, and counts with percentages for categorical variables.
- `group_by()` + `summarise()` (Level 5) extends single-variable descriptive statistics into grouped comparisons, like blood pressure by county.
- Every number in a written report should be traceable back to a specific, reproducible calculation — never eyeballed or estimated.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Which piece of information should always accompany a reported mean for a variable with missing data?
   a) The variable's storage type
   b) The count (or percentage) of missing values
   c) The variable's alphabetical position in the data frame
   d) Nothing extra is needed

2. **True or false.** A complete descriptive report only needs numeric summaries; categorical variables don't need to be described.

3. **Predict the output.** Given the reported percentages (52% female, 48% male), would you expect these two numbers to sum to 100?

4. **Short answer.** Why is it good practice to re-verify every number in a written report paragraph against its source calculation before finalizing it?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The count or percentage of missing values, so a reader knows how much of the sample the reported mean actually reflects.
2. **False.** Categorical variables (like `sex` and `smoker`) need their own counts and percentages — a complete report describes every relevant variable, numeric and categorical alike.
3. Yes — since every participant has exactly one recorded sex in this dataset, the two percentages should sum to 100.
4. Numbers are easy to transpose, mis-round, or misremember when translating from code output into prose — a final verification pass catches these errors before they reach a reader who may take the report at face value.

</details>
:::
