# `arrange()` {#sec-arrange}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor wants to see the five oldest participants in the study first, to spot-check whether the oldest enrollees have unusually high blood pressure. The dataset isn't sorted by age at all — it's in enrollment order — so you need to reorder the entire dataset by age before you can even look at the top of it.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Sort a data frame's rows by one or more columns, in ascending or descending order.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If you wanted to sort a plain vector of numbers, what base R function (from earlier levels) might you reach for?
- Sorting a whole *data frame* by one column needs to move every other column's values along with it, row by row — is that different from sorting a single, standalone vector?
:::

## The R Concept

**`arrange()`** sorts a data frame's rows by one or more columns, in ascending order by default:

```{r}
#| eval: false
arrange(df, column)          # ascending (smallest first)
arrange(df, desc(column))    # descending (largest first)
```

```{r}
#| eval: false
arrange(participants, age)          # youngest first
arrange(participants, desc(age))    # oldest first
```

Unlike sorting a plain vector (where only the values themselves move), `arrange()` reorders *entire rows* together — every column's value for a given participant stays correctly matched to that same participant, just in a new row position.

You can also sort by multiple columns; ties in the first column are broken by the second:

```{r}
#| eval: false
arrange(participants, county, desc(age))   # by county, then oldest-first within each county
```

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

youngest_first <- arrange(participants, age)
head(youngest_first, 3)
```

```
  id age    sex     county systolic_bp diastolic_bp  bmi smoker
1  6 18 Female  Riverdale         117           64 24.0     No
2 33 18 Female    Oakford         149           94 30.3    Yes
3 36 18   Male Springvale         152           88 23.1     No
```

```{r}
#| eval: false
oldest_first <- arrange(participants, desc(age))
head(oldest_first, 3)
```

```
  id age    sex      county systolic_bp diastolic_bp  bmi smoker
1 62 81   Male   Millbrook         153           95 30.3    Yes
2 19 80 Female Cedar Falls         109           79 28.4     No
3 17 79 Female  Springvale         117           73 21.1    Yes
```

## Explore

Try `arrange(participants, county, desc(age))` — sorting first by county (alphabetically), then by age (oldest first) *within* each county. Compare this to `arrange(participants, desc(age), county)` — sorting by age first, with county only breaking ties. Notice these produce genuinely different orderings, since the *first* column listed always takes priority.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

oldest_first <- arrange(participants, desc(age))

head(oldest_first, 5)
```

```
  id age    sex      county systolic_bp diastolic_bp  bmi smoker
1 62  81   Male   Millbrook         153           95 30.3    Yes
2 19  80 Female Cedar Falls         109           79 28.4     No
3 17  79 Female  Springvale         117           73 21.1    Yes
4 50  74   Male   Millbrook         114           75 33.5     No
5 31  70   Male     Oakford         145           85   NA     No
```

The five oldest participants (ages 81, 80, 79, 74, and 70) are now at the top, ready for your professor to spot-check — and notably, two of the five oldest participants (ages 81 and 70) do have elevated systolic blood pressure, echoing the pattern the previous page's `filter()` began to surface.

## Interpret the Result

`arrange(participants, desc(age))` reordered every row by age, oldest first, while keeping each participant's full record intact and correctly matched — participant #62 (Millbrook, age 81) is now row 1, with every one of their other values (blood pressure, BMI, smoking status) correctly still attached to them, not shuffled independently.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `desc()` for descending order.** `arrange(df, age)` sorts smallest-first by default; without wrapping the column in `desc()`, you'll get the opposite order from what you intended for a "largest first" request.
- **Assuming `arrange()` changes the underlying data.** Like `filter()` and `select()`, `arrange()` returns a *new*, reordered data frame — the original `participants` object is completely unaffected unless you explicitly reassign the result back to it.
- **Not realizing column order matters for multi-column sorts.** `arrange(df, a, b)` and `arrange(df, b, a)` can produce different orderings — the first-listed column always takes sorting priority, with later columns only breaking ties.
:::

## Practice Problem

Use `arrange()` to sort `participants` by `bmi` in ascending order, and identify the participant with the lowest recorded BMI.

## Challenge Problem

Sort `participants` by `county` (alphabetical, ascending) and then by `systolic_bp` (descending) within each county, and report the participant with the highest blood pressure in Springvale specifically.

## Key Takeaways

- `arrange(df, column)` sorts rows in ascending order by default; `arrange(df, desc(column))` sorts descending.
- Sorting a data frame reorders entire rows together, keeping every column correctly matched to its original participant.
- Multiple sort columns break ties in order — the first column listed always takes priority.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `arrange(df, desc(age))` do?
   a) Sorts by age, youngest first
   b) Sorts by age, oldest first
   c) Removes the age column
   d) Filters out missing ages

2. **True or false.** `arrange()` modifies the original data frame in place, without needing reassignment.

3. **Predict the output.** After `arrange(df, x)`, which row would have the smallest value of `x`?

4. **Write the code.** Write the code to sort `participants` by `bmi`, largest first.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Sorts by age, oldest first.
2. **False.** Like other dplyr verbs, `arrange()` returns a new data frame; the original is unchanged unless reassigned.
3. The first row.
4. `arrange(participants, desc(bmi))`

</details>
:::
