# `mutate()` {#sec-mutate}

::: {.pbl-box .pbl-problem}
**The Problem**

Back in Level 4, you built `pulse_pressure` and `bmi_category` with base R's `df$new_col <- ...` pattern, one assignment line per new variable. Your professor now wants three new variables added at once — pulse pressure, BMI category, and a hypertension flag — and typing three separate `participants$` assignment lines, one after another, feels repetitive for something that's really one coherent step: "add these derived variables to the dataset."
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Create one or more new columns in a single, unified step using `dplyr::mutate()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 4, what did `participants$pulse_pressure <- participants$systolic_bp - participants$diastolic_bp` do?
- If a single function could create several new columns at once, each specified by name, would you expect to need `participants$` repeated in front of every column reference inside it?
:::

## The R Concept

**`mutate()`** creates (or modifies) one or more columns at once, using column names directly with no `df$` prefix — and, crucially, later new columns *within the same `mutate()` call* can reference earlier ones:

```{r}
#| eval: false
mutate(df,
  new_col1 = <calculation>,
  new_col2 = <calculation using new_col1>
)
```

```{r}
#| eval: false
mutate(participants,
  pulse_pressure = systolic_bp - diastolic_bp,
  is_hypertensive = as.numeric(systolic_bp >= 130)
)
```

This mirrors Level 4's `case_when()` and `ifelse()` exactly in spirit — `mutate()` is simply the tidyverse's version of "assign a new column," the same way `filter()` was the tidyverse's version of logical row indexing.

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

participants_new <- mutate(participants,
  pulse_pressure = systolic_bp - diastolic_bp,
  is_hypertensive = as.numeric(systolic_bp >= 130)
)

names(participants_new)
```

```
[1] "id"              "age"             "sex"             "county"
[5] "systolic_bp"     "diastolic_bp"    "bmi"             "smoker"
[9] "pulse_pressure"  "is_hypertensive"
```

```{r}
#| eval: false
head(participants_new[, c("systolic_bp", "diastolic_bp", "pulse_pressure", "is_hypertensive")], 3)
```

```
  systolic_bp diastolic_bp pulse_pressure is_hypertensive
1         125           79             46               0
2         123           80             43               0
3         130           77             53               1
```

## Explore

Try referencing a just-created column within the same `mutate()` call: `mutate(participants, pulse_pressure = systolic_bp - diastolic_bp, high_pp = as.numeric(pulse_pressure > 50))`. Confirm `high_pp` correctly uses `pulse_pressure`, even though both were created in the very same call — `mutate()` processes its arguments in order, top to bottom.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

participants <- mutate(participants,
  pulse_pressure = systolic_bp - diastolic_bp,
  bmi_category = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi < 25   ~ "Normal",
    bmi < 30   ~ "Overweight",
    TRUE       ~ "Obese"
  ),
  is_hypertensive = as.numeric(systolic_bp >= 130)
)

table(participants$bmi_category, useNA = "always")
```

```
Normal Obese Overweight Underweight       <NA>
    28    25         39           3           5
```

All three derived variables from your professor's request were added in a single `mutate()` call, and the `bmi_category` counts (28 Normal, 25 Obese, 39 Overweight, 3 Underweight, 5 missing) exactly match the counts you first computed with nested `ifelse()` back in Level 2 — the same result, produced more readably.

## Interpret the Result

Three new derived variables — `pulse_pressure`, `bmi_category`, and `is_hypertensive` — now exist in `participants`, all added in one step whose logic is easy to read top to bottom, rather than three disconnected `participants$... <- ...` lines scattered through a script.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `mutate()` returns a new data frame.** Like `filter()`, `select()`, and `arrange()`, `mutate()` doesn't modify `participants` in place — you must reassign the result (`participants <- mutate(participants, ...)`) to keep the new columns.
- **Assuming column order within `mutate()` doesn't matter.** It does — a later argument can reference an earlier one *within the same call* (as in the Explore step), but not the reverse; arguments are evaluated top to bottom, in the order written.
- **Recreating Level 4's `case_when()` logic from scratch instead of reusing it.** `mutate()` doesn't replace `case_when()`, `ifelse()`, or arithmetic — it's the *container* they go inside; all of Level 4's conditional tools remain exactly as useful here.
:::

## Practice Problem

Use `mutate()` to add a `bp_ratio` column (`systolic_bp / diastolic_bp`) and an `age_group` column (`"Senior"` for 65+, `"Non-senior"` otherwise, using `ifelse()`) in a single call.

## Challenge Problem

Use `mutate()` to add a `risk_score` column that equals `1` for every hypertension criterion a participant meets, summed across being 65+, having BMI 30+, and being a current smoker (hint: this needs three separate `as.numeric()` comparisons added together within one `mutate()` expression). Report the maximum `risk_score` observed in the dataset.

## Key Takeaways

- `mutate(df, new_col = ...)` creates or modifies one or more columns at once, using bare column names with no `df$` prefix.
- Later columns within the same `mutate()` call can reference earlier ones, evaluated top to bottom.
- Like other dplyr verbs, `mutate()` returns a new data frame — reassign the result to keep the changes.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `mutate(df, x2 = x * 2)` do?
   a) Removes column `x`
   b) Creates a new column `x2` equal to `x` times 2
   c) Filters rows where `x` equals 2
   d) Renames `x` to `x2`

2. **True or false.** Within a single `mutate()` call, a later column can reference one created earlier in the same call.

3. **Predict the output.** After `df <- mutate(df, y = x + 1)`, if the original `df` had no column `y`, does `df` now have one?

4. **Write the code.** Write the `mutate()` call to add a column `bmi_rounded` equal to `round(bmi, 0)`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Creates a new column `x2` equal to `x` times 2.
2. **True.** `mutate()` evaluates its arguments in order, so later ones can reference earlier ones within the same call.
3. Yes — since the result was reassigned back to `df`, the new `y` column now exists.
4. `mutate(df, bmi_rounded = round(bmi, 0))`

</details>
:::
