# `group_by()` {#sec-group-by}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor's real question isn't "what's the average blood pressure across the whole study" — it's "how does average blood pressure differ *by county*?" `summarise()` alone collapses everything into one overall row, throwing away exactly the comparison you need. In Level 2 you might have written five separate `mean(participants$systolic_bp[participants$county == "..."], na.rm = TRUE)` calls, one per county — tedious, and easy to mistype a county name.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Split a data frame into groups and compute summary statistics separately for each group, using `dplyr::group_by()` together with `summarise()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 4, what did `tapply(participants$systolic_bp, participants$county, mean, na.rm = TRUE)` compute?
- If `summarise()` alone produces one overall row, what would you expect it to produce if the data were first split into five county-sized pieces?
:::

## The R Concept

**`group_by()`** doesn't change how the data looks when printed — it tags the data frame as "grouped," so that every subsequent `summarise()` (or `mutate()`) is computed *separately within each group* instead of across the whole dataset:

```{r}
#| eval: false
grouped <- group_by(df, grouping_column)
summarise(grouped, stat = <summary function>(column))
```

```{r}
#| eval: false
by_county <- group_by(participants, county)
summarise(by_county, mean_bp = mean(systolic_bp, na.rm = TRUE))
```

The result has **one row per group** — five rows for five counties — instead of `summarise()`'s usual single overall row.

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

by_county <- group_by(participants, county)

county_summary <- summarise(by_county,
  n_participants = n(),
  mean_bp = round(mean(systolic_bp, na.rm = TRUE), 2)
)

county_summary
```

```
# A tibble: 5 x 3
  county      n_participants mean_bp
  <chr>                <int>   <dbl>
1 Cedar Falls             20  128.16
2 Millbrook               21  133.30
3 Oakford                 17  129.71
4 Riverdale               16  124.36
5 Springvale              26  131.46
```

Five rows, one per county — `n()` inside a grouped `summarise()` counts rows *within each group*, not the whole dataset.

## Explore

Try grouping by two columns at once: `group_by(participants, county, smoker)`, then `summarise(..., n = n())`. Confirm the result has one row for every county-by-smoking-status combination that actually occurs in the data (not necessarily all 10 possible combinations, if some don't occur).

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

by_county <- group_by(participants, county)

county_summary <- summarise(by_county,
  n_participants = n(),
  mean_age = round(mean(age, na.rm = TRUE), 2),
  mean_bp  = round(mean(systolic_bp, na.rm = TRUE), 2)
)

county_summary
```

```
# A tibble: 5 x 4
  county      n_participants mean_age mean_bp
  <chr>                <int>    <dbl>   <dbl>
1 Cedar Falls             20    50.58  128.16
2 Millbrook               21    48.00  133.30
3 Oakford                 17    42.20  129.71
4 Riverdale               16    51.27  124.36
5 Springvale              26    45.40  131.46
```

Millbrook has the highest mean blood pressure (133.30) of all five counties, while Riverdale has the lowest (124.36) — exactly the county-by-county comparison your professor asked for, computed in two `group_by()` + `summarise()` lines instead of five repeated subset-and-mean calls.

## Interpret the Result

Splitting `participants` by `county` and summarising within each group reveals real variation base R's single overall mean (129.84 from the previous page) completely hid: Millbrook's average blood pressure (133.30) is nearly nine points higher than Riverdale's (124.36) — the kind of pattern a single ungrouped summary can never surface.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `group_by()` alone doesn't summarise anything.** `group_by(participants, county)` just tags the data as grouped — printing it looks nearly identical to the ungrouped data frame. The grouping only takes visible effect once you call `summarise()` (or `mutate()`) afterward.
- **Leaving a data frame grouped after you're done with it.** If you `group_by()` and then later run `mutate()` for something unrelated, the mutation will unexpectedly happen *within groups*. Use `ungroup()` afterward if you need fully ungrouped data again.
- **Forgetting `na.rm = TRUE` still applies per group.** Missing values are excluded within each group's calculation, not just once overall — the rule from Level 2 still holds, just applied five times instead of once.
:::

## Practice Problem

Use `group_by()` and `summarise()` to compute the number of participants and the mean `bmi` for each `smoker` status (`"Yes"` vs `"No"`).

## Challenge Problem

Use `group_by(participants, county)` followed by `summarise()` to find which county has both the highest mean BMI *and* the highest proportion of smokers. Are they the same county?

## Key Takeaways

- `group_by(df, column)` tags a data frame as grouped, without changing its printed appearance by itself.
- Following `group_by()` with `summarise()` computes statistics separately within each group, producing one row per group.
- Always `ungroup()` (or start a fresh pipeline) once you're done with grouped operations, to avoid accidentally grouping later steps.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `group_by(df, county)` do by itself, before any `summarise()`?
   a) Immediately computes county means
   b) Tags the data frame as grouped by county, with no visible summary yet
   c) Deletes all columns except `county`
   d) Sorts the data frame by county

2. **True or false.** `summarise()` after `group_by(df, county)` produces one row per county.

3. **Predict the output.** If `participants` has 5 distinct counties, how many rows would `summarise(group_by(participants, county), n = n())` return?

4. **Find the error.**
   ```r
   grouped <- group_by(participants, county)
   grouped
   # expecting to already see county-level means here
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Tags the data frame as grouped by county, with no visible summary yet.
2. **True.** Grouped `summarise()` computes one row per distinct group value.
3. 5 rows — one per distinct county.
4. `group_by()` alone doesn't compute anything — it only sets up the grouping. A `summarise()` call must follow to actually see county-level statistics.

</details>
:::
