# `count()` {#sec-count}

::: {.pbl-box .pbl-problem}
**The Problem**

You just used `group_by()` + `summarise(n = n())` to count participants per county — three lines of code for a task as simple as "how many rows have each value of this column?" Back in Level 2, `table(participants$county)` did this in one call. Is there a tidyverse equivalent that's just as short?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Count how many rows fall into each category of one or more columns, using `dplyr::count()` — the one-line shortcut for `group_by()` + `summarise(n = n())`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 2, what did `table(participants$county)` return?
- If `count()` is a shortcut for `group_by()` + `summarise(n = n())`, would you expect its output to look like `table()`'s output, or like a `summarise()` data frame's output?
:::

## The R Concept

**`count()`** counts rows per group in one step — no `group_by()` needed first:

```{r}
#| eval: false
count(df, column)
```

```{r}
#| eval: false
count(participants, county)
```

This is exactly equivalent to `summarise(group_by(participants, county), n = n())` from the previous page, but in a single readable call. You can also count combinations of multiple columns at once, and sort by frequency with `sort = TRUE`:

```{r}
#| eval: false
count(participants, county, smoker)          # every county-by-smoker combination
count(participants, county, sort = TRUE)     # counts ordered largest first
```

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

count(participants, county)
```

```
       county  n
1 Cedar Falls 20
2   Millbrook 21
3     Oakford 17
4   Riverdale 16
5  Springvale 26
```

Note the column is automatically named `n` — matching exactly what `table(participants$county)` reported back in Level 2, just returned as a proper data frame instead of a table object.

```{r}
#| eval: false
count(participants, county, sort = TRUE)
```

```
       county  n
1  Springvale 26
2   Millbrook 21
3 Cedar Falls 20
4     Oakford 17
5   Riverdale 16
```

`sort = TRUE` reorders the rows largest-count-first, instantly surfacing that Springvale is the largest county in the study.

## Explore

Try `count(participants, smoker)` and confirm it matches Level 2's `table(participants$smoker)` result (75 "No", 25 "Yes"). Then try `count(participants, county, smoker)` — notice the result has one row for every county-and-smoking-status combination that actually appears, not all 10 mathematically possible combinations.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

county_smoker_counts <- count(participants, county, smoker)

county_smoker_counts
```

```
        county smoker  n
1  Cedar Falls     No 17
2  Cedar Falls    Yes  3
3    Millbrook     No 15
4    Millbrook    Yes  6
5      Oakford     No 11
6      Oakford    Yes  6
7    Riverdale     No 12
8    Riverdale    Yes  4
9   Springvale     No 20
10  Springvale    Yes  6
```

All 10 county-by-smoking-status combinations appear (every county has at least one smoker and one non-smoker), giving your professor exactly the cross-tabulation `table()` alone (Level 2) could only produce for a single column at a time without extra work.

## Interpret the Result

`county_smoker_counts` breaks the earlier single `smoker` count (75 No, 25 Yes) down further by county — for example, Oakford has the highest proportion of smokers relative to its size (6 of 17, roughly 35%), while Cedar Falls has the lowest (3 of 20, 15%).

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Trying to `count()` a column that doesn't exist or is misspelled.** Like every dplyr verb, `count()` uses bare column names — a typo (`count(participants, contry)`) produces a "column not found" error, not a silent empty result.
- **Forgetting `count()` already includes an implicit `group_by()`.** You don't need to call `group_by()` first — `count(participants, county)` handles the grouping and summarising in one step internally.
- **Confusing `count()`'s output column name.** The count column is always named `n` by default (not `count` or `freq`) — referencing `county_smoker_counts$count` will fail; it must be `county_smoker_counts$n`.
:::

## Practice Problem

Use `count()` to find how many participants fall into each combination of `sex` and `smoker`. Which combination is least common?

## Challenge Problem

Use `count(participants, county, sort = TRUE)` to confirm Springvale is the largest county, then use `count(participants, county, smoker)` to determine whether Springvale also has the largest *number* of smokers (not necessarily the same as the largest proportion).

## Key Takeaways

- `count(df, column)` is a one-line shortcut for `group_by()` + `summarise(n = n())`, with the result column always named `n`.
- `count()` can take multiple columns to count every combination that actually occurs in the data.
- `sort = TRUE` reorders the result largest-count-first, useful for quickly spotting the most (or least) common categories.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What is `count(df, column)` equivalent to?
   a) `filter(df, column)`
   b) `group_by(df, column)` + `summarise(n = n())`
   c) `arrange(df, column)`
   d) `select(df, column)`

2. **True or false.** The count column produced by `count()` is always named `n`.

3. **Predict the output.** What would `count(participants, smoker)` return, given 75 "No" and 25 "Yes" values?

4. **Write the code.** Write the `count()` call to count participants per combination of `county` and `sex`, sorted largest-first.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `group_by(df, column)` + `summarise(n = n())`.
2. **True.** `count()`'s result column is always named `n` by default.
3. A two-row data frame: `smoker = "No", n = 75` and `smoker = "Yes", n = 25`.
4. `count(participants, county, sex, sort = TRUE)`

</details>
:::
