# `distinct()` {#sec-distinct}

::: {.pbl-box .pbl-problem}
**The Problem**

Before analyzing `county` any further, your professor wants a quick sanity check: exactly how many *different* counties actually appear in the dataset, and what are their names — without caring yet how many participants are in each. `count()` from the last page tells you counts, but sometimes you just want the unique *values* themselves, nothing more.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

List the unique combinations of values in one or more columns, using `dplyr::distinct()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 3, what did `unique(participants$county)` return?
- If `count()` gives you unique values *plus* how many rows have each, what would you expect a function that gives you *only* the unique values (no counts) to look like?
:::

## The R Concept

**`distinct()`** returns the unique rows (or unique combinations of specified columns), with no count attached:

```{r}
#| eval: false
distinct(df, column)
```

```{r}
#| eval: false
distinct(participants, county)
```

This is the tidyverse counterpart to Level 3's `unique(participants$county)` — but because `distinct()` operates on a data frame rather than a single vector, it can also find unique *combinations* across several columns at once:

```{r}
#| eval: false
distinct(participants, county, sex)   # every county-sex combination that occurs
```

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

distinct(participants, county)
```

```
       county
1     Oakford
2   Riverdale
3   Millbrook
4  Springvale
5 Cedar Falls
```

Five distinct counties — the rows appear in the order they were first encountered in the dataset (row 1 is `Oakford` because participant #1 lives there), not alphabetically.

```{r}
#| eval: false
nrow(distinct(participants, county))
```

```
[1] 5
```

## Explore

Try `distinct(participants, county, sex)` and check `nrow()` on the result. With 5 counties and 2 sex categories, there are at most 10 possible combinations — confirm whether all 10 actually occur in the data, or whether some combination is missing entirely.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

unique_counties <- distinct(participants, county)

nrow(unique_counties)
```

```
[1] 5
```

```{r}
#| eval: false
unique_county_sex <- distinct(participants, county, sex)

nrow(unique_county_sex)
```

```
[1] 10
```

Five distinct counties confirmed, matching what `unique(participants$county)` reported in Level 3 and what `count(participants, county)` implied by returning 5 rows on the previous page. All 10 possible county-and-sex combinations occur at least once in the data.

## Interpret the Result

`distinct()` confirms there are exactly 5 unique counties in the study — no unexpected typos or extra category hiding in the data (recall Level 3's messy dataset, where inconsistent capitalization made "13 distinct-looking values" collapse to only 5 real counties after cleaning). This clean dataset's `county` column has no such problem.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing `distinct()` with `count()`.** `distinct()` returns only the unique values themselves, with no counts attached; `count()` returns the unique values *plus* how many rows had each. Reach for `distinct()` when you only care "what values exist," and `count()` when you also care "how many of each."
- **Expecting `distinct()` to sort its output.** Unlike `arrange()`, `distinct()` returns unique values in the order they first appear in the data, not alphabetically or numerically — combine with `arrange()` afterward if sorted output matters.
- **Forgetting `distinct()` on a whole data frame (no columns specified) checks every column.** `distinct(participants)` with no column names looks for entire duplicate *rows* across all columns — a different (and much rarer) kind of duplicate than a single repeated `county` value.
:::

## Practice Problem

Use `distinct()` to find every unique combination of `smoker` and `sex` that occurs in the dataset, and report how many there are.

## Challenge Problem

Compare `nrow(distinct(participants, county, smoker))` to the number of rows produced by last page's `count(participants, county, smoker)`. Should these two numbers be the same? Explain why or why not.

## Key Takeaways

- `distinct(df, column)` returns the unique values of a column (or unique combinations of several columns), with no counts attached.
- `distinct()` is the tidyverse counterpart to Level 3's `unique()`, extended to work across multiple columns at once.
- Rows appear in first-encountered order, not sorted — pair with `arrange()` if sorted output is needed.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** How does `distinct()` differ from `count()`?
   a) `distinct()` returns unique values with counts; `count()` doesn't
   b) `distinct()` returns unique values with no counts; `count()` returns unique values with counts
   c) They are identical in every way
   d) `distinct()` only works on numeric columns

2. **True or false.** `distinct(participants, county)` sorts the counties alphabetically.

3. **Predict the output.** If `smoker` has only two possible values ("Yes" and "No"), how many rows would `distinct(participants, smoker)` return?

4. **Write the code.** Write the code to find every unique combination of `county` and `smoker` in `participants`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `distinct()` returns unique values with no counts; `count()` returns unique values with counts.
2. **False.** `distinct()` preserves first-encountered order, not alphabetical order.
3. 2 rows — one for `"Yes"`, one for `"No"`.
4. `distinct(participants, county, smoker)`

</details>
:::
