# `rename()` {#sec-rename}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor asks you to rename `systolic_bp` and `diastolic_bp` to the clearer `sbp` and `dbp` before sharing the dataset with a collaborator. In Level 4's base R, you did this with `names(df)[names(df) == "old"] <- "new"` — correct, but you had to look up the column's *position* indirectly through `names()` matching, rather than just naming old and new directly.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Rename one or more columns directly, using `dplyr::rename()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 4, what did `names(participants)[names(participants) == "systolic_bp"] <- "sbp"` actually do, step by step?
- If a function let you just write "new name equals old name" directly, would you still need to go through `names()` matching at all?
:::

## The R Concept

**`rename()`** renames columns with a direct `new_name = old_name` mapping — no `names()` indirection required:

```{r}
#| eval: false
rename(df, new_name = old_name)
```

```{r}
#| eval: false
rename(participants, sbp = systolic_bp, dbp = diastolic_bp)
```

Read the argument order carefully: it's **new name first, old name second** — the reverse of how you might instinctively say it aloud ("rename systolic_bp to sbp"), but consistent with every other tidyverse verb's `new_column = <expression>` pattern (`mutate()`, `summarise()`).

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

renamed <- rename(participants, sbp = systolic_bp, dbp = diastolic_bp)

names(renamed)
```

```
[1] "id"      "age"     "sex"     "county"  "sbp"     "dbp"     "bmi"
[8] "smoker"
```

`systolic_bp` and `diastolic_bp` are gone from the column names; `sbp` and `dbp` have taken their place, in the same column positions.

## Explore

Try `rename(participants, participant_id = id)` and confirm with `names()` that only `id` changed, while every other column name — including its position in the data frame — stayed exactly the same. `rename()` never reorders columns, unlike `select()`.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

for_collaborator <- rename(participants,
  sbp = systolic_bp,
  dbp = diastolic_bp
)

names(for_collaborator)
```

```
[1] "id"      "age"     "sex"     "county"  "sbp"     "dbp"     "bmi"
[8] "smoker"
```

```{r}
#| eval: false
head(for_collaborator[, c("sbp", "dbp")], 3)
```

```
  sbp dbp
1 125  79
2 123  80
3 130  77
```

The dataset is ready to share with your professor's collaborator, using the clearer `sbp`/`dbp` names — with every value untouched, only the column labels changed.

## Interpret the Result

`for_collaborator` has identical data to `participants`, just with two columns relabeled. This is purely cosmetic — no values changed, no rows were reordered, and `nrow()`/`ncol()` are identical to the original.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Reversing the argument order.** `rename(participants, systolic_bp = sbp)` would try to rename a column called `sbp` (which doesn't exist yet) to `systolic_bp` — the *opposite* of the intended renaming. Always read it as "new = old."
- **Expecting `rename()` to reorder columns.** Unlike `select()`, `rename()` never changes column order or drops any columns — it only relabels the ones you specify, leaving everything else (including position) untouched.
- **Forgetting to reassign the result.** Like every other dplyr verb covered so far, `rename()` returns a new data frame; `participants` itself is unaffected unless you write `participants <- rename(participants, ...)`.
:::

## Practice Problem

Use `rename()` to rename `bmi` to `body_mass_index` and `county` to `region`, and confirm with `names()` that both changes took effect while all other column names and their order stayed the same.

## Challenge Problem

Compare Level 4's base R approach (`names(df)[names(df) == "old"] <- "new"`) against `rename()` for renaming three columns at once. Which approach scales better as the number of renamed columns grows, and why?

## Key Takeaways

- `rename(df, new_name = old_name)` relabels a column directly — argument order is new name first, old name second.
- `rename()` never reorders or drops columns; it only changes labels, leaving positions and all other columns untouched.
- Like other dplyr verbs, `rename()` returns a new data frame and must be reassigned to keep the change.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In `rename(df, new_name = old_name)`, which comes first?
   a) The old name
   b) The new name
   c) Order doesn't matter
   d) Neither — both must be quoted strings in a vector

2. **True or false.** `rename()` can change the order of columns in a data frame.

3. **Predict the output.** After `df2 <- rename(df, patient_id = id)`, does `df` still have a column called `id`?

4. **Find the error.**
   ```r
   rename(participants, systolic_bp = sbp)
   # intended to rename systolic_bp to sbp
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The new name comes first, then `=`, then the old name.
2. **False.** `rename()` only changes column labels — it never reorders or drops columns.
3. Yes — `rename()` returns a new data frame (`df2`); the original `df` is completely unaffected.
4. The argument order is reversed. Since `participants` has a column named `systolic_bp` (not `sbp`), this call tries to rename a nonexistent `sbp` column. It should be `rename(participants, sbp = systolic_bp)`.

</details>
:::
