# The Native Pipe `|>` {#sec-native-pipe}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor asks you to filter to seniors (65+), then look at just the first 3 rows. Nesting the calls works — `head(filter(participants, age >= 65), 3)` — but notice you have to read it **inside-out**: R evaluates `filter()` first, even though it's written second, buried inside `head()`'s parentheses. As soon as you chain three or four steps together, this inside-out reading becomes genuinely hard to follow.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Chain multiple operations together in the order you actually think about them, using R's native pipe operator `|>`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- In `head(filter(participants, age >= 65), 3)`, which function runs *first* — the one written first, or the one written innermost?
- If you could rewrite that same operation top-to-bottom, left-to-right, in the exact order the steps actually happen, what would that look like?
:::

## The R Concept

The **native pipe**, `|>` (built into base R since version 4.1, no package required), takes whatever is on its left and inserts it as the **first argument** of the function on its right:

```{r}
#| eval: false
x |> f()
# is exactly equivalent to:
f(x)
```

```{r}
#| eval: false
participants |> filter(age >= 65)
# is exactly equivalent to:
filter(participants, age >= 65)
```

Chained together, a pipe reads top-to-bottom in the exact order the steps happen — no more inside-out nesting:

```{r}
#| eval: false
participants |> filter(age >= 65) |> head(3)
```

This is identical to `head(filter(participants, age >= 65), 3)` — same result, same functions, same arguments — but now written in the order you'd describe it aloud: "take `participants`, filter to age 65+, then take the first 3."

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

# Nested version - reads inside-out
nested_result <- head(filter(participants, age >= 65), 3)

# Piped version - reads top-to-bottom
piped_result <- participants |> filter(age >= 65) |> head(3)

identical(nested_result, piped_result)
```

```
[1] TRUE
```

```{r}
#| eval: false
piped_result
```

```
  id age    sex      county systolic_bp diastolic_bp  bmi smoker
1 17  79 Female  Springvale         117           73 21.1    Yes
2 19  80 Female Cedar Falls         109           79 28.4     No
3 31  70   Male     Oakford         145           85   NA     No
```

## Explore

Try adding a third step to the pipeline: `participants |> filter(age >= 65) |> arrange(desc(systolic_bp)) |> head(3)`. Read it aloud as you look at it — "take `participants`, filter to seniors, sort by blood pressure descending, take the top 3" — and notice the code reads in that exact same order.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

at_risk_seniors <- participants |>
  filter(age >= 65 & systolic_bp >= 130) |>
  arrange(desc(systolic_bp))

at_risk_seniors
```

```
  id age    sex      county systolic_bp diastolic_bp  bmi smoker
1 62  81   Male   Millbrook         153           95 30.3    Yes
2 31  70   Male     Oakford         145           85   NA     No
3 84  67 Female   Riverdale         134           79 26.8     No
4 51  66   Male Cedar Falls         133           77 34.5     No
```

Filtering and sorting are now two clearly ordered steps in one pipeline — read top to bottom, this code says exactly what it does: "take `participants`, filter to at-risk seniors, sort by blood pressure with the highest first."

## Interpret the Result

The same 4 at-risk seniors identified back on the `filter()` page are now sorted by systolic blood pressure, highest first — participant #62 (age 81, BP 153) stands out as the most urgent case, followed by #31, #84, and #51. The pipe made the two-step process ("filter, then sort") visibly sequential rather than nested inside one another.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing the native pipe `|>` with the older magrittr pipe `%>%`.** You may see `%>%` in older tidyverse tutorials and Stack Overflow answers — it behaves almost identically, but `|>` is built into base R itself (no `library()` call needed) and is now the standard choice for new code.
- **Forgetting the pipe inserts as the *first* argument only.** `participants |> filter(age >= 65)` becomes `filter(participants, age >= 65)` — `participants` always lands in the first argument slot. If a function's data argument isn't first, the plain pipe can't simply slot it in.
- **Piping into something that isn't a function call.** `participants |> age` is an error — the right-hand side of `|>` must be a function call like `filter(...)` or `head(...)`, not a bare column reference.
:::

## Practice Problem

Rewrite `arrange(select(participants, id, age, bmi), desc(bmi))` as a piped chain using `|>`, and confirm with `identical()` that both versions produce the same result.

## Challenge Problem

Write a three-step pipeline that starts from `participants`, filters to non-smokers (`smoker == "No"`), sorts by `age` ascending, and keeps only the `id`, `age`, and `county` columns — using `|>` to chain `filter()`, `arrange()`, and `select()` in that order.

## Key Takeaways

- `x |> f()` is exactly equivalent to `f(x)` — the pipe inserts its left-hand side as the function's first argument.
- Chained pipes read top-to-bottom, in the actual order operations happen — the opposite of nested function calls, which read inside-out.
- The native pipe `|>` is built into base R (since R 4.1) and requires no package; you may also encounter the older, package-based `%>%` pipe in existing code.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What is `participants |> filter(age >= 65)` equivalent to?
   a) `filter(age >= 65, participants)`
   b) `filter(participants, age >= 65)`
   c) `age >= 65 |> filter(participants)`
   d) None of these — the pipe changes the result

2. **True or false.** The native pipe `|>` requires `library(magrittr)` or `library(dplyr)` to work.

3. **Predict the output.** What would `participants |> nrow()` return?

4. **Find the error.**
   ```r
   participants |> age
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `filter(participants, age >= 65)` — the pipe's left side becomes the first argument.
2. **False.** The native pipe `|>` is built into base R (since version 4.1) and needs no package.
3. `100` — the number of rows in `participants`, since `nrow()` takes a data frame and `participants` is piped in as that argument.
4. The right-hand side of `|>` must be a function call, not a bare column reference. `age` alone isn't a function call, so this errors; you'd need something like `participants |> pull(age)` (or simply `participants$age`) instead.

</details>
:::
