# Chaining Operations {#sec-chaining-operations}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor's final request for this level: "Among participants aged 50 and older, which county has the highest average blood pressure?" This single question actually requires *four* steps — import the data, filter to age 50+, group by county, and compute + sort mean blood pressure. Doing this in base R (Level 4) would mean four or five separate intermediate objects (`filtered <- ...`, `grouped_means <- tapply(...)`, `sorted <- sort(...)`) cluttering your environment. This is exactly the moment every tool from this level comes together.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Combine `filter()`, `group_by()`, `summarise()`, and `arrange()` into a single chained pipeline using `|>`, answering a real research question in one readable block of code.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- In what order would you need to perform these four steps: sort by blood pressure, group by county, filter to age 50+, compute the mean? (Hint: you can't group data that hasn't been filtered yet, and you can't sort a summary that hasn't been computed yet.)
- Once you know the right order, could you write all four steps as one pipeline using `|>`, the way you did on the previous page?
:::

## The R Concept

A full analytical pipeline chains every relevant verb together, each one passing its result to the next:

```{r}
#| eval: false
result <- df |>
  filter(condition) |>
  group_by(grouping_column) |>
  summarise(stat = <function>(column, na.rm = TRUE)) |>
  arrange(desc(stat))
```

Each step in this chain does exactly one job — filter narrows the rows, `group_by()` sets up per-group processing, `summarise()` collapses each group to one row, and `arrange()` orders the final result — and the pipe (`|>`) makes the whole sequence read top-to-bottom, in the order the steps actually happen.

## Demonstration

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

older_participants_summary <- participants |>
  filter(age >= 50) |>
  group_by(county) |>
  summarise(
    n = n(),
    mean_bp = round(mean(systolic_bp, na.rm = TRUE), 2)
  ) |>
  arrange(desc(mean_bp))

older_participants_summary
```

```
# A tibble: 5 x 3
  county          n mean_bp
  <chr>       <int>   <dbl>
1 Millbrook       9  130.38
2 Oakford         3  130.00
3 Springvale     13  125.77
4 Cedar Falls    11  125.40
5 Riverdale       9  122.00
```

One pipeline, four chained steps, answering the exact question asked — no intermediate objects needed along the way.

## Explore

Try removing the final `arrange(desc(mean_bp))` step and re-running the pipeline. Notice the same five rows appear, just in whatever order `group_by()` happened to encounter the counties — confirming that `arrange()` only affects *display order*, not which rows or values are present.

## Solve the Problem

```{r}
#| eval: false
library(dplyr)

participants <- read.csv("data/participants.csv")

bp_by_county_50plus <- participants |>
  filter(age >= 50) |>
  group_by(county) |>
  summarise(
    n = n(),
    mean_bp = round(mean(systolic_bp, na.rm = TRUE), 2)
  ) |>
  arrange(desc(mean_bp))

bp_by_county_50plus
```

```
# A tibble: 5 x 3
  county          n mean_bp
  <chr>       <int>   <dbl>
1 Millbrook       9  130.38
2 Oakford         3  130.00
3 Springvale     13  125.77
4 Cedar Falls    11  125.40
5 Riverdale       9  122.00
```

Among participants aged 50 and older, **Millbrook** has the highest average systolic blood pressure (130.38, across 9 participants), narrowly ahead of Oakford (130.00, but only 3 participants). Riverdale has the lowest (122.00) — your professor's exact question, answered in one pipeline.

## Interpret the Result

This single chained pipeline — filter, group, summarise, arrange — replaced what would otherwise be four or five separate base R steps with intermediate objects at each stage. Notice too that Oakford's second-place mean (130.00) rests on only 3 participants, a much shakier estimate than Millbrook's 9 — a reminder from Level 11 (statistics) that group size matters when interpreting a mean, not just the mean itself.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Grouping before filtering.** `group_by()` should come *after* `filter()` in most pipelines — you generally want to narrow the rows first, then split what remains into groups, not the reverse.
- **Forgetting `arrange()` only changes the final display order.** Placing `arrange()` in the middle of a pipeline (before `summarise()`, for instance) sorts rows that will just get collapsed a step later, having no effect on the final grouped result — `arrange()` belongs at the very end unless you have a specific reason otherwise.
- **Overlooking small group sizes when interpreting grouped summaries.** A mean based on 3 participants (Oakford) and a mean based on 9 (Millbrook) aren't equally trustworthy, even though `summarise()` reports them with the same apparent precision.
:::

## Practice Problem

Write a chained pipeline that starts from `participants`, filters to current smokers (`smoker == "Yes"`), groups by `county`, computes the mean `bmi` and count per county, and sorts by mean `bmi` descending. Which county has the highest mean BMI among smokers?

## Challenge Problem

Write a pipeline that filters to participants with `bmi >= 30` (obese, from Level 4's `case_when()` categories), groups by `county` and `smoker`, computes the count for each county-and-smoking-status combination, and sorts by count descending. Which single county-and-smoking-status combination has the most obese participants?

## Key Takeaways

- A full pipeline chains `filter()` → `group_by()` → `summarise()` → `arrange()` (in that order, generally) using `|>`, replacing multiple intermediate base R objects with one readable block.
- `arrange()` belongs at the end of a pipeline — it only affects display order and has no effect if placed before a `summarise()` that will collapse the rows anyway.
- Small group sizes can produce misleadingly precise-looking summary statistics — always check `n` alongside any grouped mean before drawing conclusions.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In a pipeline `df |> filter(...) |> group_by(...) |> summarise(...) |> arrange(...)`, what does `arrange()` do?
   a) Filters rows before grouping
   b) Changes the final display order of the summarised result
   c) Groups the data a second time
   d) Removes duplicate rows

2. **True or false.** Placing `arrange()` in the middle of a pipeline, right before `summarise()`, changes the final grouped-and-summarised result.

3. **Predict the output.** In `bp_by_county_50plus`, which county has the second-highest mean blood pressure, and how many participants does that count rely on?

4. **Find the error.**
   ```r
   participants |>
     group_by(county) |>
     filter(age >= 50) |>
     summarise(mean_bp = mean(systolic_bp, na.rm = TRUE))
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Changes the final display order of the summarised result.
2. **False.** `arrange()` before `summarise()` sorts rows that are about to be collapsed anyway — it has no effect on the final grouped result. Only `arrange()` placed *after* `summarise()` affects the final output's order.
3. Oakford, with mean blood pressure 130.00 — but that mean rests on only 3 participants, far fewer than Millbrook's 9.
4. This isn't strictly an error — `group_by()` before `filter()` still works, since `filter()` operates fine on grouped data — but it's misleading style: readers expect a pipeline to narrow rows *before* grouping them, and grouping first can silently change behavior in more complex pipelines involving group-specific filters (e.g. `filter(age == max(age))`, which behaves differently before vs. after grouping).

</details>
:::
