# Writing CSV Files {#sec-writing-csv}

::: {.pbl-box .pbl-problem}
**The Problem**

You've spent an hour cleaning and recoding the `participants` dataset — fixing capitalization, adding a BMI category column, filtering out a pilot test row. Your supervisor asks you to email the cleaned version to a colleague at another clinic, who doesn't use R at all. You need to get this data *out* of R and into a plain file they can open in Excel.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Save a data frame or tibble from R to a CSV file on disk.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If `read_csv()` reads a file *into* R, what do you imagine the mirror-image function for writing *out* of R might be called?
- Your colleague doesn't use R — what file format would be usable by literally anyone, regardless of what software they have?
:::

## The R Concept

**`write_csv()`** (from `readr`) saves a data frame or tibble to a CSV file:

```{r}
#| eval: false
write_csv(data, "path/to/output.csv")
```

Base R has an equivalent, `write.csv()`, but with one especially important difference worth knowing: base R's `write.csv()` includes row names as an extra, unnamed first column by default, unless you explicitly turn that off (`row.names = FALSE`) — a common source of a mysterious extra column when colleagues open your file. `readr`'s `write_csv()` never adds row names, which is one of several reasons this course prefers it.

```{r}
#| eval: false
write_csv(participants, "participants_cleaned.csv")       # readr - no extra column, ever
write.csv(participants, "participants_cleaned.csv", row.names = FALSE)  # base R equivalent
```

## Demonstration

```{r}
#| eval: false
library(readr)

# A small cleaned example
cleaned_sample <- data.frame(
  id = c(1, 2, 3),
  age = c(46, 53, 59),
  county = c("Oakford", "Riverdale", "Oakford")
)

write_csv(cleaned_sample, "cleaned_sample.csv")
```

```{r}
#| eval: false
# Reading it back in confirms it saved correctly
read_csv("cleaned_sample.csv")
```

```
# A tibble: 3 x 3
     id   age county
  <dbl> <dbl> <chr>
1     1    46 Oakford
2     2    53 Riverdale
3     3    59 Oakford
```

## Explore

Try the base R equivalent without `row.names = FALSE`: `write.csv(cleaned_sample, "test_with_rownames.csv")`, then open the resulting file (or read it back with `read_csv()`) and look for an unexpected, unnamed first column containing `1, 2, 3` — the row names base R added automatically. Compare that to `write_csv()`'s output, which never has this issue.

## Solve the Problem

```{r}
#| eval: false
library(readr)

participants <- read_csv("data/participants.csv")

# ... cleaning steps happen here (Level 4 covers this properly) ...

write_csv(participants, "participants_cleaned.csv")
```

The file `participants_cleaned.csv` is now saved in your working directory, ready to attach to an email — a plain CSV that opens correctly in Excel, Google Sheets, or any other spreadsheet software your colleague uses, with no R installation required on their end at all.

## Interpret the Result

`write_csv()` completing with no error message means the file was written successfully to disk. Unlike an R object living only in your current session (gone the moment you close RStudio, unless saved), a CSV file on disk persists permanently and can be opened by anyone — this is the entire point of *exporting* data, as opposed to just holding it in memory.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `row.names = FALSE` with base R's `write.csv()`.** Without it, an unwanted, unnamed row-number column gets added to the output file — a frequent source of confusion for whoever opens it next.
- **Overwriting the original raw data file by accident.** Writing your cleaned output to the *same* filename as your original import (`data/participants.csv`) permanently destroys your original raw data — always write cleaned output to a clearly different filename.
- **Forgetting the file extension.** `write_csv(data, "output")` (no `.csv`) still creates a file, just without the extension most software expects — always include `.csv` explicitly.
:::

## Practice Problem

Create a small data frame with 3 participants' `id` and `bmi_category`, and write it to a file called `bmi_summary.csv` using `write_csv()`. Then read it back in to confirm it saved correctly.

## Challenge Problem

Explain, in one or two sentences, why writing your cleaned output to a new filename (like `participants_cleaned.csv`) rather than overwriting `data/participants.csv` is a form of the same "reproducibility" principle from Level 1 — specifically, what would you lose if the original raw file were gone and only the cleaned version remained?

## Key Takeaways

- `write_csv()` (readr) saves a data frame or tibble to disk as a CSV file, without adding an unwanted row-name column.
- Base R's `write.csv()` requires `row.names = FALSE` explicitly to avoid that same extra column.
- Always write cleaned or processed output to a new filename — never overwrite your original raw data file.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `write_csv()` do differently from base R's `write.csv()` by default?
   a) It adds row names automatically
   b) It never adds row names
   c) It only works on numeric data
   d) It requires a `.txt` extension

2. **True or false.** It's safe to save your cleaned data by overwriting the original raw data file.

3. **Predict the output.** After `write_csv(df, "out.csv")` runs successfully with no error, where does the resulting file live?

4. **Write the code.** Write the code to save a tibble called `results` to a file named `results_final.csv`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** It never adds row names, unlike base R's `write.csv()` without `row.names = FALSE`.
2. **False.** Always preserve the original raw file; write cleaned output to a new filename.
3. In your current working directory, unless a different path was specified.
4. `write_csv(results, "results_final.csv")`

</details>
:::
