# Reading RDS Files {#sec-reading-rds}

::: {.pbl-box .pbl-problem}
**The Problem**

A colleague sends you `participants_cleaned.rds` instead of a CSV, saying "this already has the BMI category factor set up correctly, and I didn't want that to get lost." You open it the way you'd open a CSV, `read_csv("participants_cleaned.rds")`, and get garbled nonsense — `.rds` is not a text format at all, and your colleague's comment about the factor "not getting lost" is a clue about *why* they chose it over a CSV in the first place.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Read an `.rds` file — R's own native format for saving a single R object exactly as it is.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Level 2, what happens to a factor's carefully specified level order if it's exported to a CSV (a plain-text format with no concept of "factor") and then re-imported?
- If R had a file format specifically designed to save *any* R object — including its exact type, factor levels, and structure — would that format need to be readable by non-R software at all?
:::

## The R Concept

**`.rds`** is R's native binary format for saving a *single* R object — a vector, a data frame, a factor, a list, anything — exactly as it exists in R, including all its type information. Unlike CSV, which flattens everything to plain text (losing factor level order, for instance), `.rds` preserves an object perfectly.

```{r}
#| eval: false
my_object <- readRDS("path/to/file.rds")
```

The trade-off: `.rds` files are R-specific — a colleague using Python or Excel cannot open one directly. This makes `.rds` the right choice for *saving your own intermediate work* (to reopen later, in R), and CSV or Excel the right choice for *sharing data with people or software outside R*.

## Demonstration

```{r}
#| eval: false
# Reading a previously saved R object back in
participants_cleaned <- readRDS("data/participants_cleaned.rds")

class(participants_cleaned$bmi_category)
```

```
[1] "factor"
```

```{r}
#| eval: false
levels(participants_cleaned$bmi_category)
```

```
[1] "Underweight" "Normal"      "Overweight"  "Obese"
```

Notice the factor's carefully specified level order (from Level 2's factors page) survived perfectly — exactly as your colleague intended when choosing `.rds` over CSV.

## Explore

Compare this to what would happen with CSV: if that same `bmi_category` factor had instead been saved with `write_csv()` and reopened with `read_csv()`, the column would come back as plain character text, with the clinically meaningful order (Underweight, Normal, Overweight, Obese) completely lost — you'd need to rebuild the factor and its level order from scratch. This is precisely the loss your colleague avoided by using `.rds`.

## Solve the Problem

```{r}
#| eval: false
# .rds is not plain text - trying to read it as CSV produces garbage
# read_csv("data/participants_cleaned.rds")  # do not run - wrong tool entirely

# The correct approach:
participants_cleaned <- readRDS("data/participants_cleaned.rds")

str(participants_cleaned)
```

```
'data.frame':   100 obs. of  9 variables:
 $ id          : int  1 2 3 4 5 6 7 8 9 10 ...
 $ age         : int  46 53 59 49 61 18 62 68 33 36 ...
 $ sex         : chr  "Female" "Female" "Female" "Female" ...
 $ county      : chr  "Oakford" "Riverdale" "Oakford" "Millbrook" ...
 $ systolic_bp : int  125 123 130 117 139 117 110 100 149 NA ...
 $ diastolic_bp: int  79 80 77 72 79 64 64 53 88 86 ...
 $ bmi         : num  35.1 27.7 28.3 30.4 21.5 24 19.7 29.7 22.4 NA ...
 $ smoker      : chr  "No" "No" "No" "No" ...
 $ bmi_category: Factor w/ 4 levels "Underweight",..: 4 3 3 4 2 2 2 3 2 NA ...
```

## Interpret the Result

`readRDS()` correctly restores `participants_cleaned` exactly as your colleague saved it, including `bmi_category` as a proper 4-level ordered factor — no reconstruction needed, unlike what would have happened if the same data had round-tripped through a CSV file instead.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Trying to open `.rds` files with CSV or Excel tools.** `.rds` is a binary R-specific format — `read_csv()`, `read_excel()`, and non-R software cannot open it at all.
- **Assuming `.rds` is a good format for sharing data outside R.** It's the opposite: `.rds` is ideal for your own R-to-R workflow, but useless for a colleague using Python, Stata, or Excel — use CSV or Excel for external sharing (as covered earlier this level).
- **Forgetting `.rds` stores exactly one object.** Unlike a CSV file (naturally one table) or an Excel workbook (multiple sheets), a single `.rds` file holds exactly one R object — if you need to save several related objects together, a list (Level 2) is often the natural container to save as one `.rds` file.
:::

## Practice Problem

Assume a colleague sends you `study_summary.rds`, described as "a list with three elements: the cleaned data, a summary table, and today's date." Write the code to read it in and confirm, using tools from Level 2, that it does indeed have exactly 3 top-level elements.

## Challenge Problem

Explain, in your own words, why `.rds` is the right choice for saving your own intermediate analysis steps within a single project, but the *wrong* choice for the final deliverable you'd email to a non-R collaborator — referencing what you learned about CSV and Excel earlier in this level.

## Key Takeaways

- `.rds` is R's native format for saving a single R object exactly as it is, including factor levels and any other type information CSV would flatten away.
- `.rds` files are R-specific — use CSV or Excel instead when sharing data with non-R collaborators or software.
- A single `.rds` file always holds exactly one object; use a list to bundle several related objects into that one file.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `.rds` preserve that CSV does not?
   a) Column names
   b) Row count
   c) Factor levels and exact R object type
   d) File size

2. **True or false.** A colleague using Python can open an `.rds` file directly with no special tools.

3. **Predict the output.** If `x` was saved as `saveRDS(x, "x.rds")` where `x` was a factor with 3 levels, what would `levels(readRDS("x.rds"))` return?

4. **Write the code.** Write the code to read `"data/model_results.rds"` into an object called `model_results`.

<details class="pbl-answer">
<summary>Show answers</summary>

1. **c)** Factor levels and the exact R object type — CSV flattens everything to plain text.
2. **False.** `.rds` is R-specific; Python and other non-R tools cannot open it directly without special conversion libraries.
3. The same 3 levels, in the same order, exactly as they existed before saving.
4. `model_results <- readRDS("data/model_results.rds")`

</details>
:::
