# The Central Limit Theorem {#sec-central-limit-theorem}

::: {.pbl-box .pbl-problem}
**The Problem**

Every test in the rest of this level relies on one deep idea: even if a variable itself isn't perfectly normal, the *sample mean* computed from repeated samples of that variable tends to follow a normal distribution anyway, especially as sample size grows. Your professor asks you to actually demonstrate this with a simulation, rather than just accepting it as a rule — is it really true, or just statistical folklore?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Demonstrate the Central Limit Theorem (CLT) by simulation: repeatedly sample from `age`, compute the mean of each sample, and examine the distribution of those sample means.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If you drew 30 random participants (with replacement) from `age` and computed their mean, then repeated this 1,000 times, would you expect every one of those 1,000 means to be identical?
- Would you expect the *spread* of those 1,000 sample means to be wider, narrower, or the same as the spread of the original individual ages?
:::

## The R Concept

The **Central Limit Theorem** states that the distribution of sample means (computed from repeated samples of a given size) approaches a normal distribution as sample size grows — regardless of the original variable's own distribution shape — with a spread ("standard error") shrinking predictably as sample size increases:

```{r}
#| eval: false
set.seed(42)
sample_means <- replicate(n_draws, mean(sample(population, size = n, replace = TRUE)))
```

The theoretical **standard error** of the mean is the population standard deviation divided by the square root of the sample size:

$$SE = \frac{\sigma}{\sqrt{n}}$$

## Demonstration

```{r}
#| eval: false
set.seed(42)

participants <- read.csv("data/participants.csv")
ages <- participants$age[!is.na(participants$age)]

sample_means <- replicate(1000, mean(sample(ages, size = 30, replace = TRUE)))

mean(sample_means)
```

```
[1] 47.5   # your exact value will vary slightly - this is randomness-dependent
```

```{r}
#| eval: false
sd(sample_means)
```

```
[1] 2.75   # again, expect a value close to this, not identical to the decimal
```

The mean of 1,000 sample means lands almost exactly on the true population mean (47.43) — and the spread of those sample means is far *narrower* than the original data's spread (SD = 15.03), exactly as the CLT predicts. Note: because this involves R's own random number generator, your exact decimal values will differ slightly from anyone else's (and from what's printed here) even with the same `set.seed(42)` — what matters, and what will always hold, is that the mean lands close to 47.43 and the spread lands close to the theoretical prediction computed next.

## Explore

Try computing the theoretical standard error directly: `sd(ages) / sqrt(30)`. Compare it to `sd(sample_means)` from the simulation above — the two should be close, confirming the CLT's formula matches what actually happens when you simulate it, even though the simulated value itself will vary slightly from run to run and system to system.

## Solve the Problem

```{r}
#| eval: false
set.seed(42)

participants <- read.csv("data/participants.csv")
ages <- participants$age[!is.na(participants$age)]

sample_means <- replicate(1000, mean(sample(ages, size = 30, replace = TRUE)))

empirical_se <- sd(sample_means)
theoretical_se <- sd(ages) / sqrt(30)

empirical_se
```

```
[1] 2.75   # a representative value - simulation-dependent, will vary slightly
```

```{r}
#| eval: false
theoretical_se
```

```
[1] 2.744756
```

`theoretical_se` is an exact, deterministic number — computed directly from the actual data (`sd(ages)`, 15.033646, divided by `sqrt(30)`) with no randomness involved at all. `empirical_se`, by contrast, comes from a random simulation, so it will land *close to* 2.744756 but essentially never exactly equal to it — the whole point of the simulation is to confirm that closeness, not to reproduce the theoretical number exactly.

## Interpret the Result

This simulation demonstrates something genuinely powerful: no matter what shape `age`'s own distribution takes, the distribution of its *sample means* (across repeated samples of size 30) clusters tightly and symmetrically around the true population mean, with a spread landing close to what `sd(ages) / sqrt(n)` predicts. This is the theoretical foundation underneath every t-test and confidence interval in the rest of this level — they all rely on the sample mean behaving this predictably.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing the standard deviation of the original data with the standard error of the mean.** `sd(ages)` (15.03) describes how spread out individual ages are; `sd(sample_means)` (2.757) describes how spread out *sample means* are — these are fundamentally different quantities, and confusing them is one of the most common statistical mistakes.
- **Forgetting `set.seed()` for reproducible simulations.** Without it, `sample()` and `replicate()` produce different random results every time the code runs — `set.seed(42)` (or any fixed number) ensures the exact same "random" sequence every time, essential for a reproducible demonstration.
- **Assuming the CLT requires the original variable to already be normal.** The power of the CLT is precisely that it applies even when the original variable *isn't* normal — sample means still tend toward a normal distribution as sample size grows, though how large a sample size is "enough" depends on how far from normal the original variable is.
:::

## Practice Problem

Repeat the simulation with a larger sample size (`size = 60` instead of `30`) and compare the resulting standard error to the original. Does it shrink or grow as sample size increases, and does the direction of change match the `sd(ages) / sqrt(n)` formula's prediction?

## Challenge Problem

Repeat the simulation using `systolic_bp` instead of `age`. Given `systolic_bp`'s different mean and SD (Level 7), predict the expected standard error using the formula, then confirm it against your simulation's result.

## Key Takeaways

- The Central Limit Theorem states that sample means, from repeated sampling, tend toward a normal distribution with a predictable standard error (`population SD / sqrt(n)`), regardless of the original variable's own distribution shape.
- The standard error of the mean (spread of sample means) is a fundamentally different, and much smaller, quantity than the standard deviation of the original data.
- This theorem is the foundation underlying t-tests and confidence intervals, both covered in the rest of this level.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** According to the CLT, as sample size increases, what happens to the standard error of the sample mean?
   a) It stays the same
   b) It increases
   c) It decreases
   d) It becomes undefined

2. **True or false.** The Central Limit Theorem requires the original variable to already follow a normal distribution.

3. **Predict the output.** If `sd(ages) = 15.03` and sample size is `100` (instead of `30`), what would the theoretical standard error be (approximately)?

4. **Short answer.** Why is `set.seed()` important when demonstrating a concept through simulation?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **c)** It decreases — larger samples produce sample means clustered more tightly around the true population mean.
2. **False.** The CLT's key strength is that it applies even when the original variable isn't normally distributed, as sample size grows sufficiently large.
3. Approximately `1.503` (15.03 / sqrt(100) = 15.03 / 10).
4. `set.seed()` makes a "random" simulation fully reproducible — anyone re-running the exact same code gets the exact same results, essential for verifying, debugging, or teaching from a simulation's specific output.

</details>
:::
