# Confidence Intervals {#sec-confidence-intervals}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 7 reported the mean systolic blood pressure as 129.84 — a single number, computed from 96 non-missing participants out of a much larger population of people this study could have sampled. Your professor asks: if you'd sampled a different 96 participants, would you expect to get exactly 129.84 again, or somewhere nearby? How far off might the true population mean plausibly be from this one sample's estimate?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Compute a 95% confidence interval for a sample mean, expressing the uncertainty around a single point estimate like 129.84.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- From Page 2's Central Limit Theorem, what did the standard error of a sample mean describe?
- If a sample mean has some uncertainty around it, would you expect a *range* of plausible values to be more informative than a single number alone?
:::

## The R Concept

A **confidence interval** gives a range of plausible values for a population parameter (like the true mean), built around a sample estimate using its standard error. R's `t.test()` — primarily a hypothesis test (next page) — conveniently also reports a confidence interval for the mean as part of its output:

```{r}
#| eval: false
t.test(x)$conf.int
```

The interval is built from the sample mean, the standard error (Page 2), and a critical value from the **t-distribution** (a close relative of the normal distribution, adjusted for the extra uncertainty of estimating the standard deviation from a limited sample):

$$\text{CI} = \bar{x} \pm t_{crit} \times SE$$

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sbp <- participants$systolic_bp[!is.na(participants$systolic_bp)]

t.test(sbp)$conf.int
```

```
[1] 126.0201 133.6674
attr(,"conf.level")
[1] 0.95
```

The sample mean (129.84) comes with a 95% confidence interval of roughly 126.02 to 133.67 — a range of plausible values for the true population mean, not just the single point estimate alone.

## Explore

Try `t.test(sbp, conf.level = 0.99)$conf.int` — a 99% confidence interval instead of the default 95%. Notice the interval widens: a *higher* confidence level requires a *wider* range to remain equally confident of capturing the true population mean.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sbp <- participants$systolic_bp[!is.na(participants$systolic_bp)]

sbp_mean <- mean(sbp)
sbp_se <- sd(sbp) / sqrt(length(sbp))
t_critical <- qt(0.975, df = length(sbp) - 1)

lower_bound <- sbp_mean - t_critical * sbp_se
upper_bound <- sbp_mean + t_critical * sbp_se

c(lower_bound, upper_bound)
```

```
[1] 126.0201 133.6674
```

```{r}
#| eval: false
t.test(sbp)$conf.int
```

```
[1] 126.0201 133.6674
attr(,"conf.level")
[1] 0.95
```

Building the interval manually (mean ± critical value × SE) produces exactly the same result as `t.test()$conf.int` — confirming the formula and the built-in function agree, and showing explicitly what `t.test()` computes behind the scenes.

## Interpret the Result

You can now say something more precise than "the mean systolic blood pressure was 129.84": you're 95% confident the true population mean lies somewhere between 126.02 and 133.67 mmHg. This is a meaningfully different, more honest statement than reporting the single point estimate alone — it explicitly communicates the uncertainty that comes from having only 96 participants rather than the entire population.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Interpreting a 95% CI as "95% probability the true mean is in this range."** The more precise interpretation: if you repeated this sampling process many times, about 95% of the resulting confidence intervals would contain the true population mean — a subtle, but real, distinction from a probability statement about this one specific interval.
- **Assuming a narrower interval always means a better estimate.** A narrower interval usually comes from a *larger* sample size, not necessarily a "better" one — comparing intervals from very different sample sizes without accounting for this can be misleading.
- **Forgetting confidence intervals depend on the same normality assumptions checked on Page 1.** For very small samples with heavily skewed data, `t.test()`'s confidence interval can become less reliable — always sanity-check the underlying assumptions, not just run the function blindly.
:::

## Practice Problem

Compute a 95% confidence interval for `bmi`'s mean using `t.test()`. Does the interval contain the value 27 (roughly the boundary between "Normal" and "Overweight," Level 4's `case_when()` categories)?

## Challenge Problem

Compute confidence intervals for `age`'s mean at three different confidence levels (90%, 95%, 99%) using `t.test(age, conf.level = ...)`. Describe the pattern in how the interval's width changes as the confidence level increases, and explain why that pattern makes sense.

## Key Takeaways

- A confidence interval expresses a range of plausible values for a population parameter, built from a sample estimate and its standard error.
- `t.test(x)$conf.int` computes this directly; the same interval can be built manually as `mean ± critical_value × SE`.
- A higher confidence level produces a wider interval — there's an inherent trade-off between how confident you are and how precise (narrow) the range can be.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does a 95% confidence interval represent?
   a) There is a 95% probability the sample mean is correct
   b) A range of plausible values for the population parameter, with a procedure that captures it about 95% of the time across repeated sampling
   c) 95% of the individual data points fall within this range
   d) The interval within which all future samples must fall

2. **True or false.** A higher confidence level (say, 99% instead of 95%) always produces a narrower interval.

3. **Predict the output.** If the sample mean is 129.84 and the 95% CI is (126.02, 133.67), would a 90% CI be wider or narrower than this range?

4. **Find the error.**
   ```r
   # "There's a 95% chance the true population mean is 129.84."
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** A range of plausible values for the population parameter, with a procedure that captures the true value about 95% of the time across repeated sampling.
2. **False.** A higher confidence level requires a *wider* interval to maintain that higher level of confidence, not a narrower one.
3. Narrower — a lower confidence level (90%) requires a smaller range to achieve that lower degree of confidence.
4. This statement misinterprets both the point estimate and the confidence interval — the sample mean (129.84) is a single computed value, not a probabilistic claim, and the correct interpretation of a 95% CI is about the interval-generating *procedure* capturing the true mean across repeated sampling, not a 95% probability attached to any single specific value.

</details>
:::
