# The One-Sample t-Test {#sec-one-sample-t-test}

::: {.pbl-box .pbl-problem}
**The Problem**

A clinical guideline states that normal systolic blood pressure is around 120 mmHg. Your professor asks: does this *study sample's* mean (129.84) differ from 120 by more than could plausibly be due to chance alone — or could the true population this sample represents actually have a mean of 120, with 129.84 just being one somewhat-higher-than-typical sample?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Test whether a sample's mean differs significantly from a specific reference value, using a one-sample t-test.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If chance alone could easily produce a sample mean of 129.84 from a true population mean of 120, would you expect that to happen often, or rarely?
- From Page 3, does the 95% confidence interval for mean systolic blood pressure (126.02 to 133.67) contain the reference value 120?
:::

## The R Concept

A **one-sample t-test** compares a sample's mean against a specific hypothesized value (`mu`), producing a **p-value**: the probability of observing a difference at least this large, purely by chance, if the true population mean actually equaled the hypothesized value:

```{r}
#| eval: false
t.test(x, mu = hypothesized_value)
```

A small p-value (conventionally, below 0.05) suggests the observed difference is unlikely to be due to chance alone — evidence *against* the hypothesized value being the true population mean.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sbp <- participants$systolic_bp[!is.na(participants$systolic_bp)]

t.test(sbp, mu = 120)
```

```

	One Sample t-test

data:  sbp
t = 5.111, df = 95, p-value = 1.655e-06
alternative hypothesis: true mean is not equal to 120
95 percent confidence interval:
 126.0201 133.6674
sample estimates:
mean of x
 129.8438
```

## Explore

Notice the confidence interval reported here (126.02 to 133.67) is exactly the same interval from Page 3 — `t.test()` always reports both a hypothesis test *and* a confidence interval together, since they're built from the same underlying calculation. Try `t.test(sbp, mu = 130)` instead — since 130 falls *inside* the confidence interval, you'd expect a much larger p-value (failing to reject that particular hypothesized value).

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sbp <- participants$systolic_bp[!is.na(participants$systolic_bp)]

test_result <- t.test(sbp, mu = 120)

test_result$statistic
```

```
       t
5.110964
```

```{r}
#| eval: false
test_result$p.value
```

```
[1] 1.655098e-06
```

The t-statistic (5.11) is large, and the p-value (0.0000017, far below 0.05) is extremely small — strong evidence that this sample's mean genuinely differs from 120, not merely due to random chance.

## Interpret the Result

The sample's mean systolic blood pressure (129.84) is statistically significantly different from the reference value of 120 (p < 0.001) — consistent with the fact that 120 doesn't even fall inside the 95% confidence interval (126.02 to 133.67) computed on Page 3. This is genuine evidence, not just a numeric coincidence: a sample mean this far from 120, in a sample of 96, would be extremely unlikely to occur by chance alone if the true population mean actually were 120.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing statistical significance with practical (clinical) importance.** A p-value this small confirms the difference from 120 is unlikely to be chance — but whether a roughly 10 mmHg difference is *clinically* meaningful is a separate, non-statistical judgment that depends on domain knowledge, not just the p-value.
- **Treating p < 0.05 as a magic threshold rather than a convention.** 0.05 is a widely used convention, not a law of nature — a p-value of 0.049 and 0.051 represent nearly identical evidence, despite falling on opposite sides of this threshold.
- **Forgetting the confidence interval and the p-value are two views of the same information.** If a hypothesized value falls *outside* the confidence interval (as 120 does here), the corresponding t-test will always produce p < 0.05 — the two are mathematically linked, not independent pieces of evidence.
:::

## Practice Problem

Test whether mean `bmi` differs significantly from 25 (the conventional upper boundary of "Normal" BMI, Level 4). Report the t-statistic, p-value, and your conclusion.

## Challenge Problem

Test whether mean `age` differs significantly from 45. Then check whether 45 falls inside or outside the 95% confidence interval for `age`'s mean (Page 3's practice problem) — confirm the two approaches (hypothesis test and confidence interval) lead to the same conclusion.

## Key Takeaways

- A one-sample t-test compares a sample mean against a specific hypothesized value, producing a p-value indicating how likely such a difference would be purely by chance.
- A small p-value (conventionally < 0.05) is evidence against the hypothesized value; it does not by itself indicate the size or practical importance of the difference.
- A hypothesized value falling outside a confidence interval and a t-test producing p < 0.05 always agree — they're two expressions of the same underlying calculation.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does the p-value from `t.test(sbp, mu = 120)` represent?
   a) The probability the true mean is exactly 120
   b) The probability of observing a difference this large purely by chance, if the true mean were actually 120
   c) The sample mean itself
   d) The percentage of participants with blood pressure above 120

2. **True or false.** A statistically significant result automatically means the difference is large enough to matter clinically.

3. **Predict the output.** If a hypothesized value falls inside a variable's 95% confidence interval, would you expect its corresponding one-sample t-test to produce p < 0.05 or p ≥ 0.05?

4. **Short answer.** Why is treating p = 0.05 as a rigid pass/fail cutoff considered poor statistical practice?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The probability of observing a difference this large purely by chance, if the true population mean were actually 120.
2. **False.** Statistical significance indicates the difference is unlikely to be due to chance — it says nothing about whether that difference is large enough to matter practically or clinically.
3. p ≥ 0.05 — a hypothesized value inside the confidence interval indicates the data is consistent with that value, corresponding to a non-significant test result.
4. Because 0.05 is an arbitrary convention, not a meaningful boundary in the underlying evidence — a p-value of 0.049 and 0.051 reflect nearly identical strength of evidence, so treating one as "significant" and the other as "not significant" overstates the importance of a fairly arbitrary line.

</details>
:::
