# The Chi-Square Test of Independence {#sec-chi-square-test}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 7's cross-tabulation found a striking-looking descriptive pattern: 19.2% of women smoke, compared to 31.2% of men — a 12-percentage-point gap. But Level 7 also cautioned that this was a *descriptive* finding, not yet a tested one. Your professor now asks the question Level 7 deliberately left open: is this sex difference in smoking rate statistically significant, or could a gap this size easily arise by chance in a sample of only 100 participants?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Test whether two categorical variables are statistically associated, using a chi-square test of independence.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- If sex and smoking status were completely unrelated, would you still expect the exact same smoking rate in both groups, or just *roughly similar* rates?
- From Level 7, how many women and men were in this sample (52 and 48)? Does a 12-point gap in a sample this size feel like a lot of data to draw a firm conclusion from?
:::

## The R Concept

The **chi-square test of independence** compares a cross-tabulation's *observed* counts against the counts you'd *expect* if the two variables were completely unrelated, producing a p-value:

```{r}
#| eval: false
chisq.test(table(var1, var2))
```

By default, R applies a continuity correction (Yates' correction) for 2×2 tables, which slightly adjusts the test to be more conservative for small samples.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sex_smoker_table <- table(participants$sex, participants$smoker)
sex_smoker_table
```

```
        No Yes
Female  42  10
Male    33  15
```

```{r}
#| eval: false
chisq.test(sex_smoker_table)
```

```

	Pearson's Chi-squared test with Yates' continuity correction

data:  sex_smoker_table
X-squared = 1.3355, df = 1, p-value = 0.2478
```

## Explore

Try `chisq.test(sex_smoker_table)$expected` to see the counts that *would* be expected if sex and smoking status were completely independent: roughly 39 female non-smokers and 13 female smokers (compared to the actual 42 and 10). The chi-square statistic measures exactly how far the observed counts (42, 10, 33, 15) deviate from these expected ones.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sex_smoker_table <- table(participants$sex, participants$smoker)

test_result <- chisq.test(sex_smoker_table)

test_result$statistic
```

```
X-squared
 1.335470
```

```{r}
#| eval: false
test_result$p.value
```

```
[1] 0.2478344
```

```{r}
#| eval: false
test_result$expected
```

```
        No Yes
Female  39  13
Male    36  12
```

A p-value of 0.248 is well above the conventional 0.05 threshold — despite the 12-percentage-point descriptive gap Level 7 found, this chi-square test provides no strong statistical evidence that sex and smoking status are genuinely associated in this sample.

## Interpret the Result

This is a genuinely important, humbling result: Level 7's 19.2%-versus-31.2% gap *looked* substantial when described in percentages, but with only 52 women and 48 men, that gap is well within the range chance alone could plausibly produce. This is exactly why Level 7 cautioned against treating a cross-tabulation's pattern as a confirmed finding — the formal test now confirms that caution was warranted, not just a reflexive disclaimer.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Treating a large descriptive gap (12 percentage points) as automatically meaningful.** Percentage-point differences can look dramatic without accounting for sample size — a chi-square test (or any formal significance test) directly addresses whether the gap is larger than chance would typically produce, not just whether it looks big.
- **Applying a chi-square test to cell counts that are too small.** `chisq.test()` can become unreliable when expected cell counts are very small (a common rule of thumb: below 5) — always check `$expected` before trusting the result; here, all expected counts (39, 13, 36, 12) are comfortably above that threshold.
- **Confusing "not statistically significant" with "definitely no association exists."** As with the t-tests earlier in this level, a non-significant chi-square result means this sample doesn't provide strong evidence of an association — it doesn't prove sex and smoking are truly unrelated in the broader population this sample represents.
:::

## Practice Problem

Run a chi-square test on `county` and `smoker` (a larger, 5×2 table). Check `$expected` first to confirm no expected cell count falls below 5, then interpret the resulting p-value.

## Challenge Problem

Compare `chisq.test(sex_smoker_table)` (with Yates' correction, R's default) to `chisq.test(sex_smoker_table, correct = FALSE)` (without it). How much does the p-value change? Research why Yates' correction tends to matter more for smaller samples and 2×2 tables specifically.

## Key Takeaways

- The chi-square test of independence compares observed cross-tabulation counts against what would be expected under complete independence, producing a p-value.
- A descriptive percentage-point gap can look substantial while still being entirely consistent with chance, especially in smaller samples — always follow up a descriptive cross-tabulation (Level 7) with a formal test before drawing conclusions.
- Check `$expected` cell counts before trusting a chi-square result — very small expected counts (conventionally below 5) can make the test unreliable.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does the chi-square test of independence evaluate?
   a) Whether two numeric variables are correlated
   b) Whether two categorical variables are associated, beyond what chance would produce
   c) Whether a sample mean differs from a reference value
   d) Whether a variable is normally distributed

2. **True or false.** A large descriptive percentage-point gap between two groups always corresponds to a statistically significant chi-square test result.

3. **Predict the output.** Given a p-value of 0.248 for the sex-smoking chi-square test, would you conclude there's strong evidence of an association?

4. **Short answer.** Why is it important to check expected cell counts before trusting a chi-square test's result?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Whether two categorical variables are associated, beyond what chance would produce.
2. **False.** As shown here, a 12-percentage-point gap produced a non-significant result (p = 0.248) — the size of a descriptive gap and its statistical significance depend heavily on sample size, and don't always align.
3. No — a p-value of 0.248 is well above the conventional 0.05 threshold, indicating no strong evidence of an association in this sample.
4. The chi-square test's approximation becomes less reliable when expected cell counts are very small (conventionally below 5) — trusting a result built on unreliable expected counts risks drawing an incorrect conclusion from a test that wasn't appropriate for that data in the first place.

</details>
:::
