# Level 11 Capstone: Testing This Course's Earlier Findings {#sec-statistics-capstone}

::: {.pbl-box .pbl-problem}
**The Problem**

This level's opening mission, finally answerable: Level 7 found smoking rates differ by sex (19.2% women, 31.2% men), and Level 8's trend line showed blood pressure declining gently with age. Both were purely descriptive findings at the time — plots and percentages, without a formal test behind them. Your professor wants a final, side-by-side verdict: which of these two patterns holds up to formal statistical testing, and which doesn't?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Formally test both of this course's earlier descriptive findings, using the tools from this entire level, and report a clear, honest verdict on each.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Which test from this level is appropriate for "does smoking rate differ by sex" — a chi-square test, a t-test, or a correlation test?
- Which test is appropriate for "does blood pressure decline with age" — the same test, or a different one?
:::

## The R Concept

Choosing the right test depends entirely on the *type* of variables involved:

```{r}
#| eval: false
# Two categorical variables -> chi-square test
chisq.test(table(categorical1, categorical2))

# One numeric, one numeric (linear relationship) -> correlation test or regression
cor.test(numeric1, numeric2)
lm(numeric1 ~ numeric2)
```

Sex and smoking status are both categorical — a chi-square test (Page 7) is the right tool. Age and blood pressure are both numeric — a correlation test (Page 8) or simple linear regression (Page 9) are both appropriate, and (as Page 9's challenge problem hinted) mathematically equivalent for a single predictor.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

# Finding 1: sex and smoking status (Level 7's cross-tabulation)
sex_smoker_table <- table(participants$sex, participants$smoker)
chisq.test(sex_smoker_table)

# Finding 2: age and blood pressure (Level 8's trend line)
cor.test(participants$age, participants$systolic_bp)
```

```

	Pearson's Chi-squared test with Yates' continuity correction

data:  sex_smoker_table
X-squared = 1.3355, df = 1, p-value = 0.2478


	Pearson's product-moment correlation

data:  participants$age and participants$systolic_bp
t = -2.1266, df = 88, p-value = 0.03626
alternative hypothesis: true correlation is not equal to 0
95 percent confidence interval:
 -0.4094278 -0.0146621
sample estimates:
       cor
-0.2210819
```

## Explore

Notice the two p-values point in opposite directions: 0.248 (not significant) for sex-and-smoking, but 0.036 (significant) for age-and-blood-pressure. Both were originally reported in this course as comparably "modest" descriptive patterns — only the formal tests reveal that one holds up statistically and the other doesn't.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

sex_smoker_test <- chisq.test(table(participants$sex, participants$smoker))
age_bp_test <- cor.test(participants$age, participants$systolic_bp)

sex_smoker_test$p.value
```

```
[1] 0.2478344
```

```{r}
#| eval: false
age_bp_test$p.value
```

```
[1] 0.03625678
```

```{r}
#| eval: false
age_bp_test$estimate
```

```
       cor
-0.2210819
```

**Finding 1 (sex and smoking status): not statistically significant** (p = 0.248). Despite a 12-percentage-point descriptive gap, this sample of 52 women and 48 men doesn't provide strong evidence of a true association between sex and smoking status.

**Finding 2 (age and blood pressure): statistically significant** (p = 0.036, r = -0.221). Older participants do show a modest but statistically real tendency toward lower systolic blood pressure in this sample — confirmed by both the correlation test here and the regression slope test on Page 9.

## Interpret the Result

This side-by-side verdict is exactly the kind of honest, complete conclusion a real analysis should reach: one of Level 7 and Level 8's two earlier patterns turned out to be statistically robust (age and blood pressure), while the other — despite looking similarly notable in a plot or percentage table — did not hold up (sex and smoking status). Neither conclusion is a "failure": correctly identifying which patterns are statistically supported and which aren't is precisely the value this entire level adds on top of Levels 5, 7, and 8's descriptive and visual work.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Reporting only the significant finding and quietly dropping the non-significant one.** A complete, honest analysis reports both results — a non-significant finding is still a real, useful piece of information (it tells you a suspected pattern isn't well-supported by this data), not something to omit because it's less exciting.
- **Assuming a smaller p-value always means a "more important" finding.** The age-blood pressure relationship (p = 0.036, r = -0.221) is statistically significant but still numerically weak (R-squared under 5%, Page 9) — statistical significance and practical importance remain two separate questions, exactly as emphasized throughout this level.
- **Forgetting that "significant" and "not significant" are both provisional conclusions from one particular sample.** A larger, independently collected sample could, in principle, find sex and smoking status significantly associated after all, or fail to replicate the age-blood pressure finding — every result in this level describes this specific sample of 100 participants, not an eternal truth about the wider population.
:::

## Practice Problem

Choose one more descriptive finding from earlier in this course (for example, Level 5's blood pressure ranking by county, or Level 8's BMI-by-county faceted histogram) and select the appropriate formal test from this level to check whether it holds up statistically.

## Challenge Problem

Write a short paragraph (as if for the final section of a real report) summarizing both findings from this capstone: state each result, its statistical significance (or lack of it), its effect size (correlation or percentage-point gap), and an honest, appropriately cautious interpretation of what both findings do and don't establish about this dataset.

## Key Takeaways

- Choosing the correct statistical test depends on variable type: chi-square for two categorical variables, correlation/regression for two numeric variables, t-tests for comparing means across groups.
- A complete analysis reports both significant and non-significant results honestly — omitting inconvenient non-significant findings misrepresents the evidence.
- Statistical significance and practical/clinical importance remain distinct: a significant result can still describe a weak relationship, and every conclusion in this level is provisional to this specific sample of 100 participants.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Which test correctly evaluates whether sex and smoking status (both categorical) are associated?
   a) A t-test
   b) A chi-square test
   c) A correlation test
   d) Simple linear regression

2. **True or false.** Both of this course's earlier descriptive findings (sex-smoking, age-blood pressure) turned out to be statistically significant when formally tested.

3. **Predict the output.** Given a p-value of 0.036 for the age-blood pressure correlation, would you conclude this relationship is statistically significant at the conventional 0.05 threshold?

4. **Short answer.** Why is it important to report a non-significant finding (like the sex-smoking result) rather than omitting it from a final report?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** A chi-square test — the appropriate tool for testing association between two categorical variables.
2. **False.** Only the age-blood pressure relationship (p = 0.036) was statistically significant; the sex-smoking association (p = 0.248) was not.
3. Yes — 0.036 is below the conventional 0.05 threshold, indicating statistical significance.
4. Omitting non-significant findings misrepresents the true state of evidence — a reader deserves to know that a plausible-looking descriptive pattern (like the sex-smoking gap) did not hold up to formal testing, just as much as they deserve to know which patterns did.

</details>
:::
