# Multiple Regression with `lm()` {#sec-multiple-regression}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 11 Page 9 fit `lm(systolic_bp ~ age)` and found a significant, but weak, relationship (R-squared = 0.049). Your professor points out that blood pressure is influenced by more than just age — BMI is a well-known factor too. She wants to know: what happens to the model once BMI is added alongside age?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Fit a multiple regression model with two numeric predictors at once, and interpret each coefficient while holding the other predictor constant.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- In `lm(y ~ x1 + x2)`, do you expect the coefficient on `x1` to mean exactly the same thing as it did in a simple regression with only `x1`?
- If BMI is also related to blood pressure, could adding it to the model change how much of blood pressure's variation age alone appears to explain?
:::

## The R Concept

Adding a second predictor to `lm()` is as simple as joining it with `+`:

```{r}
#| eval: false
model <- lm(outcome ~ predictor1 + predictor2, data = df)
summary(model)
```

Each coefficient in a multiple regression is now interpreted **holding the other predictor(s) constant** — the age coefficient represents the predicted change in the outcome per year of age, *for participants with the same BMI*. This is a subtly different (and usually more honest) interpretation than a simple regression's coefficient, which does not account for anything else.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

bp_age_bmi_model <- lm(systolic_bp ~ age + bmi, data = participants)

summary(bp_age_bmi_model)
```

```

Call:
lm(formula = systolic_bp ~ age + bmi, data = participants)

Residuals:
    Min      1Q  Median      3Q     Max
-33.454 -12.720  -1.673  10.982  44.223

Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept) 123.1658    13.4226   9.176  3.21e-14 ***
age          -0.3052     0.1337  -2.283    0.0250 *
bmi           0.7345     0.4400   1.669    0.0989 .
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 18.36 on 82 degrees of freedom
  (15 observations deleted due to missingness)
Multiple R-squared:  0.08657,	Adjusted R-squared:  0.06429
F-statistic: 3.886 on 2 and 82 DF,  p-value: 0.02439
```

## Explore

Try comparing this model's R-squared (0.087) to Page 9's age-only model (0.049, on a slightly different sample of 90 due to `bmi`'s own missing values). Adding BMI roughly doubled the explained variation — but notice the *sample size* also dropped, since `lm()` drops any row missing either `age`, `bmi`, or `systolic_bp`.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

bp_age_bmi_model <- lm(systolic_bp ~ age + bmi, data = participants)

coef(bp_age_bmi_model)
```

```
(Intercept)         age         bmi
123.1658246  -0.3052224   0.7345009
```

```{r}
#| eval: false
summary(bp_age_bmi_model)$r.squared
summary(bp_age_bmi_model)$adj.r.squared
```

```
[1] 0.08656528
[1] 0.06428639
```

The fitted equation is **predicted blood pressure = 123.17 − 0.305 × age + 0.735 × bmi**: holding BMI constant, each additional year of age predicts about 0.305 mmHg lower systolic blood pressure; holding age constant, each additional BMI unit predicts about 0.735 mmHg higher systolic blood pressure. Age remains statistically significant (p = 0.025); BMI is not quite significant at the conventional 0.05 threshold (p = 0.099).

## Interpret the Result

Age's coefficient barely changed once BMI joined the model (−0.305 here vs. −0.276 in Page 9's simple model, on a smaller matched sample) — a first hint that BMI isn't strongly confounding the age-blood-pressure relationship in this data (Page 4 investigates this properly). The model overall explains more variation than age alone (R-squared 0.087 vs. 0.049), though both predictors together still leave over 90% of blood pressure's variation unexplained — entirely consistent with blood pressure being influenced by many factors this course's dataset doesn't capture.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Interpreting a multiple regression coefficient the same way as a simple regression coefficient.** In `lm(bp ~ age + bmi)`, age's coefficient means "the effect of age *holding BMI constant*" — not "the overall effect of age ignoring everything else," which is what Page 9's simple model estimated instead.
- **Not noticing the sample size changed.** Adding a predictor with its own missing values (`bmi` has 5 missing entries) can shrink the usable sample further than a single-predictor model — always check "observations deleted due to missingness" in the output before comparing models.
- **Assuming more predictors always means a better model.** R-squared can only stay the same or increase as predictors are added (Page 5 shows why adjusted R-squared, and a formal test, are the honest way to judge whether a new predictor earns its place).
:::

## Practice Problem

Fit `lm(systolic_bp ~ age + diastolic_bp, data = participants)`. Given Level 7's very strong correlation (0.92) between systolic and diastolic blood pressure, do you expect `diastolic_bp`'s coefficient to be large or small, and its p-value to be significant or not?

## Challenge Problem

Fit a three-predictor model: `lm(systolic_bp ~ age + bmi + diastolic_bp, data = participants)`. Compare the age and bmi coefficients here to the two-predictor model above — do they change much once `diastolic_bp` also joins the model? This previews Page 7's multicollinearity discussion, since systolic and diastolic blood pressure are very strongly correlated with each other.

## Key Takeaways

- `lm(outcome ~ predictor1 + predictor2, data = df)` fits a multiple regression; each coefficient represents that predictor's effect holding the other predictor(s) constant.
- Adding predictors can change both the estimated coefficients and the usable sample size (due to each predictor's own missing values) — always check both.
- A higher R-squared with more predictors doesn't automatically mean a meaningfully better model — Page 5 covers the formal way to test this.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In `lm(systolic_bp ~ age + bmi)`, what does the coefficient on `age` represent?
   a) The overall correlation between age and blood pressure
   b) The effect of age on blood pressure, holding BMI constant
   c) The effect of BMI on blood pressure, holding age constant
   d) The combined effect of age and BMI together

2. **True or false.** Adding a second predictor to a regression model can only increase (or leave unchanged) the model's R-squared, never decrease it.

3. **Predict the output.** If a model drops from `n = 90` (one predictor) to `n = 85` (two predictors), what is the most likely explanation?

4. **Find the error.**
   ```r
   # "Age's coefficient is -0.305 in the two-predictor model, meaning age alone
   #  explains a 0.305 mmHg drop in blood pressure per year, regardless of BMI."
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The effect of age on blood pressure, holding BMI constant.
2. **True.** R-squared is mathematically guaranteed to never decrease when a predictor is added (even a useless one) — this is exactly why adjusted R-squared and formal model comparison tests (Page 5) matter.
3. The second predictor (`bmi`) likely has its own missing values in rows that were previously usable when only `age` was in the model — `lm()` drops any row missing a value for *any* variable used in the model.
4. This statement is backwards — the coefficient of −0.305 explicitly *is* age's effect while accounting for (not ignoring) BMI; describing it as "regardless of BMI" would actually describe the *simple* regression's coefficient (Page 9's −0.276), not this multiple regression's.

</details>
:::
