# Regression Diagnostics {#sec-regression-diagnostics}

::: {.pbl-box .pbl-problem}
**The Problem**

Every `lm()` result so far has been reported at face value — coefficients, p-values, R-squared. But `lm()`'s p-values are only trustworthy if certain assumptions roughly hold: that the residuals (prediction errors) are approximately normally distributed, and that they're spread out evenly regardless of the predicted value. Your professor asks: how would you actually check whether Page 5's model meets these assumptions, rather than just trusting the output blindly?
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Check a fitted regression model's residuals for approximate normality and constant variance, using both a numeric test and a visual check.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- A **residual** is the difference between an observed value and the model's prediction for that row. If `lm()`'s assumptions hold well, would you expect residuals to be scattered evenly, or to show an obvious pattern?
- Level 11 Page 1 used `shapiro.test()` conceptually to check normality. Could the same test be applied to a model's residuals instead of raw data?
:::

## The R Concept

**`residuals(model)`** extracts every observation's prediction error. Two common diagnostic checks:

```{r}
#| eval: false
shapiro.test(residuals(model))       # tests residual normality
plot(model, which = 1)               # residuals vs. fitted values (constant variance check)
```

A `shapiro.test()` p-value above 0.05 suggests no strong evidence against normality (Level 11 Page 1's same logic, applied to residuals instead of raw data). For constant variance ("homoscedasticity"), the **Breusch-Pagan test** (from the `lmtest` package) formalizes what a residuals-vs-fitted plot shows visually:

```{r}
#| eval: false
library(lmtest)
bptest(model)
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

complete_rows <- participants[complete.cases(participants[, c("age", "bmi", "smoker", "systolic_bp")]), ]
complete_rows$smoker_binary <- as.numeric(complete_rows$smoker == "Yes")

model_big <- lm(systolic_bp ~ age + bmi + smoker_binary, data = complete_rows)

shapiro.test(residuals(model_big))
```

```

	Shapiro-Wilk normality test

data:  residuals(model_big)
W = 0.98499, p-value = 0.4303
```

## Explore

Try `plot(model_big, which = 1)` (a residuals-vs-fitted plot) and `plot(model_big, which = 2)` (a Q-Q plot) — the two most common visual diagnostics. A Q-Q plot with points falling roughly along the diagonal line supports the same normality conclusion the Shapiro-Wilk test gives numerically.

## Solve the Problem

```{r}
#| eval: false
library(lmtest)

participants <- read.csv("data/participants.csv")

complete_rows <- participants[complete.cases(participants[, c("age", "bmi", "smoker", "systolic_bp")]), ]
complete_rows$smoker_binary <- as.numeric(complete_rows$smoker == "Yes")

model_big <- lm(systolic_bp ~ age + bmi + smoker_binary, data = complete_rows)

shapiro.test(residuals(model_big))$p.value
bptest(model_big)
```

```
[1] 0.4303420
```

```

	studentized Breusch-Pagan test

data:  model_big
BP = 1.262, df = 3, p-value = 0.7382
```

Both diagnostic tests come back reassuring: Shapiro-Wilk's p-value (0.430) shows no significant evidence against residual normality, and Breusch-Pagan's p-value (0.738) shows no significant evidence against constant variance. Neither test result *proves* the assumptions hold perfectly — but neither raises a red flag either.

## Interpret the Result

This model's residuals behave close to what `lm()` assumes: roughly normal and roughly evenly spread. That means the p-values reported throughout Pages 1-5 (for `age`, `bmi`, `smoker_binary`, and the model comparisons) can be reasonably trusted — if either diagnostic test had come back strongly significant instead, it would be a signal to interpret those p-values with real caution, or consider a different modeling approach entirely.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Treating a non-significant diagnostic test as definitive proof assumptions are perfectly met.** `shapiro.test()` and `bptest()` failing to detect a violation isn't the same as guaranteeing there isn't one, especially in smaller samples (n = 85 here) where these tests have limited power to detect subtle problems.
- **Skipping diagnostics entirely because a model's coefficients "look reasonable."** A model's coefficients and p-values can look perfectly plausible even when its underlying assumptions are badly violated — diagnostics are a required step, not an optional one, for any model whose conclusions matter.
- **Running diagnostics but never acting on a bad result.** If a diagnostic test flags a real problem (severe non-normality or non-constant variance), the appropriate response is usually to reconsider the model — a data transformation, a different type of model, or robust standard errors — not to report the flawed model's p-values unchanged.
:::

## Practice Problem

Run `shapiro.test()` and `bptest()` on Page 1's simpler two-predictor model (`systolic_bp ~ age + bmi`). Do both diagnostics still look reassuring?

## Challenge Problem

Run `plot(model_big, which = 1)` and describe, in your own words, what an *ideal* residuals-vs-fitted plot looks like (in terms of the spread and pattern of points), versus what a *problematic* one (indicating non-constant variance) would look like instead.

## Key Takeaways

- `shapiro.test(residuals(model))` checks whether a regression model's residuals are approximately normally distributed.
- `bptest(model)` (from the `lmtest` package) formally tests for constant variance ("homoscedasticity") across the range of predicted values.
- A non-significant diagnostic test result supports (but never definitively proves) that a model's assumptions are reasonably met — always pair numeric diagnostic tests with a visual check like `plot(model)`.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `shapiro.test(residuals(model))` check?
   a) Whether the predictors are correlated with each other
   b) Whether the model's residuals are approximately normally distributed
   c) Whether the model has enough predictors
   d) Whether the outcome variable has missing values

2. **True or false.** A non-significant Breusch-Pagan test result definitively proves a model has constant variance with no possible exceptions.

3. **Predict the output.** If `bptest()` returns a p-value of 0.738, is there strong evidence of non-constant variance in the residuals?

4. **Short answer.** Why is it important to pair a numeric diagnostic test (like `shapiro.test()`) with a visual check (like a Q-Q plot)?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** Whether the model's residuals are approximately normally distributed.
2. **False.** A non-significant result supports, but never definitively proves, that the assumption holds — especially in smaller samples where the test has limited power.
3. No — a p-value of 0.738 is far above 0.05, showing no significant evidence of non-constant variance.
4. Numeric tests provide an overall summary, but a visual check can reveal patterns (like a specific range of values behaving oddly) that a single p-value might not fully capture — used together, they give a more complete diagnostic picture.

</details>
:::
