# `while` Loops {#sec-while-loops}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor asks a strange but genuine question: "If we kept removing our oldest participant, one at a time, how many would we need to remove before the remaining sample's mean age drops below 40?" A `for` loop (Page 1) needs to know in advance exactly how many times to run — but here, you don't know that number ahead of time. You only know the *condition* you're waiting for ("mean age is still 40 or above"), not the exact iteration count.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Repeat a block of code until a condition is no longer true, using a `while` loop.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- A `for` loop runs a fixed, known number of times. If you don't know in advance how many iterations you'll need, is a `for` loop still the right tool?
- Every `while` loop needs something inside it that eventually makes its condition `FALSE` — what would that "something" be for the mean-age-removal question above?
:::

## The R Concept

A **`while` loop** repeats a block of code for as long as a condition remains `TRUE`, checking that condition again before every iteration:

```{r}
#| eval: false
while (condition) {
  # code repeated as long as condition is TRUE
  # something inside here must eventually make condition FALSE
}
```

Unlike a `for` loop, a `while` loop doesn't know in advance how many times it will run — it simply keeps going until its condition becomes `FALSE`. This makes `while` loops appropriate for exactly the kind of "keep going until..." question `for` loops aren't naturally suited for.

## Demonstration

```{r}
#| eval: false
count <- 0

while (count < 5) {
  print(count)
  count <- count + 1
}
```

```
[1] 0
[1] 1
[1] 2
[1] 3
[1] 4
```

The loop checks `count < 5` before every iteration, stopping the moment `count` reaches `5` — note that `count <- count + 1` inside the loop is what eventually makes the condition `FALSE`; without it, this loop would run forever (an "infinite loop," a common bug to watch for).

## Explore

Try removing the `count <- count + 1` line from the loop above (don't actually run it, just consider it). Since `count` would never change, `count < 5` would remain `TRUE` forever — this is exactly why every `while` loop needs something inside it that moves its condition toward becoming `FALSE`.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

ages <- sort(participants$age[!is.na(participants$age)], decreasing = TRUE)

removed_count <- 0

while (mean(ages) >= 40) {
  ages <- ages[-1]           # remove the current oldest (first, since sorted descending)
  removed_count <- removed_count + 1
}

removed_count
```

```
[1] 29
```

```{r}
#| eval: false
mean(ages)
```

```
[1] 39.95385
```

```{r}
#| eval: false
length(ages)
```

```
[1] 65
```

The loop removed the oldest participant, one at a time, checking the mean age after each removal — stopping after exactly 29 removals, once the remaining 65 participants' mean age dropped to 39.95, just below the 40 threshold.

## Interpret the Result

Nearly a third of the original 94 non-missing ages (29 of 94) had to be removed — specifically, the oldest third — before the sample's mean age dropped below 40. This is a genuinely different kind of question than anything answerable with a `for` loop directly: you needed to keep going until a *condition* was met, with no way to know in advance exactly how many iterations that would take.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting to include something that changes the condition inside the loop.** This is the single most common `while` loop bug — an infinite loop, where the condition never becomes `FALSE` because nothing inside the loop actually moves toward that outcome.
- **Using `while` when a `for` loop (or vectorized approach) would be simpler.** If you already know exactly how many iterations you need, a `for` loop is usually clearer — `while` loops are best reserved for situations where the stopping point genuinely depends on a condition evaluated during the loop, not a known count in advance.
- **Not having an escape plan for a `while` loop that might never terminate.** If there's any chance the condition could never become `FALSE` (a data problem, a logic error), consider adding a safety counter (e.g., stop after 1000 iterations regardless) to prevent a truly infinite loop from freezing your session.
:::

## Practice Problem

Write a `while` loop that removes the *youngest* participant, one at a time, until the remaining sample's mean age exceeds 55. How many participants need to be removed?

## Challenge Problem

Modify the blood-pressure equivalent of this question: starting from all non-missing `systolic_bp` values, repeatedly remove the single highest reading until the mean drops below 125. How many removals are required, and what does this suggest about how much a handful of high readings are pulling up the overall mean?

## Key Takeaways

- `while (condition) { ... }` repeats a block of code for as long as `condition` remains `TRUE`, without needing to know the iteration count in advance.
- Every `while` loop must contain something that eventually makes its condition `FALSE` — omitting this creates an infinite loop.
- `while` loops are the right tool specifically when a stopping point depends on a condition evaluated during the loop, unlike `for` loops (Page 1), which need a known number of iterations up front.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** When is a `while` loop the more appropriate choice over a `for` loop?
   a) When you know exactly how many iterations you need in advance
   b) When the number of iterations depends on a condition that's only known while the loop runs
   c) Never — `for` loops can always replace `while` loops
   d) Only when looping over character vectors

2. **True or false.** Every `while` loop is guaranteed to eventually stop on its own, without any special care needed.

3. **Predict the output.** If `count <- 0` and the loop is `while (count < 3) { print(count); count <- count + 1 }`, how many times does it print?

4. **Find the error.**
   ```r
   count <- 0
   while (count < 5) {
     print(count)
   }
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** When the number of iterations depends on a condition that's only known while the loop runs, not a fixed count decided in advance.
2. **False.** A `while` loop only stops if something inside it actually changes the condition toward `FALSE` — without that, it runs forever (an infinite loop).
3. 3 times, printing `0`, `1`, `2`.
4. `count` is never updated inside the loop, so `count < 5` remains `TRUE` forever, creating an infinite loop that would need to be manually interrupted. It should include `count <- count + 1` inside the loop body.

</details>
:::
