# `vapply()` and `tapply()` {#sec-vapply-tapply}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 9's `sapply()` is convenient but has a quirk (also raised by `purrr::map()`, Level 9): its return type isn't guaranteed in advance. Separately, you want each county's mean blood pressure — Level 5 solved this with `group_by()` + `summarise()`, but your professor asks whether base R (predating the tidyverse) has an equally direct way to compute a grouped statistic without loading `dplyr` at all.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Use `vapply()` for a type-safe alternative to `sapply()`, and `tapply()` for a grouped summary statistic without `group_by()`.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 9's `map_dbl()` guarantees a numeric return type, erroring if a function doesn't comply. Would you expect a base R equivalent to `sapply()` with the same kind of guarantee?
- From Level 5, `group_by(participants, county)` followed by `summarise(mean_bp = mean(systolic_bp, na.rm = TRUE))` computed one mean per county. What three pieces of information would a single base R function need to do the same thing directly?
:::

## The R Concept

**`vapply()`** works like `sapply()`, but requires you to specify the expected return type upfront (`FUN.VALUE`) — exactly the base R equivalent of Level 9's `map_dbl()`/`map_chr()` guarantee:

```{r}
#| eval: false
vapply(df, function_name, FUN.VALUE = numeric(1))
```

If `function_name` ever returns something other than a single number for any column, `vapply()` errors immediately, rather than silently producing unexpected output.

**`tapply()`** computes a summary statistic for a numeric vector, split by a grouping vector — in one call, without `group_by()`:

```{r}
#| eval: false
tapply(numeric_vector, grouping_vector, function_name, na.rm = TRUE)
```

## Demonstration

```{r}
#| eval: false
pct_missing <- function(x) {
  sum(is.na(x)) / length(x) * 100
}

participants <- read.csv("data/participants.csv")

vapply(participants, pct_missing, FUN.VALUE = numeric(1))
```

```
         id         age         sex      county systolic_bp diastolic_bp
          0           6           0           0           4            0
        bmi      smoker
          5           0
```

The same results as Level 9's `sapply()` and `map_dbl()` — but `vapply()` additionally guarantees, and checks, that `pct_missing()` returns exactly one number per column, erroring immediately if that promise were ever broken.

```{r}
#| eval: false
tapply(participants$systolic_bp, participants$county, mean, na.rm = TRUE)
```

```
Cedar Falls   Millbrook     Oakford   Riverdale  Springvale
   128.1579    133.3000    129.7059    124.3571    131.4615
```

One line, no `group_by()` or `summarise()` needed — `tapply()` splits `systolic_bp` by `county` and applies `mean(..., na.rm = TRUE)` to each group directly, matching Level 5's grouped means exactly.

## Explore

Try `vapply(participants, class, FUN.VALUE = character(1))` — note `FUN.VALUE = character(1)` here, since `class()` returns text, not a number. Using the wrong `FUN.VALUE` type (say, `numeric(1)` for a function that returns text) produces an immediate, clear error.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

county_bp_means <- tapply(participants$systolic_bp, participants$county, mean, na.rm = TRUE)

round(county_bp_means, 2)
```

```
Cedar Falls   Millbrook     Oakford   Riverdale  Springvale
     128.16      133.30      129.71      124.36      131.46
```

```{r}
#| eval: false
county_age_means <- tapply(participants$age, participants$county, mean, na.rm = TRUE)

round(county_age_means, 2)
```

```
Cedar Falls   Millbrook     Oakford   Riverdale  Springvale
      50.58       48.00       42.20       51.27       45.40
```

Both grouped means computed with a single `tapply()` call each — matching Level 5's `group_by()` + `summarise()` results exactly (blood pressure) and Level 5's earlier county age findings (mean age), all without loading `dplyr` at all.

## Interpret the Result

`tapply()` proves the tidyverse isn't strictly necessary for grouped summaries — base R could always do this, just with a less pipe-friendly single-function-call style. `vapply()` similarly proves base R had a type-safety mechanism (`FUN.VALUE`) analogous to `purrr`'s type-specific `map_*()` variants long before `purrr` existed. Neither replaces the tidyverse's readability advantages for longer, multi-step pipelines — but both are worth recognizing when reading older R code that predates the tidyverse's popularity.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `FUN.VALUE` must match what the function actually returns.** `vapply(df, pct_missing, FUN.VALUE = character(1))` would error immediately, since `pct_missing()` returns numbers, not text — the mismatch is exactly the point of `vapply()`'s stricter checking, but it does mean you must know your function's output type correctly.
- **Confusing `tapply()`'s argument order.** `tapply(x, INDEX, FUN)` takes the *values* first, then the *grouping* vector, then the *function* — reversing the first two arguments (grouping by the wrong variable) is an easy mistake when moving quickly.
- **Expecting `tapply()`'s output to be a data frame.** `tapply()` returns a named array (here, effectively a named vector) — not a data frame like `summarise()` produces, so operations expecting data-frame structure (like further `filter()`-ing) won't work directly on its result without additional conversion.
:::

## Practice Problem

Use `tapply()` to compute the mean `bmi` for each smoking status (`participants$smoker`), and compare the result to Page 5's boxplot-based finding from Level 8 (non-smokers median 27.8, smokers median 26.25 — note `tapply()` here computes the *mean*, not median, so some difference is expected).

## Challenge Problem

Use `vapply()` with `FUN.VALUE = numeric(1)` to apply a function that computes each numeric column's range width (`max - min`, Level 9's practice problem) across `participants`. You'll need to first restrict to only the numeric columns (Level 5's `select()` with `where(is.numeric)`, or manual column selection), since `vapply()`'s strict type-checking will error on character columns where range width isn't meaningful.

## Key Takeaways

- `vapply(df, function_name, FUN.VALUE = ...)` works like `sapply()` but requires specifying the expected return type, erroring immediately if that promise is broken — the base R equivalent of `purrr`'s `map_dbl()`/`map_chr()`.
- `tapply(values, grouping, function_name)` computes a summary statistic split by group in one call, without needing `group_by()`.
- Both functions predate the tidyverse and prove grouped, type-safe operations were always possible in base R — the tidyverse mainly improves readability and pipe-friendliness, not raw capability.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `FUN.VALUE = numeric(1)` tell `vapply()`?
   a) To convert every column to numeric first
   b) To expect the applied function to return exactly one number per element
   c) To skip any column that isn't numeric
   d) Nothing — it's an optional, unused argument

2. **True or false.** `tapply()` requires the tidyverse's `dplyr` package to work.

3. **Predict the output.** What would `tapply(participants$age, participants$smoker, mean, na.rm = TRUE)` compute?

4. **Find the error.**
   ```r
   vapply(participants, class, FUN.VALUE = numeric(1))
   ```

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** To expect the applied function to return exactly one number per element, erroring if it doesn't.
2. **False.** `tapply()` is base R, requiring no additional packages at all.
3. The mean age for each smoking status ("No" and "Yes"), split by group in one call.
4. `class()` returns character (text) values, but `FUN.VALUE = numeric(1)` tells `vapply()` to expect numbers — this mismatch causes an immediate error. It should be `FUN.VALUE = character(1)` instead.

</details>
:::
