# Environments {#sec-environments}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 9's `03-default-arguments.qmd` mentioned that a variable created *inside* a function disappears once the function finishes, and never affects anything outside it. But you've never actually seen *why* that happens. Your professor asks you to explain, and demonstrate, what's really going on when R decides which variable a name like `x` refers to.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Understand what an environment is, how R looks up variable names inside one, and how to create and use your own.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 9 Page 6 (Scope) showed a variable defined inside a function doesn't exist outside it. If every variable in R "lives" somewhere, where does a function's local variable actually live?
- If two completely separate environments each had their own variable named `x`, would changing one affect the other?
:::

## The R Concept

An **environment** is R's container for variable names and their values — every time you assign `x <- 5`, that binding is stored in *some* environment. `environment()` reveals which one a function belongs to, and `new.env()` creates a brand-new, empty one:

```{r}
#| eval: false
new_environment <- new.env()
assign("count", 0, envir = new_environment)
get("count", envir = new_environment)
```

Your R session's main workspace is called the **global environment** (`globalenv()`). Every time a function runs, R creates a *new*, temporary environment just for that call — which is exactly why a function's local variables vanish once it returns: that temporary environment (and everything in it) is discarded.

## Demonstration

```{r}
#| eval: false
counter_env <- new.env()
assign("count", 0, envir = counter_env)

get("count", envir = counter_env)
```

```
[1] 0
```

```{r}
#| eval: false
assign("count", get("count", envir = counter_env) + 1, envir = counter_env)
get("count", envir = counter_env)
```

```
[1] 1
```

```{r}
#| eval: false
exists("count", envir = globalenv())
```

```
[1] FALSE
```

## Explore

Try `ls(envir = counter_env)` to list every variable name stored in `counter_env` — just like `ls()` (Level 1) lists everything in your global environment. Notice `count` exists inside `counter_env`, but `exists("count")` (checking the global environment by default) still returns `FALSE`.

## Solve the Problem

```{r}
#| eval: false
threshold_env <- new.env()
assign("threshold", 140, envir = threshold_env)

participants <- read.csv("data/participants.csv")

count_above_threshold <- function(values, env) {
  cutoff <- get("threshold", envir = env)
  sum(values >= cutoff, na.rm = TRUE)
}

count_above_threshold(participants$systolic_bp, threshold_env)
```

```
[1] 29
```

```{r}
#| eval: false
assign("threshold", 130, envir = threshold_env)
count_above_threshold(participants$systolic_bp, threshold_env)
```

```
[1] 46
```

Storing `threshold` inside `threshold_env` (rather than as a plain global variable) means `count_above_threshold()` always reads whatever value is currently stored there — changing the environment's value from 140 to 130 changes the function's behavior without touching the function itself, reproducing Level 4's Stage 2 (140+) count of 29 and Stage 1-or-above (130+) count of 46.

## Interpret the Result

Every function call in R creates its own temporary environment for local variables, which is discarded when the function returns — this is precisely why Level 9's scope rules work the way they do. Explicitly creating and passing around an environment (like `threshold_env` here) is a more advanced technique for when you deliberately want state to *persist* between function calls, rather than disappearing — exactly the idea Page 2's closures build on next.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Forgetting `get()` and `assign()` need an explicit `envir` argument to target a non-default environment.** Without `envir = threshold_env`, both functions default to the current function's local environment, not the one you intended.
- **Assuming a variable inside a custom environment is visible everywhere.** `exists("count")` without specifying `envir` only checks the calling environment (usually global) by default — a variable stored inside `counter_env` is invisible to a plain `exists("count")` call.
- **Creating a new environment when a simple return value would do.** Most of the time, a function should simply `return()` a value (Level 9) rather than reach for an environment — environments are for the specific cases (like closures, Page 2) where genuinely persistent, mutable state is needed.
:::

## Practice Problem

Create a new environment called `bmi_env`, store a `cutoff` value of 30 inside it, and write a function that counts how many participants' `bmi` exceeds that cutoff using `get(..., envir = bmi_env)`.

## Challenge Problem

Modify the `count_above_threshold()` function so that, in addition to returning the count, it also updates a `last_count` variable inside the same environment using `assign()`. Call the function twice with different threshold values and show that `get("last_count", envir = threshold_env)` reflects the most recent call each time.

## Key Takeaways

- An environment is R's container for variable name-to-value bindings; `new.env()` creates a new, empty one, and `assign()`/`get()` read and write values inside a specific environment.
- Every function call creates its own temporary environment for local variables, which is discarded when the function returns — the mechanism behind Level 9's scoping rules.
- Explicitly created environments let state persist and be shared across multiple function calls, unlike a function's normal, temporary local variables.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Why does a variable created inside a function disappear once the function returns?
   a) R deletes all variables starting with lowercase letters
   b) The function's temporary environment is discarded when the function finishes
   c) Local variables are never actually created
   d) R only allows one variable per script

2. **True or false.** `get("x", envir = some_env)` will find `x` even if it's only defined in the global environment, not in `some_env`.

3. **Predict the output.** If `assign("threshold", 130, envir = threshold_env)` is run after `threshold` was previously 140, what would `get("threshold", envir = threshold_env)` return next?

4. **Short answer.** Why might a function that reads its cutoff value from an environment (rather than a hardcoded number) be more flexible than a function with the cutoff written directly into its code?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The function's temporary environment is discarded when the function finishes, taking its local variables with it.
2. **False.** `get(..., envir = some_env)` looks specifically inside `some_env` (and, by default, its parent environments) — it will not find a variable that exists only in a completely unrelated environment outside that chain.
3. 130 — `assign()` overwrites the previous value stored in that environment.
4. Reading the cutoff from an environment lets the same function's behavior be changed later (by updating the environment) without editing or re-defining the function itself — exactly the flexibility closures (Page 2) build on directly.

</details>
:::
