# Robust Error Handling with `tryCatch()` {#sec-trycatch}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 9's `stop()` (Page 10) halts a function entirely the moment something goes wrong — appropriate for a single calculation, but a real problem for Level 10's automation capstone, which loops over multiple clinic files. If just *one* file is missing or malformed, `stop()` would crash the entire loop, losing every result already computed for the files that *did* work. Your professor wants the loop to skip a bad file and keep going, not give up entirely.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Use `tryCatch()` to catch an error when it happens, respond to it gracefully, and let the rest of your code keep running.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 9's `stop()` immediately halts execution. If a loop needs to process 5 files and the 3rd one causes an error, would you rather the whole loop stop, or just that one iteration be skipped?
- If you could "catch" an error instead of letting it crash your script, what would you want to happen instead — print a warning message? Return a placeholder value?
:::

## The R Concept

**`tryCatch()`** runs a block of code and lets you specify exactly what should happen if it produces an error (or a warning), instead of letting it crash your script:

```{r}
#| eval: false
tryCatch({
  risky_code_that_might_fail()
}, error = function(e) {
  cat("Something went wrong:", conditionMessage(e), "\n")
  NA   # a fallback value to use instead
})
```

`conditionMessage(e)` extracts the actual error text, letting you report *what* went wrong rather than just that something did. The `error = function(e) { ... }` block only runs if an error actually occurs — otherwise, `tryCatch()` simply returns the result of the first block, unaffected.

## Demonstration

```{r}
#| eval: false
safe_read_csv <- function(path) {
  tryCatch({
    read.csv(path)
  }, error = function(e) {
    cat("Could not read", path, "-", conditionMessage(e), "\n")
    NULL
  })
}

result_ok <- safe_read_csv("data/participants.csv")
nrow(result_ok)
```

```
[1] 100
```

```{r}
#| eval: false
result_missing <- safe_read_csv("data/does_not_exist.csv")
```

```
Could not read data/does_not_exist.csv - cannot open the connection
```

```{r}
#| eval: false
is.null(result_missing)
```

```
[1] TRUE
```

## Explore

Try wrapping Level 10's automation capstone loop (reading each file in `data/clinics/`) with `safe_read_csv()` instead of a plain `read.csv()`. Then try adding a genuinely broken path to the file list — the loop should print an error message for that one file and continue processing the rest, instead of crashing entirely.

## Solve the Problem

```{r}
#| eval: false
clinic_files <- c(
  list.files("data/clinics", pattern = "\\.csv$", full.names = TRUE),
  "data/clinics/clinic_nonexistent.csv"   # deliberately broken path
)

safe_read_csv <- function(path) {
  tryCatch({
    read.csv(path)
  }, error = function(e) {
    cat("Skipping", path, "-", conditionMessage(e), "\n")
    NULL
  })
}

all_results <- vector("list", length(clinic_files))
for (i in seq_along(clinic_files)) {
  all_results[[i]] <- safe_read_csv(clinic_files[i])
}

successful_results <- all_results[!sapply(all_results, is.null)]
length(successful_results)
```

```
Skipping data/clinics/clinic_nonexistent.csv - cannot open the connection
```

```
[1] 5
```

```{r}
#| eval: false
combined <- do.call(rbind, successful_results)
nrow(combined)
```

```
[1] 100
```

The broken file path printed a clear message and was skipped — `all_results` still contains `NULL` in its place — while the 5 genuine clinic files were read successfully and combined into the same 100-row dataset Level 10's capstone reconstructed, exactly as if the broken path had never been in the list.

## Interpret the Result

`tryCatch()` transforms a single bad file from "the whole script crashes" into "one file is skipped, with a clear message explaining why, while everything else proceeds normally." This is the difference between a fragile script that works only in perfect conditions and a robust one that degrades gracefully — an essential quality for any automation (Level 10) meant to run repeatedly on real-world data that occasionally has missing or malformed files.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Wrapping too much code in one `tryCatch()` block.** If an entire multi-step process is wrapped in a single `tryCatch()`, one small error hides *which* step actually failed — wrapping the smallest risky piece (like just the `read.csv()` call) gives more precise, useful error messages.
- **Forgetting to filter out the `NULL` placeholders afterward.** `all_results` still contains a `NULL` for each skipped file — passing it directly to `do.call(rbind, ...)` without first removing `NULL`s (via `sapply(..., is.null)`, as shown here) can cause a confusing downstream error instead.
- **Using `tryCatch()` to silently hide every error with no message at all.** An `error` handler that does nothing but `return(NULL)` with no `cat()` or logging leaves you with no way to know *which* files failed or why — always surface at least a brief message.
:::

## Practice Problem

Write a `safe_mean <- function(x)` that wraps `mean(x)` in a `tryCatch()`, returning `NA` and printing a message if `x` isn't numeric (this will require `stop()`-ing manually inside the tried code if `!is.numeric(x)`, since `mean()` on character data doesn't error by default — try `mean(c("a","b"))` first to see what actually happens).

## Challenge Problem

Extend the clinic-file-reading loop above so that, in addition to skipping unreadable files, it also uses `tryCatch()`'s `warning = function(w) {...}` argument to catch and report any *warnings* (not just errors) produced while reading a file. Test it against a file path that would produce a warning, such as one with inconsistent column types.

## Key Takeaways

- `tryCatch({ code }, error = function(e) { ... })` catches an error when it happens, letting you respond gracefully (a message, a fallback value) instead of crashing the whole script.
- `conditionMessage(e)` extracts the actual error text inside an error handler, letting you report specifically what went wrong.
- Wrapping the smallest risky piece of code (not an entire multi-step process) in `tryCatch()` gives the most precise, useful error handling.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What is the main advantage of `tryCatch()` over letting `stop()` crash a script?
   a) It runs code faster
   b) It lets you catch an error and continue running the rest of your code
   c) It automatically fixes broken file paths
   d) It prevents all errors from ever occurring

2. **True or false.** If no error occurs inside `tryCatch({ code })`, the `error` handler function never runs.

3. **Predict the output.** If a loop uses `tryCatch()` to skip 1 broken file out of 6 total files, how many results would end up in a list of successfully-read files?

4. **Short answer.** Why is wrapping just `read.csv(path)` in `tryCatch()` better than wrapping an entire 10-line data-processing pipeline in one `tryCatch()` block?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** It lets you catch an error and continue running the rest of your code, rather than crashing entirely.
2. **True.** The `error` handler is only invoked if an error actually occurs during the tried code's execution.
3. 5 — one broken file is skipped, leaving the other 5 successfully read.
4. Wrapping only the risky line pinpoints exactly which operation failed and lets every other step run normally on all other iterations — wrapping an entire pipeline means any error anywhere inside it looks identical and stops that entire iteration's processing, even parts that would have worked fine.

</details>
:::
