# Functional Programming with `Reduce()` {#sec-functional-programming}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 10's `for` loops accumulated results by explicitly pre-allocating a vector and filling it in, step by step (Page 2). Level 9's `sapply()`/`map_dbl()` apply a function to *each* element independently, but never combine those elements together one after another. Your professor wants a third pattern: repeatedly combining a whole sequence of values into a single result — like a running total — without writing an explicit loop at all.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Use `Reduce()` to combine a sequence of values into a single result, one pairwise operation at a time.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- `sum(c(1, 2, 3, 4))` adds all four numbers at once. What if, instead, you combined them two at a time: first `1 + 2`, then that result `+ 3`, then that result `+ 4`? Would you end up at the same final answer?
- Level 10 Page 2 pre-allocated a vector and looped with an index to accumulate results. Could the same "combine one at a time" idea be expressed without an explicit loop or index at all?
:::

## The R Concept

**`Reduce(function, x)`** repeatedly applies a two-argument function to combine elements of `x` one at a time, carrying an accumulated result forward:

```{r}
#| eval: false
Reduce(`+`, c(1, 2, 3, 4))
# step by step: (((1 + 2) + 3) + 4) = 10
```

Adding `accumulate = TRUE` returns *every* intermediate step, not just the final result — useful for seeing a running total, similar in spirit to Level 10's manually-built accumulator loop, but expressed in a single line.

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

ages <- na.omit(participants$age)

Reduce(`+`, ages)
```

```
[1] 4458
```

```{r}
#| eval: false
sum(ages)   # confirms Reduce("+", ages) matches R's built-in sum()
```

```
[1] 4458
```

## Explore

Try `Reduce(`+`, ages, accumulate = TRUE)` on just the first 5 ages (`ages[1:5]`) to see every running-total step along the way, rather than only the final sum. Compare it to `cumsum(ages[1:5])` — `Reduce(..., accumulate = TRUE)` and `cumsum()` produce the same running totals, just through a more general mechanism that works with any two-argument combining function, not only addition.

## Solve the Problem

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")

ages <- na.omit(participants$age)

running_totals <- Reduce(`+`, ages[1:5], accumulate = TRUE)
running_totals
```

```
[1]  46  99 158 207 268
```

```{r}
#| eval: false
Reduce(function(current_max, next_val) if (next_val > current_max) next_val else current_max, ages)
```

```
[1] 81
```

The running totals (46, 99, 158, 207, 268) confirm `Reduce(accumulate = TRUE)` matches `cumsum()`'s step-by-step behavior exactly. The second `Reduce()` call — using a custom "keep the larger value" function instead of `+` — correctly finds the maximum age (81) across the entire dataset, demonstrating `Reduce()` works with *any* two-argument combining function, not just arithmetic ones.

## Interpret the Result

`Reduce()` generalizes the "combine values one at a time, carrying a result forward" pattern beyond simple addition — the exact same function works whether you're summing numbers, finding a maximum, or (as the challenge problem explores) combining data frames. This is a genuinely different mental model from Level 9's `sapply()`/`map_dbl()` (which apply a function to each element *independently*, with no memory of prior elements) and Level 10's explicit accumulator loops (which require manually managing an index and a pre-allocated vector).

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Confusing `Reduce()` (combines elements together) with `sapply()`/`map()` (transforms each element independently).** `sapply(x, f)` returns one output per input element; `Reduce(f, x)` returns a single combined result (or, with `accumulate = TRUE`, one result per step of the combination) — these solve different problems.
- **Forgetting the combining function needs exactly two arguments.** A custom function passed to `Reduce()`, like the maximum-finder above, must accept exactly two arguments — the accumulated result so far, and the next element — anything else will error.
- **Not realizing `do.call(rbind, list_of_dataframes)` (Level 10's capstone) is itself a `Reduce()`-like pattern.** `do.call(rbind, list)` combines every data frame in a list into one, conceptually similar to `Reduce(rbind, list)` — both express "combine a list of things into one," just with slightly different syntax.
:::

## Practice Problem

Use `Reduce()` with a custom two-argument function to find the *minimum* age in `participants$age` (removing missing values first), without using R's built-in `min()`.

## Challenge Problem

Level 10's automation capstone used `do.call(rbind, clinic_data_list)` to combine 5 clinic data frames into one. Rewrite that combination step using `Reduce(rbind, clinic_data_list)` instead, and confirm it produces the identical 100-row combined data frame both ways.

## Key Takeaways

- `Reduce(function, x)` repeatedly applies a two-argument function to combine all elements of `x` into a single result, carrying an accumulated value forward at each step.
- `accumulate = TRUE` returns every intermediate step of the combination, not just the final result — comparable to `cumsum()`, but generalized to any two-argument combining function.
- `Reduce()` solves a fundamentally different problem than `sapply()`/`map()` (which transform elements independently) or an explicit accumulator loop (Level 10) — all three are valid tools for different situations.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** What does `Reduce(`+`, c(1, 2, 3, 4))` compute?
   a) A vector `c(1, 2, 3, 4)` unchanged
   b) The single combined value `10`
   c) Four separate sums
   d) An error, since `+` isn't a valid function name

2. **True or false.** `Reduce(f, x, accumulate = TRUE)` returns only the final combined result, discarding every intermediate step.

3. **Predict the output.** What would `Reduce(`+`, c(46, 53, 59, 49, 61), accumulate = TRUE)` return?

4. **Short answer.** How is `Reduce()` fundamentally different from `sapply()` or `purrr::map_dbl()`?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The single combined value `10` — computed step by step as `((1 + 2) + 3) + 4`.
2. **False.** `accumulate = TRUE` returns *every* intermediate step of the combination, not just the final result.
3. `c(46, 99, 158, 207, 268)` — the running total after each successive addition.
4. `sapply()`/`map_dbl()` apply a function to each element of a vector *independently*, producing one output per input with no memory between elements; `Reduce()` instead combines elements together sequentially, carrying an accumulated result forward from one step to the next.

</details>
:::
