# Level 13 Capstone: A Resilient File-Reading Toolkit {#sec-advanced-r-capstone}

::: {.pbl-box .pbl-problem}
**The Problem**

Level 10's automation capstone combined all 5 clinic files with `list.files()` and a `for` loop — but it assumed every file would be present and well-formed. Your professor wants a genuinely production-ready version: one that uses a regex pattern to find files (Page 7), skips unreadable ones gracefully instead of crashing (Page 4), and returns a clean, custom-printed summary object (Page 3) reporting exactly what happened — not just the combined data.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Combine this level's tools — closures, S3 objects, error handling, and regex — into one resilient, reusable function that reads a folder of files and reports on its own success.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Level 10's capstone used `list.files(pattern = "\\.csv$")` to find files. If one of those files were deleted or corrupted, would the original loop keep going, or crash entirely?
- If a function's return value needs to report *both* the combined data *and* metadata about what succeeded or failed, would a single data frame be enough, or would a custom S3 object (Page 3) fit better?
:::

## The R Concept

This capstone's function combines four separate ideas from across this level into one:

```{r}
#| eval: false
read_clinic_folder <- function(folder_path) {
  csv_files <- list.files(folder_path, pattern = "\\.csv$", full.names = TRUE)  # Page 7 (regex)

  results <- vector("list", length(csv_files))
  failed <- character(0)

  for (i in seq_along(csv_files)) {
    results[[i]] <- tryCatch({                    # Page 4 (error handling)
      read.csv(csv_files[i])
    }, error = function(e) {
      failed <<- c(failed, csv_files[i])
      NULL
    })
  }

  successful <- results[!sapply(results, is.null)]

  structure(list(                                  # Page 3 (S3 object)
    data = do.call(rbind, successful),
    n_files_found = length(csv_files),
    n_files_read = length(successful),
    failed_files = failed
  ), class = "clinic_batch")
}

print.clinic_batch <- function(x, ...) {
  cat(sprintf("Clinic batch: %d/%d files read successfully (%d rows total)\n",
              x$n_files_read, x$n_files_found, nrow(x$data)))
  if (length(x$failed_files) > 0) {
    cat("Failed files:", paste(x$failed_files, collapse = ", "), "\n")
  }
}
```

## Demonstration

```{r}
#| eval: false
batch <- read_clinic_folder("data/clinics")

batch
```

```
Clinic batch: 5/5 files read successfully (100 rows total)
```

```{r}
#| eval: false
class(batch)
```

```
[1] "clinic_batch"
```

## Explore

Try running `read_clinic_folder()` against a folder containing a deliberately broken or missing file — copy `data/clinics/` to a new folder, add a non-CSV file, or remove read permissions from one file, and observe how `batch$failed_files` reports exactly which file(s) failed, without crashing the whole function.

## Solve the Problem

```{r}
#| eval: false
read_clinic_folder <- function(folder_path) {
  csv_files <- list.files(folder_path, pattern = "\\.csv$", full.names = TRUE)

  results <- vector("list", length(csv_files))
  failed <- character(0)

  for (i in seq_along(csv_files)) {
    results[[i]] <- tryCatch({
      read.csv(csv_files[i])
    }, error = function(e) {
      failed <<- c(failed, csv_files[i])
      NULL
    })
  }

  successful <- results[!sapply(results, is.null)]

  structure(list(
    data = do.call(rbind, successful),
    n_files_found = length(csv_files),
    n_files_read = length(successful),
    failed_files = failed
  ), class = "clinic_batch")
}

print.clinic_batch <- function(x, ...) {
  cat(sprintf("Clinic batch: %d/%d files read successfully (%d rows total)\n",
              x$n_files_read, x$n_files_found, nrow(x$data)))
  if (length(x$failed_files) > 0) {
    cat("Failed files:", paste(x$failed_files, collapse = ", "), "\n")
  }
}

batch <- read_clinic_folder("data/clinics")

nrow(batch$data)
batch$n_files_read
```

```
[1] 100
```

```
[1] 5
```

```{r}
#| eval: false
table(batch$data$county)
```

```

Cedar Falls    Millbrook      Oakford    Riverdale   Springvale
         20           21           17           16           26
```

All 5 clinic files were found (via the same `"\\.csv$"` regex pattern from Page 7) and successfully read (using `tryCatch()`-wrapped reads from Page 4), producing the identical 100-row combined dataset and county breakdown that Level 10's capstone reconstructed — but this version, unlike Level 10's, would have survived a missing or corrupted file, reporting it in `failed_files` instead of crashing.

## Interpret the Result

This capstone function is a genuine step up from Level 10's automation: it doesn't just automate a repeated task, it does so *resiliently*, reporting exactly what succeeded and what didn't through a clean, custom-printed S3 object rather than a raw data frame or a script that halts on the first problem. This combination — regex-based file discovery, graceful error handling, and an informative custom result type — is close to what a genuinely reusable, shareable R function (Page 10's package structure) should look like.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Using `<-` instead of `<<-` inside the `tryCatch()` error handler.** `failed <<- c(failed, csv_files[i])` uses the super-assignment operator (Page 1's environments) specifically because the error handler runs in its own function environment — a plain `<-` would create a *new* local `failed` variable instead of updating the one the loop can see afterward.
- **Forgetting `do.call(rbind, successful)` fails if `successful` is empty.** If every single file fails to read, `successful` is an empty list, and `do.call(rbind, list())` produces `NULL` rather than an error — code using `batch$data` afterward should check for this case before assuming it's a proper data frame.
- **Not testing the failure path at all.** It's tempting to only test a function like this against files that work — deliberately testing against a missing or malformed file (as the Explore section suggests) is the only way to confirm the error-handling logic actually behaves as intended.
:::

## Practice Problem

Modify `print.clinic_batch()` to also report the mean `systolic_bp` across all successfully-combined rows, alongside the existing file-count summary.

## Challenge Problem

Add a `summary.clinic_batch()` S3 method (Page 3's pattern) that returns a data frame with one row per county, showing each county's row count and mean BMI — effectively reproducing Level 5's `group_by(county) %>% summarise()` output, but generated from this capstone's resilient file-reading function instead of a single pre-combined CSV.

## Key Takeaways

- A genuinely resilient automation function combines several ideas at once: pattern-based file discovery (regex), graceful error handling (`tryCatch()`), and an informative, custom-printed result object (S3) — rather than any single technique alone.
- `<<-` (super-assignment) is often necessary inside a `tryCatch()` error handler or other nested function, specifically to update a variable in an enclosing environment rather than creating a new local one.
- Testing a function's failure path deliberately (not just its success path) is the only reliable way to confirm error-handling logic actually works as intended.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** Why does the error handler inside `read_clinic_folder()` use `failed <<- c(failed, csv_files[i])` instead of `failed <- c(failed, csv_files[i])`?
   a) `<<-` is required inside every `tryCatch()` call
   b) `<<-` updates the `failed` variable in the enclosing function's environment, rather than creating a new local one
   c) `<-` doesn't work with character vectors
   d) There is no actual difference between the two

2. **True or false.** This capstone's function will crash entirely if even one file in the folder is unreadable.

3. **Predict the output.** If `read_clinic_folder()` is run on a folder with 5 valid CSV files, what would `batch$n_files_read` equal?

4. **Short answer.** Why is combining regex, error handling, and an S3 result object a meaningfully more resilient design than Level 10's original automation capstone?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** `<<-` updates the `failed` variable in the enclosing function's environment, rather than creating a new local variable that would disappear once the error handler finishes.
2. **False.** The whole point of wrapping each file's `read.csv()` call in `tryCatch()` is that one failed file is skipped and recorded, while the rest of the loop continues normally.
3. 5 — all 5 valid files would be successfully read.
4. Level 10's original version assumed every file would always be present and well-formed, and would crash entirely on the first bad file; this version discovers files by pattern, survives individual file failures gracefully, and reports exactly what happened through a clean, custom-printed summary object — a combination that degrades gracefully instead of failing all-or-nothing.

</details>
:::
