# Level 14 Capstone: A Complete Clinic Report Template {#sec-reproducible-research-capstone}

::: {.pbl-box .pbl-problem}
**The Problem**

Your professor now wants everything from this level combined into one deliverable: a single, parameterized clinic report template that can regenerate a polished, properly-captioned, cited summary for any of Level 10's five clinics — county-appropriate statistics, a formatted table, a captioned figure, and a bibliography crediting the tools used — all from one reusable source file, rendered five separate times.
:::

::: {.pbl-box .pbl-mission}
**Your Mission**

Build a complete, reproducible report template combining parameters, inline code, tables, figures, cross-references, and citations into one cohesive document.
:::

::: {.pbl-box .pbl-think}
**Think Before Coding**

- Of this level's tools — parameters (Page 7), tables (Page 4), figures (Page 4), cross-references (Page 5), and citations (Page 6) — which ones need to change their *behavior* based on which county is selected, and which ones stay the same regardless?
- If this template is rendered 5 times (once per county), should the bibliography's content differ between renders, or stay identical every time?
:::

## The R Concept

A complete parameterized report brings every piece from this level together in one YAML block and a handful of chunks:

```yaml
---
title: "Clinic Summary: ˋr params$countyˋ"
params:
  county: "Oakford"
bibliography: references.bib
---
```

```{r}
#| eval: false
#| label: tbl-clinic-summary-concept
#| tbl-cap: paste("Summary statistics for", params$county)
participants <- read.csv("data/participants.csv")
clinic_data <- participants[participants$county == params$county, ]

summary_table <- data.frame(
  Metric = c("Patients", "Mean SBP", "Mean BMI", "Current Smokers"),
  Value = c(nrow(clinic_data),
            round(mean(clinic_data$systolic_bp, na.rm = TRUE), 1),
            round(mean(clinic_data$bmi, na.rm = TRUE), 1),
            sum(clinic_data$smoker == "Yes"))
)
knitr::kable(summary_table)
```

## Demonstration

```{r}
#| eval: false
participants <- read.csv("data/participants.csv")
clinic_data <- participants[participants$county == "Oakford", ]

summary_table <- data.frame(
  Metric = c("Patients", "Mean SBP", "Mean BMI", "Current Smokers"),
  Value = c(nrow(clinic_data),
            round(mean(clinic_data$systolic_bp, na.rm = TRUE), 1),
            round(mean(clinic_data$bmi, na.rm = TRUE), 1),
            sum(clinic_data$smoker == "Yes"))
)
knitr::kable(summary_table)
```

```
|Metric          | Value|
|:---------------|-----:|
|Patients        |  17.0|
|Mean SBP        | 129.7|
|Mean BMI        |  28.0|
|Current Smokers |   6.0|
```

## Explore

Try re-running the same two chunks above with `"Millbrook"` in place of `"Oakford"` — every number in the table changes (Millbrook has 21 patients, not 17), while the table's structure, caption pattern, and surrounding prose template remain completely unchanged.

## Solve the Problem

```yaml
---
title: "Clinic Summary: ˋr params$countyˋ"
params:
  county: "Oakford"
bibliography: references.bib
---

## Overview {#sec-clinic-overview}

This report summarizes patients enrolled at the ˋr params$countyˋ clinic
site, as part of the broader study described in @sec-clinic-overview.
Data cleaning followed tidyverse conventions [@wickham2023r4ds].

## Summary Statistics
```

```{r}
#| eval: false
#| label: tbl-clinic-summary
#| tbl-cap: paste("Summary statistics for", params$county)
participants <- read.csv("data/participants.csv")
clinic_data <- participants[participants$county == params$county, ]

summary_table <- data.frame(
  Metric = c("Patients", "Mean SBP", "Mean BMI", "Current Smokers"),
  Value = c(nrow(clinic_data),
            round(mean(clinic_data$systolic_bp, na.rm = TRUE), 1),
            round(mean(clinic_data$bmi, na.rm = TRUE), 1),
            sum(clinic_data$smoker == "Yes"))
)
knitr::kable(summary_table)
```

```{r}
#| eval: false
#| label: fig-clinic-bp-distribution
#| fig-cap: paste("Systolic blood pressure distribution,", params$county)
library(ggplot2)
ggplot(clinic_data, aes(x = systolic_bp)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Count")
```

```
## References
```

```
Rendered with params: county: "Oakford" -> "Clinic Summary: Oakford" with
Table 1 showing 17 patients, mean SBP 129.7; Figure 1 showing Oakford's
blood pressure histogram; and a References section citing R for Data
Science.

Rendered again with params: county: "Riverdale" -> the identical template
instead produces "Clinic Summary: Riverdale" with Table 1 showing 16
patients, mean SBP 124.4; a completely different Figure 1 for Riverdale's
data; and the exact same References section, since the cited sources
don't depend on which county was selected.
```

## Interpret the Result

This capstone demonstrates the complete reproducible-research workflow: one parameterized template (Page 7) produces a correctly-labeled, cross-referenced (Page 5) table and figure (Page 4) with inline-code-driven prose (Page 3) and a properly cited bibliography (Page 6) — and every single piece of this automatically regenerates correctly for any of the 5 counties, just by changing one `params$county` value. Nothing about the report's *content* needed to be rewritten by hand between counties; only the parameter changed, and every downstream number, table, figure, and caption followed automatically.

::: {.pbl-box .pbl-mistake}
**Common Mistakes**

- **Hardcoding a county-specific number somewhere in the prose "just this once."** Even a single hardcoded value (like manually typing "17 patients" instead of using inline code) breaks the whole point of parameterization — that same sentence would be silently wrong the moment the template renders for a different county.
- **Forgetting the bibliography and cross-references don't need to change between parameter values.** Only the county-dependent calculations (the table, the figure, county-specific inline numbers) should reference `params$county` — citations and section cross-references stay identical across every rendered version of the report.
- **Not testing the template with more than one parameter value before considering it finished.** A template that only renders correctly for its YAML default (`"Oakford"`) but silently breaks for `"Millbrook"` or `"Springvale"` isn't actually a working parameterized report — always test at least 2-3 different values before trusting it.
:::

## Practice Problem

Extend the summary table to include a fifth row, `"Mean Age"`, computed from `clinic_data$age`, and re-render the template (mentally, or by tracing through the code) for `params$county = "Springvale"`.

## Challenge Problem

Write the R code (using `quarto::quarto_render()` inside a `for` loop, Level 10) that would render this template once for each of the 5 counties in `participants$county`, saving each output as `clinic_report_<county>.html`, and briefly explain how this single capstone script replaces what would otherwise require 5 separately maintained report files.

## Key Takeaways

- A complete reproducible report combines this level's tools together: parameters drive what data is shown, inline code keeps prose numbers live, tables and figures carry proper captions, cross-references link content reliably, and citations credit sources automatically.
- Testing a parameterized template against multiple different parameter values (not just its YAML default) is the only way to confirm it's genuinely reusable, not accidentally hardcoded somewhere.
- This entire course — every level, every page, every verified statistic — was itself built following exactly this same reproducible-research discipline, from Level 1's first `read.csv()` call through this very capstone.

::: {.pbl-box .pbl-check}
**Knowledge Check**

1. **Multiple choice.** In this capstone's template, what changes between rendering for `"Oakford"` versus `"Riverdale"`?
   a) The bibliography's cited sources
   b) The table's and figure's underlying data and captions, driven by `params$county`
   c) The cross-reference labels used in the document
   d) Nothing changes between renders

2. **True or false.** A parameterized report template should be tested with only its YAML default value before being considered complete.

3. **Predict the output.** If `params$county` is set to `"Millbrook"` (21 patients, per Level 10), what would the `tbl-clinic-summary` table's "Patients" row show?

4. **Short answer.** Why does hardcoding even one county-specific number in this template's prose undermine the entire point of building it as a parameterized report?

<details class="pbl-answer">
<summary>Show answers</summary>

1. **b)** The table's and figure's underlying data and captions — both driven by `params$county` — change between renders; the bibliography and cross-reference structure stay the same.
2. **False.** A template should be tested against multiple different parameter values to confirm it's genuinely reusable and not accidentally hardcoded somewhere.
3. 21 — matching Level 10's known Millbrook row count.
4. A single hardcoded number defeats the purpose of parameterization for that specific value — the report would continue displaying that one incorrect, frozen number no matter which county is actually selected via `params$county`, silently undermining the reliability of the entire template.

</details>
:::
